---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# diseasenowcasting

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/diseasenowcasting)](https://CRAN.R-project.org/package=diseasenowcasting)
[![Codecov test coverage](https://codecov.io/gh/RodrigoZepeda/diseasenowcasting/graph/badge.svg)](https://app.codecov.io/gh/RodrigoZepeda/diseasenowcasting)
[![R-CMD-check](https://github.com/RodrigoZepeda/diseasenowcasting/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RodrigoZepeda/diseasenowcasting/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`diseasenowcasting` is an R package for **nowcasting** time series of
epidemiological cases. 

## What is nowcasting?

Surveillance systems have a delay between the
**true date of an event** (`event_date`, e.g. symptom onset or testing) and the
**report date** for that event (`report_date`, e.g. when the case was entered
into the database). Because recent event-dates are still missing reports that
have not yet arrived, the latest case counts look artificially low. A *nowcast*
corrects for this reporting delay and estimates the eventual, fully-observed
totals.

To do this `diseasenowcasting` fits censored Bayesian
models. The reporting delay is modeled directly as a stochastic process,
jointly with the epidemic dynamics, through a censored likelihood. Inference runs
on R's Template Model Builder
([`RTMB`](https://cran.r-project.org/package=RTMB)), so no Stan/JAGS compilation is required. The
package supports several [epidemic processes](https://rodrigozepeda.github.io/diseasenowcasting/reference/epidemic_process.html), [several delay
families](https://rodrigozepeda.github.io/diseasenowcasting/reference/delay_process.html), stratified data,
[extreme value detection](https://rodrigozepeda.github.io/diseasenowcasting/reference/extreme_values.html) and [backtesting](https://rodrigozepeda.github.io/diseasenowcasting/reference/backtest.html).

> :warning: `diseasenowcasting` is currently under active development and parts
> of the interface might still change.

## Installing

You can install `diseasenowcasting` (and its companion data package
[`tbl.now`](https://github.com/RodrigoZepeda/tbl.now)) from GitHub with
[`pak`](https://pak.r-lib.org/):

```{r, eval = FALSE}
# install.packages("pak") # <- uncomment if you do not have `pak`
pak::pkg_install("RodrigoZepeda/tbl.now")
pak::pkg_install("RodrigoZepeda/diseasenowcasting")
```

## Example

`diseasenowcasting` works with data organised as a `tbl_now` object from the
companion [`tbl.now`](https://rodrigozepeda.github.io/tbl.now/) package: a data
frame annotated with the **roles** of its columns (event date, report date, and
optionally the `strata` and the `now` date up to which everything is observed).

```{r setup, message=FALSE}
set.seed(6728)
library(diseasenowcasting)
library(tbl.now)
library(dplyr)

# Example surveillance data (dengue in Puerto Rico) shipped with tbl.now
data(denguedat)

#Simulate a nowcast on January 1st 1991
#Only cases and reports before that date will be used
dengue_data <- denguedat |> 
  filter(onset_week <= as.Date("1991-01-01") & report_week <= as.Date("1991-01-01"))
```

We annotate the data with `tbl_now()`, pointing at the event and report dates
(here the symptom-onset and report weeks) and the strata (`gender`):

```{r tblnow}
dengue_tbl <- tbl_now(
  dengue_data,
  event_date  = onset_week,    # when symptoms started
  report_date = report_week,   # when the case was registered
  strata      = gender,
  data_type   = "linelist",    # use "count-incidence" for pre-aggregated counts
  now         = as.Date("1991-01-01"),
  t_effects   = temporal_effects(seasons = 52) #Weekly seasonal effect (Fourier)  
)
```

Once the data is a `tbl_now()`, a single call to `nowcast()` does the rest (the
nowcast is automatically stratified by the strata declared above):

```{r nowcast}
ncast <- nowcast(dengue_tbl, n_draws = 1000)
```

Results can be visualised with `autoplot()` (bars show the median, error bars the
90% credible interval) and extracted with `predict()` / `summary()`:

```{r plot, fig.cap = "_Nowcast for the dengue example, by gender._"}
autoplot(ncast)
```

```{r predict, eval = TRUE}
pred <- predict(ncast)   # full posterior-predictive nowcast at every event-time
```

```{r summary, eval = FALSE}
summary(pred)
```

```{r summary2, eval = TRUE, echo = FALSE}
summary(pred) |> dplyr::as_tibble() |> tail(6)
```

You can choose a different epidemic process, delay family or likelihood by passing a
`model()` to `nowcast()`:

```{r model, eval = TRUE}
model_1 <- model(
  likelihood = nb_likelihood(),          # negative binomial (recommended)
  epidemic   = ar1_epidemic(),           # AR(1) process 
  delay      = generalized_gamma_delay() # generalized gamma reporting delay
)
```

```{r, eval = FALSE}
nowcast(dengue_tbl, model = model_1)
```

And `backtest` and `score` to evaluate how different models would have performed in the past

```{r backtest}
model_2 <- model(
  likelihood = nb_likelihood(),  # negative binomial (recommended)
  epidemic   = hsgp_epidemic(),  # recommended for long-running epidemics
  delay      = lognormal_delay() # generalized gamma reporting delay
)

#Set ndates to the number of dates you want for the backtest
bt <- backtest(dengue_tbl, list(model_1, model_2), n_dates = 3)
```

And get the scores

```{r}
autoplot(bt)
```
You can access these quantities directly with `score`

```{r}
score(bt)
```

## Handling extreme delays

When new data arrives, `update()` re-scores every incoming report against 
the previously fitted delay distribution and **warns** about anything that 
seems implausibly delayed. 
The flagged reports are available via `extreme_values()`:

```{r censor-detect, eval = FALSE, warning=FALSE}
#Simulate getting new data
dengue_update <- denguedat |> 
  filter(onset_week <= as.Date("1991-01-14") & report_week <= as.Date("1991-01-14"))

# level stands for when should we flag something is an extreme delay
# level = 0.99 means delays with probability 1  - 0.99 = 0.01 are considered extreme
nc_updated <- update(ncast, new_data = dengue_update, level = 0.99)

# The model identified two extreme (very unlikely) delays of 9 and 10 weeks: 
# A human should check them based on domain knowledge and decide whether to keep
# them or to censor them. 
extreme_values(nc_updated)  
```

A complete tutorial on handling extreme delays is available at the
[Handling Outlier Delays with Censoring](https://rodrigozepeda.github.io/diseasenowcasting/articles/Handling_Outlier_Delays_with_Censoring.html) vignette.

## See also

A good place to start is the
**[Introduction to diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/articles/introduction.html)**,
which walks through building a `tbl_now`, fitting with `nowcast()`, inspecting
results with `predict()` / `summary()` / `autoplot()`, and comparing models with
`backtest()` / `score()`.

From there, depending on what you want to do next:

- **[Nowcasting at the Start of an Epidemic](https://rodrigozepeda.github.io/diseasenowcasting/articles/Nowcasting_at_the_start_of_an_Epidemic.html)**
  — A worked, end-to-end case study of monitoring an outbreak in real time:
  choosing a model when data are scarce, reading the nowcast as the epidemic
  grows, and using the prior to encode what you already believe before much data
  has arrived.

- **[Understanding Priors in diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/articles/Understanding_Priors.html)**
  — *Make the model say what you mean.* The package's default priors, how to
  tighten or loosen them, and how to *see* what a prior implies before fitting
  (`nowcast(..., prior_only = TRUE)`).

- **[Handling Outlier Delays with Censoring](https://rodrigozepeda.github.io/diseasenowcasting/articles/Handling_Outlier_Delays_with_Censoring.html)**
  — *Robustness to reporting glitches.* How the censored likelihood copes with
  unusually long delays, and how to flag and censor extreme delays in your
  surveillance stream.
  
- **[Using alongside an LLM](LLM_Usage.html)**
  — *Use AI.* How to use the [`SKILL.md`](https://github.com/RodrigoZepeda/diseasenowcasting/blob/master/SKILL.md)
  to teach a Large Language Model how to you develop your nowcasts with `diseasenowcasting`. 

- **[Benchmark (diseasenowcasting vs NobBS and epinowcast)](https://rodrigozepeda.github.io/diseasenowcasting/articles/Benchmark.html)**
  — *How does it compare?* A reproducible backtest comparing `diseasenowcasting`
  against the `NobBS` and `epinowcast` packages.

- **[Mathematical Foundations of diseasenowcasting](https://rodrigozepeda.github.io/diseasenowcasting/articles/Mathematics.html)**
  — *Under the hood.* The censored likelihood, the epidemic processes (HSGP,
  AR(1), SIR), the delay families, and the Laplace-approximation inference that
  powers `RTMB`.

See `?nowcast`, `?backtest`, `?score`, and `?model` for the full function
documentation.

