---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# diseasenowcasting

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/diseasenowcasting)](https://CRAN.R-project.org/package=diseasenowcasting)
[![Codecov test coverage](https://codecov.io/gh/RodrigoZepeda/diseasenowcasting/graph/badge.svg)](https://app.codecov.io/gh/RodrigoZepeda/diseasenowcasting)
[![R-CMD-check](https://github.com/RodrigoZepeda/diseasenowcasting/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RodrigoZepeda/diseasenowcasting/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->


`diseasenowcasting` is an R package for nowcasting time series of epidemiological cases. Epidemiologic surveillance tools usually have an intrinsic delay between the **true date** of an event and the **report date** for that event. Some examples include the **true date** being symptom onset or testing time and the **report date** corresponds to when the case was registered in the system. `diseasenowcasting` uses Bayesian structural time series models (via the probabilistic programming language [Stan](https://mc-stan.org/)) to infer the cases that have not yet been reported thus providing a prediction of the final number of cases. 

> :warning: `diseasenowcasting` is currently under active development and some interface might change from the final version. 

## Installing

To install `diseasenowcasting` you can do it via `pak`:

```{r, eval = FALSE}
#install.packages("pak") # <- Uncomment if you have not installed the `pak` package
pak::pkg_install("RodrigoZepeda/diseasenowcasting")
```

Currently you also need to install the [tbl.now](https://github.com/rodrigozepeda/tbl.now) package

```{r, eval = FALSE}
pak::pkg_install("RodrigoZepeda/tbl.now")
```

## Example

```{r, message = FALSE, warning = FALSE}
set.seed(6728)
library(diseasenowcasting)
library(dplyr)
```

For this example we will use the `denguedat` dataset that comes with the package. 

```{r}
#Load example dataset
data(denguedat)
```

The [nowcast()] function performs the `nowcast`. The things required are the `true_date` and `report_date` and `strata` (if exists). 

```{r}
#Use just a subsample of the data for the example
denguedat <- denguedat |> 
  filter(report_week <= as.Date("1990/08/01", format = "%Y/%m/%d"))

#Run the nowcast model stratified by gender. Refresh = 0 makes it quiet
ncast <- nowcast(denguedat, true_date = "onset_week", report_date = "report_week", 
                       strata = "gender", refresh = 0)
```

The `summary` and `plot` options allow you to use the predictions:

```{r}
#Create a nice plot for the predictions
plot(ncast, datesbrakes = "1 month")
```

## Tutorials

You can read the following articles for a more in-depth introduction to the package:

+ [An introduction to the package:](https://rodrigozepeda.github.io/diseasenowcasting/articles/Introduction.html) provides an overview of the main functions. 
+ [Advanced nowcast options:](https://rodrigozepeda.github.io/diseasenowcasting/articles/Advanced-nowcast-options.html) provides an in-depth look at the `nowcast()` function and its options. 
+ [Comparison to other methods:](https://rodrigozepeda.github.io/diseasenowcasting/articles/Comparison-to-other-methods.html) provides a comparison in terms of metrics to other nowcasting packages. 

## Developing

If you are interested on developing the package, after downloading the repository you might need to run 

```{r, eval = FALSE}
rstantools::rstan_config()
```

before the 

```{r, eval = FALSE}
devtools::load_all(".")
```

in order for the `C++` files to be generated for your specific system. 