---
title: "Benchmark (diseasenowcasting vs NobBS and epinowcast)"
author: "Rodrigo Zepeda-Tello"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Benchmark (diseasenowcasting vs NobBS and epinowcast)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 9,
  fig.height = 4.5,
  fig.align='center',
  collapse = TRUE,
  comment  = "#>"
)
```

## TL; DR

We compared different datasets across distinct `diseasenowcasting` models as well as
NobBS and epinowcast. The `HSGP/LogNormal` and `HSGP/GeneralizedGamma` are the suggested models as they are
consistently the ranking among the top performers. 


## Overview

This vignette presents a systematic benchmark comparison of `diseasenowcasting` against two widely
used nowcasting methods: [**NobBS**](https://cran.r-project.org/web/packages/NobBS/index.html)  and 
[**Epinowcast**](https://github.com/epinowcast/epinowcast).  We compare across three real-world disease surveillance datasets:

| Dataset | Disease | Temporal unit | Source |
|---------|---------|---------------|--------|
| `denguedat` (tbl.now) | Dengue fever | Weekly | [@mcgough2020nowcasting] |
| `mpoxdat` (tbl.now)   | Mpox | Daily | [@rohrer2025nowcasting] |
| `covid_colombia` (diseasenowcasting) | COVID-19 | Daily | [@colombia] |

The comparison uses **50 evaluation dates** per disease, selected at random. Forecasts are scored for a nowcast at a 0 delay, that is, the newest event time 
at each evaluation date. 


## Scoring methodology

All methods are scored using the **Weighted Interval Score (WIS)**.  WIS decomposes
into:

- **Overprediction** — penalty for predicted intervals that are too high relative to the truth.
- **Underprediction** — penalty for predicted intervals that are too low.
- **Dispersion** — penalty for uncertainty intervals that are too wide.

Coverage at 50% and 90% credible intervals is also reported.  A good model should
have empirical coverage close to the corresponding level.

All comparisons are performed on a **common date set** — dates where both `diseasenowcasting` and the
comparison method have produced a valid nowcast — ensuring fairness.


## Results

You can access the results from inside the package:

```{r load-data}
library(diseasenowcasting)
library(dplyr)

# Pre-computed benchmark results (50 evaluation dates per disease)
scoring_file <- system.file("extdata", "comparison_scores.rds", package = "diseasenowcasting")
scores_df    <- readRDS(scoring_file) 

#Rename the columns
scores_df <- scores_df |> 
  rename(Model = model, WIS = wis, Overprediction = over, Underprediction = under,
         Dispersion = disp, Bias = bias, Cov50 = cov50, Cov90 = cov90, `N dates` = n_dates)
```

### Dengue fever (weekly, Colombia)

```{r dengue-table, echo = FALSE}
#Read the dengue tab
dengue_tab <- scores_df |> 
  filter(disease == "dengue") |> 
  select(Model, WIS, Overprediction, Underprediction, Dispersion, Bias,
         Cov50, Cov90, `N dates`)


knitr::kable(dengue_tab, digits = 2, row.names = FALSE,
             caption = "Dengue nowcast scores at d*=0 (50 common evaluation dates, weekly).")
```

**Key findings (dengue):**

- All `diseasenowcasting` HSGP variants beat NobBS and all Epinowcast variants on WIS.
- Best `diseasenowcasting`: **HSGP / GenGamma**, WIS ≈ 7.0, cov90 ≈ 0.94.

### Mpox (daily, USA)

```{r mpox-table, echo = FALSE}
mpox_tab <- scores_df |> 
  filter(disease == "mpox") |> 
  select(Model, WIS, Overprediction, Underprediction, Dispersion, Bias,
         Cov50, Cov90, `N dates`)

knitr::kable(mpox_tab, digits = 2, row.names = FALSE,
             caption = "Mpox nowcast scores at d*=0 (49 common evaluation dates, daily).")
```

**Key findings (mpox):**

- `diseasenowcasting` AR1 variants achieve the lowest WIS, beating NobBS by a wide margin (~7×).
- Best `diseasenowcasting`: **AR1 / Gamma**, 

### COVID-19 (daily, Colombia)

```{r covid-table, echo = FALSE}
covid_tab <- scores_df |> 
  filter(disease == "covid") |> 
  select(Model, WIS, Overprediction, Underprediction, Dispersion, Bias,
         Cov50, Cov90, `N dates`)

knitr::kable(covid_tab, digits = 2, row.names = FALSE,
             caption = "COVID-19 nowcast scores at d*=0 (50 common evaluation dates, daily).")
```

**Key findings (COVID-19):**

- `diseasenowcasting` beats NobBS on WIS for COVID as well, but the margin depends on the delay family.
- Best `diseasenowcasting`: **HSGP / GenGamma**.

## Conclusion

`HSGP/LogNormal` and `HSGP/GeneralizedGamma` are the suggested models as they are
consistently the ranking among the top performers. 

## How to reproduce

The full benchmark — the **complete own-model grid** (HSGP / AR(1) / SIR x
{LogNormal, Gamma, Generalized-Gamma, Dirichlet}) fit with `nowcast_twostage()`
and the seasonal / day-of-week covariates, plus **NobBS** and **three
Epinowcast variants** (point effect / random effect, default, random walk),
scored on a common evaluation-date set — lives in a single self-contained
script:

```
devel/benchmark_full.R
```

Run it (after `R CMD INSTALL` of this package, so the parallel workers can load
it):

```sh
Rscript devel/benchmark_full.R                                  # all 3 diseases, 50 dates
N_DATES=6 RUN_COVID=FALSE RUN_MPOX=FALSE Rscript devel/benchmark_full.R   # quick smoke
```

It prints a ranked WIS table per disease and saves the tidy scores to
`devel/results/benchmark_scores.rds`.

> Requirements: `NobBS` and `epinowcast` (the latter needs `cmdstanr` + a working
> CmdStan). The COVID series uses the aggregated Colombia data at the path in the
> `COVID_RDS` environment variable.

> **Note.** Earlier this section inlined a *simplified* reimplementation that
> fit only three own models, a single Epinowcast variant, and ran the own models
> through `backtest()` **without** the `X` covariates — which understated the
> diseasenowcasting models and inflated NobBS. `devel/benchmark_full.R` is the
> faithful, covariate-aware pipeline that reproduces the tables above.

## References