---
title: "Handling Outlier Delays with Censoring"
author: "diseasenowcasting package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling Outlier Delays with Censoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 4,
  collapse = TRUE, comment = "#>"
)
set.seed(1)
```

```{r load}
library(diseasenowcasting)
library(tbl.now)
library(dplyr)
```

## The problem: a single absurd delay

Real surveillance data contain the occasional report with an enormous reporting
delay — a case entered into the system months after it occurred.  In the
Colombia COVID-19 data (`covid_colombia`) the bulk of reports arrive within a
week or two, but a handful take **more than 300 days**:

```{r delay-hist, fig.cap="Reporting-delay distribution of COVID-19 Colombia. A few reports arrive hundreds of days late."}
library(ggplot2)
covid_colombia |>
  mutate(delay = as.numeric(diagnosis_date - notification_date)) |>
  filter(delay >= 0) |>
  ggplot(aes(delay)) +
  geom_histogram(binwidth = 5, fill = dn_palette()["reported"]) +
  scale_y_sqrt() +
  labs(x = "Reporting delay (days)", y = "Reports (sqrt scale)",
       title = "Most reports are quick; a long tail of extreme delays") +
  theme_diseasenowcasting()
```

When a parametric delay model (log-normal, gamma, …) is fit to data containing
such an outlier, the extreme value **drags the estimated delay distribution to
the right**.  The model then believes delays are longer than they really are,
inflates the most recent nowcasts, and scores worse.

The mathematical framework behind `diseasenowcasting` offers a principled fix
(Zepeda-Tello & Shaman, 2025, §"Missing report dates"): treat such a report as
**right-censored**.  Instead of telling the model "this case had delay exactly
330", we tell it only "this case arrived *by* delay 330" — i.e. its delay is an
**upper bound**.  In the likelihood the report then contributes
$\log G_D(330)$ (which is $\approx \log 1 = 0$ for any sensible delay
distribution) instead of the tiny tail density $\log \Delta G_D(330)$.  The
outlier stops distorting the fit.

## Step 1 — fit a model and discover the surprise

We work with an early-pandemic window and fit a one-stage nowcast.

```{r fit-window}
covid <- covid_colombia |>
  filter(notification_date <= as.Date("2020-07-15")) |>
  group_by(notification_date, diagnosis_date) |>
  summarise(n = sum(n), .groups = "drop")

tn <- tbl_now(covid,
              event_date  = notification_date,
              report_date = diagnosis_date,
              case_count  = n,
              data_type   = "count-incidence",
              verbose     = FALSE)

mdl <- model(nb_likelihood(), hsgp_epidemic(), lognormal_delay())
```

To play out the "new data arrives" story we first fit on the reports available
**as of** an early date, keeping only those reported by then:

```{r fit-asof}
asof  <- as.Date("2020-05-01")
tn_asof <- covid |>
  filter(diagnosis_date <= asof) |>
  tbl_now(event_date  = notification_date, report_date = diagnosis_date,
          case_count  = n, data_type = "count-incidence", verbose = FALSE)

nc <- nowcast(tn_asof, mdl, now = asof, type = "one_stage",
              n_draws = 500, temporal_effects = "none", seed = 1)
```

Now the next report arrives, and it carries an absurd 330-day delay.  `update()`
automatically scores the new report against the old fit and **warns** when
something is surprising:

```{r surprise, warning=TRUE}
new_batch <- data.frame(
  notification_date = as.Date("2020-05-02"),
  diagnosis_date    = as.Date("2020-05-02") + 330,  # an absurd 330-day delay
  n                 = 1                              # count-incidence needs the count column
)

nc_updated <- update(nc, new_batch, now = as.Date("2020-05-03"),
                     temporal_effects = "none")
```

The warning tells us *exactly* what was unexpected (a reporting delay far longer
than the model expects) and suggests the remedy.  The full table is available
via `surprise_result()`:

```{r surprise-table}
sr <- surprise_result(nc_updated)
subset(sr$delay_surprise, is_surprising)
```

## Step 2 — censor the outlier and re-fit

We follow the warning's advice: flag every report whose delay exceeds a sensible
bound (here 45 days, comfortably above the 99th percentile of ~50 days) as
censored.  `censor_delays_above()` does this — it sets `is_censored = TRUE` for
those reports, turning their delay into an upper bound.  `nowcast()` then reads
the `is_censored` flag automatically.

```{r censor}
tn_censored <- censor_delays_above(tn, max_delay = 45)
```

```{r refit}
nc_plain    <- nowcast(tn,          mdl, now = as.Date("2020-05-01"),
                       type = "one_stage", n_draws = 500,
                       temporal_effects = "none", seed = 1)
nc_censored <- nowcast(tn_censored, mdl, now = as.Date("2020-05-01"),
                       type = "one_stage", n_draws = 500,
                       temporal_effects = "none", seed = 1)
```

The fitted delay mean drops once the outliers are no longer taken literally:

```{r delay-compare}
data.frame(
  fit       = c("plain", "censored"),
  delay_mu  = round(c(coef(nc_plain)["delay_mu"], coef(nc_censored)["delay_mu"]), 3),
  delay_sd  = round(c(coef(nc_plain)["delay_sigma"], coef(nc_censored)["delay_sigma"]), 2)
)
```

## Step 3 — does it nowcast better?  A backtest

Finally we check that controlling the surprise actually *improves accuracy*.  We
backtest the same model on the plain data and on the censored data across a set
of as-of dates, scoring the d\* = 0 nowcast against the eventual truth with the
Weighted Interval Score (WIS; lower is better).

```{r backtest-compare}
eval_dates <- as.Date(c("2020-04-15", "2020-05-01", "2020-05-15", "2020-06-01"))

bt_plain <- backtest(tn,          mdl, dates = eval_dates, type = "one_stage",
                     n_draws = 300, max_delay = Inf, temporal_effects = "none", seed = 1)
bt_cens  <- backtest(tn_censored, mdl, dates = eval_dates, type = "one_stage",
                     n_draws = 300, max_delay = Inf, temporal_effects = "none", seed = 1)

rbind(
  plain    = score(bt_plain, report = FALSE)[, c("wis", "coverage_90")],
  censored = score(bt_cens,  report = FALSE)[, c("wis", "coverage_90")]
)
```

Censoring the handful of outlier delays gives a **lower (better) WIS** without
hurting coverage: the delay distribution is no longer pulled to the right by a
few extreme values, so the most recent nowcasts are better calibrated.

## Summary — the surprise → censor → refit loop

```
1. Fit a nowcast.
2. New data arrives -> update() scores it and WARNS about surprises.
3. A "delay too long" warning flags an outlier report.
4. censor_delays_above(tn, bound) turns its delay into an upper bound.
5. Re-fit -> the delay distribution is no longer distorted -> better WIS.
```

This is the practical payoff of modelling the reporting delay as a stochastic
process with proper censoring rather than a nuisance to be cleaned away: the
model can *tell you* when an observation does not fit, and the censoring
machinery lets you respond without discarding data.

> **Note.** The improvement is largest for the **one-stage** fit, whose joint
> likelihood is directly distorted by outlier delays.  The **two-stage** cascade
> estimates the delay from a recent window first and is already fairly robust to
> such outliers, so it benefits less.

## References

- Zepeda-Tello, R. & Shaman, J. (2025). Nowcasting from an epidemic's
  reporting-delay: a censored regression approach. *Preprint*.
- Günther, F. et al. (2021). Nowcasting the COVID-19 pandemic in Bavaria.
  *Biometrical Journal*.
