---
title: "diseasenowcasting vs NobBS and Epinowcast: Benchmark Comparison"
author: "diseasenowcasting package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diseasenowcasting vs NobBS and Epinowcast: Benchmark Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 9,
  fig.height = 4.5,
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

This vignette presents a systematic benchmark comparison of `diseasenowcasting` against two widely
used nowcasting methods: **NobBS** (McGough et al., 2020) and **Epinowcast** (Abbott et al., 2023).
We compare across three real-world disease surveillance datasets:

| Dataset | Disease | Temporal unit | Source |
|---------|---------|---------------|--------|
| `denguedat` (tbl.now) | Dengue fever | Weekly | Colombia |
| `mpoxdat` (tbl.now)   | Mpox | Daily | USA 2022 |
| `covid_colombia` (diseasenowcasting) | COVID-19 | Daily | Colombia 2020–2023 |

The comparison uses **50 evaluation dates** per disease, equally spaced across the
available history, chosen with a shared random seed for reproducibility.  Forecasts are
scored at **d* = 0** (the newest event time at each evaluation date) — the hardest and most
policy-relevant target.

---

## Scoring methodology

All methods are scored using the **Weighted Interval Score (WIS)**, a proper scoring rule
that simultaneously rewards calibration and sharpness (Bracher et al., 2021).  WIS decomposes
into:

- **Overprediction** — penalty for predicted intervals that are too high relative to the truth.
- **Underprediction** — penalty for predicted intervals that are too low.
- **Dispersion** — penalty for uncertainty intervals that are too wide.

Coverage at 50% and 90% credible intervals is also reported.  A well-calibrated model should
have empirical coverage close to the nominal level.

All comparisons are performed on a **common date set** — dates where both `diseasenowcasting` and the
comparison method have produced a valid nowcast — ensuring fairness.

---

## Results

```{r load-data}
library(diseasenowcasting)
library(ggplot2)

# Pre-computed benchmark results (50 evaluation dates per disease)
sc_file <- system.file("extdata", "comparison_scores.rds", package = "diseasenowcasting")
sc <- readRDS(sc_file)
```

### Dengue fever (weekly, Colombia)

```{r dengue-table}
dengue_tab <- sc[sc$disease == "dengue", ]
dengue_tab <- dengue_tab[, c("model","wis","over","under","disp","bias","cov50","cov90","n_dates")]
names(dengue_tab) <- c("Model","WIS","Overpred","Underpred","Dispersion","Bias","Cov50","Cov90","N dates")
dengue_tab$Model <- ifelse(dengue_tab$Model %in% names(table(dengue_tab$Model)),
                           dengue_tab$Model, dengue_tab$Model)
knitr::kable(dengue_tab, digits = 2, row.names = FALSE,
             caption = "Dengue nowcast scores at d*=0 (50 common evaluation dates, weekly).")
```

**Key findings (dengue):**

- All `diseasenowcasting` HSGP variants beat NobBS and all Epinowcast variants on WIS.
- Best `diseasenowcasting`: **HSGP / GenGamma**, WIS ≈ 7.0, cov90 ≈ 0.94.
- NobBS WIS ≈ `r round(sc$wis[sc$disease=="dengue" & grepl("NobBS",sc$model)][1], 1)`, cov90 ≈ `r round(sc$cov90[sc$disease=="dengue" & grepl("NobBS",sc$model)][1], 2)`.
- Best Epinowcast WIS ≈ `r round(min(sc$wis[sc$disease=="dengue" & grepl("Epinowcast",sc$model)], na.rm=TRUE), 1)`.

### Mpox (daily, USA)

```{r mpox-table}
mpox_tab <- sc[sc$disease == "mpox", c("model","wis","over","under","disp","bias","cov50","cov90","n_dates")]
names(mpox_tab) <- c("Model","WIS","Overpred","Underpred","Dispersion","Bias","Cov50","Cov90","N dates")
knitr::kable(mpox_tab, digits = 2, row.names = FALSE,
             caption = "Mpox nowcast scores at d*=0 (49 common evaluation dates, daily).")
```

**Key findings (mpox):**

- `diseasenowcasting` AR1 variants achieve the lowest WIS, beating NobBS by a wide margin (~7×).
- Best `diseasenowcasting`: **AR1 / Gamma**, WIS ≈ `r round(sc$wis[sc$disease=="mpox" & !sc$is_comparison][1], 1)`, cov90 ≈ `r round(sc$cov90[sc$disease=="mpox" & !sc$is_comparison][1], 2)`.
- NobBS WIS ≈ `r round(sc$wis[sc$disease=="mpox" & grepl("NobBS",sc$model)][1], 1)` — substantially worse due to poor calibration (cov90 ≈ `r round(sc$cov90[sc$disease=="mpox" & grepl("NobBS",sc$model)][1], 2)`).
- The mpox epidemic has a very short, sharp curve where the AR1 model's local-trend structure outperforms the global GP.

### COVID-19 (daily, Colombia)

```{r covid-table}
covid_tab <- sc[sc$disease == "covid", c("model","wis","over","under","disp","bias","cov50","cov90","n_dates")]
names(covid_tab) <- c("Model","WIS","Overpred","Underpred","Dispersion","Bias","Cov50","Cov90","N dates")
knitr::kable(covid_tab, digits = 2, row.names = FALSE,
             caption = "COVID-19 nowcast scores at d*=0 (50 common evaluation dates, daily).")
```

**Key findings (COVID-19):**

- `diseasenowcasting` beats NobBS on WIS for COVID as well, but the margin depends on the delay family.
- Best `diseasenowcasting`: **HSGP / GenGamma**, WIS ≈ `r round(sc$wis[sc$disease=="covid" & !sc$is_comparison][1], 0)`, cov90 ≈ `r round(sc$cov90[sc$disease=="covid" & !sc$is_comparison][1], 2)`.
- COVID's long daily series (700+ days) makes the GenGamma delay family advantageous: its flexible shape adapts to the changing delay distribution across epidemic waves.

---

## Summary across all diseases

```{r summary-plot, fig.cap="WIS comparison (d*=0): diseasenowcasting (HSGP best-per-disease) vs NobBS vs Epinowcast best. Lower is better."}
# Extract best diseasenowcasting per disease and comparison models
best_diseasenowcasting <- do.call(rbind, lapply(c("dengue","mpox","covid"), function(d) {
  sub <- sc[sc$disease == d & !sc$is_comparison, ]
  sub[which.min(sub$wis), ]
}))
best_diseasenowcasting$group <- "diseasenowcasting (best)"

cmp_rows <- sc[sc$is_comparison, ]
cmp_rows$group <- ifelse(grepl("NobBS", cmp_rows$model), "NobBS",
                         paste0("Epinowcast (", sub("Epinowcast \\((.*)\\)","\\1",cmp_rows$model), ")"))
# Best Epinowcast per disease
best_enw <- do.call(rbind, lapply(c("dengue","mpox","covid"), function(d) {
  sub <- cmp_rows[cmp_rows$disease == d & grepl("Epinowcast", cmp_rows$group), ]
  if (nrow(sub) == 0) return(NULL)
  sub[which.min(sub$wis), ]
}))

plot_data <- rbind(best_diseasenowcasting, cmp_rows[grepl("NobBS", cmp_rows$group), ], best_enw)
plot_data$disease_lab <- factor(plot_data$disease,
  levels = c("dengue","mpox","covid"),
  labels = c("Dengue\n(weekly)", "Mpox\n(daily)", "COVID-19\n(daily)"))
plot_data$group <- factor(plot_data$group,
  levels = c("diseasenowcasting (best)", "NobBS", sort(unique(plot_data$group[grepl("Epinowcast",plot_data$group)]))))

ggplot(plot_data, aes(x = disease_lab, y = wis, colour = group, group = group)) +
  geom_point(size = 3.5) +
  geom_line(aes(group = group), linetype = "dashed", alpha = 0.5) +
  scale_y_log10(labels = scales::comma) +
  scale_colour_manual(values = c("diseasenowcasting (best)" = "#e31a1c",
                                  "NobBS"          = "#1f78b4",
                                  "Epinowcast (weekly RE)" = "#33a02c",
                                  "Epinowcast (default)"  = "#b2df8a",
                                  "Epinowcast (rw)"       = "#a6cee3",
                                  "Epinowcast (point effect)" = "#fb9a99")) +
  labs(x = NULL, y = "WIS (log scale, lower = better)",
       colour = "Method", title = "Nowcast accuracy at d* = 0 (50 evaluation dates each)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

```{r summary-table}
summary_tab <- do.call(rbind, lapply(c("dengue","mpox","covid"), function(d) {
  sub <- sc[sc$disease == d, ]
  best_d <- sub[!sub$is_comparison, ][which.min(sub$wis[!sub$is_comparison]), ]
  nb     <- sub[grepl("NobBS", sub$model), ][1, ]
  enw    <- sub[grepl("Epinowcast", sub$model), ]; enw <- enw[which.min(enw$wis), ]
  data.frame(
    Disease = d,
    `Best diseasenowcasting` = best_d$model, `diseasenowcasting WIS` = best_d$wis, `diseasenowcasting cov90` = best_d$cov90,
    `NobBS WIS` = nb$wis, `NobBS cov90` = nb$cov90,
    `Best ENW WIS` = enw$wis, `Best ENW cov90` = enw$cov90,
    `diseasenowcasting beats NobBS` = best_d$wis < nb$wis,
    check.names = FALSE
  )
}))
knitr::kable(summary_tab, digits = 2, row.names = FALSE,
             caption = "Summary: best diseasenowcasting model per disease vs comparison methods.")
```

---

## How to reproduce with backtest()

The results above were generated by running `devel/run_own_models_diseasenowcasting.sh` (which
parallelises 50-date × 12-model fits via `own_models_diseasenowcasting.R`) and scoring with
`devel/score_diseasenowcasting.R`.  You can reproduce the same evaluation methodology at smaller scale
using `diseasenowcasting`'s built-in `backtest()` function.

`backtest()` fits one nowcast per (date × model) cell and dispatches those cells
in parallel via the [future](https://future.futureverse.org/) framework.  It is
sequential unless you register a parallel plan first — set
`future::plan(multisession, workers = N)` before calling `backtest()` to use
`N` CPU cores (and `future::plan(sequential)` to restore serial execution).

```{r reproduce-code, eval=FALSE}
library(diseasenowcasting)
library(tbl.now)
library(future)

# Set up parallel execution (adjust workers to your CPU count)
plan(multisession, workers = 8)

# ── Dengue (weekly) ──────────────────────────────────────────────────────────
dn <- tbl_now(denguedat, event_date = onset_week, report_date = report_week,
              data_type = "linelist", verbose = FALSE)

models_to_compare <- list(
  model(nb_likelihood(), hsgp_epidemic(), lognormal_delay()),
  model(nb_likelihood(), hsgp_epidemic(), generalized_gamma_delay()),
  model(nb_likelihood(), ar1_epidemic(),  lognormal_delay()),
  model(nb_likelihood(), sir_epidemic(),  lognormal_delay())
)

# 10 evaluation dates (use 50 for the full benchmark)
set.seed(8765)
dates_dengue <- sort(sample(seq(as.Date("1993-01-01"), as.Date("2010-04-01"), by="week"), 10))

bt_dengue <- backtest(
  dn,
  models   = models_to_compare,
  dates    = dates_dengue,
  type     = "two_stage",
  K        = 25,
  n_draws  = 200,
  seed     = 42
)

score(bt_dengue, metric = "wis", report = TRUE)
autoplot(bt_dengue)
```

To add NobBS and Epinowcast as comparison baselines, format their quantile outputs with
the same `(date_run, .event_num, q2.5…q97.5, final)` column schema and bind them to the
`@results` slot of the `backtest_class` object before calling `score()`.

---

## References

- McGough, S. F. et al. (2020). NobBS: Nowcasting by Bayesian Smoothing. *PLOS Computational Biology*.
- Abbott, S. et al. (2023). Epinowcast: Flexible hierarchical nowcasting. *Journal of Open Source Software*.
- Bracher, J. et al. (2021). Evaluating epidemic forecasts in an interval format. *PLOS Computational Biology*.
- Zepeda-Tello, R. & Shaman, J. Nowcasting from an epidemic's reporting-delay: a censored regression approach. *Preprint*.
