---
title: "Custom Delays and Epidemic Processes"
author: "Rodrigo Zepeda-Tello"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Delays and Epidemic Processes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.align  = "center",
  fig.height = 4,
  collapse   = TRUE,
  comment    = "#>"
)
```

`diseasenowcasting` ships four built-in reporting-delay families (LogNormal, Gamma, Generalized Gamma, Dirichlet). With `custom_delay()` you can plug in **any** parametric distribution whose CDF you can write in RTMB-traceable R.

## Requirements: RTMB-traceable CDF factories

The only constraint comes from RTMB's automatic differentiation (AD) tape: every arithmetic operation inside `cdf_factory` must be AD-safe.

**Allowed:** `+ - * /`, `exp`, `log`, `sqrt`, `pnorm`, `pgamma`, `lgamma`, `abs`, `sum`, scalar multiplication, fixed-length `for` loops (length must not depend on parameter values).

**Not allowed inside the tape:** `if`/`ifelse` on parameter values, `pmax`/`pmin` on AD types (replace `pmax(x, 0)` with `(x + abs(x)) / 2`), external ODE solvers, any non-differentiable function.

## Setup

```{r load}
library(diseasenowcasting)
library(ggplot2)
set.seed(2025)
```

## Example 1 — Weibull delay via `custom_delay()`

The Weibull CDF is $F(d) = 1 - \exp(-(d/\lambda)^k)$.  We parameterise on the log scale so both parameters are unconstrained:

```{r weibull-factory}
weibull_cdf_factory <- function(theta) {
  shape <- exp(theta[1])   # k > 0
  scale <- exp(theta[2])   # lambda > 0
  list(
    cdf          = function(d) 1 - exp(-(d / scale)^shape),
    log_cdf      = function(d) log(1 - exp(-(d / scale)^shape)),
    log_survival = function(d) -(d / scale)^shape
  )
}
```

We then wrap it in `custom_delay()`:

```{r weibull-delay}
weibull_delay_spec <- custom_delay(
  cdf_factory = weibull_cdf_factory,
  n_params    = 2L,
  priors      = list(
    normal_prior(0, 1),          # log_shape ~ N(0,1)
    normal_prior(log(7), 1)      # log_scale ~ N(log 7, 1)
  ),
  name        = "Weibull",
  param_names = c("log_shape", "log_scale"),
  inits       = c(0, log(7))
)
```

Use `validate_custom_delay()` to confirm the factory is RTMB-traceable before fitting:

> **Note** This validator is experimental and some functions that
are AD-safe will not validate. If you know what you are doing 
you can skip this step.

```{r validate}
validate_custom_delay(weibull_delay_spec)
```

## Preparing the data

We generate synthetic case data with a known Weibull(shape = 2, scale = 7) delay so
the fitted parameters are verifiable:

```{r data-prep}
n_times     <- 40L
true_shape  <- 2.0
true_scale  <- 7.0
true_lambda <- 80 * exp(0.04 * seq_len(n_times))  # growing epidemic

rows <- list()
for (t in seq_len(n_times)) {
  n_cases <- rpois(1, true_lambda[t])
  if (n_cases == 0L) next
  delays      <- pmin(ceiling(rweibull(n_cases, shape = true_shape, scale = true_scale)), 21L)
  report_t    <- t + delays
  keep        <- report_t <= n_times
  if (!any(keep)) next
  delays_kept <- delays[keep]
  agg <- tapply(rep(1L, sum(keep)), delays_kept, sum)
  for (d in names(agg)) {
    rows[[length(rows) + 1L]] <- c(event_time = t, count = agg[[d]], delay = as.integer(d))
  }
}
m_mat <- do.call(rbind, rows)  # columns: event_time, count, delay
```

## Fitting with the custom Weibull delay

We build a model, pass `weibull_delay_spec` as the delay, and call `fit()` directly
on the prepared data (no `tbl_now` required for the low-level API):

```{r fit-weibull}
mod_weibull <- model(
  likelihood = nb_likelihood(),
  epidemic   = ar1_epidemic(),
  delay      = weibull_delay_spec
)

data_weibull  <- prepare_data(mod_weibull, m = m_mat, max_time = n_times)
priors_weibull <- default_priors(mod_weibull, data_weibull)
fit_w <- fit(mod_weibull, data_weibull, priors_weibull)
```

The custom parameters live in `fit$parList$custom_delay_params`:

```{r weibull-params}
theta_w <- as.numeric(fit_w$parList$custom_delay_params)

cat("Fitted Weibull (true: shape = 2, scale = 7):\n")
cat("  shape =", round(exp(theta_w[1]), 3), "\n")
cat("  scale =", round(exp(theta_w[2]), 3), "days\n")
```

## Example 2 — Generalized Gamma with Q fixed to 1

The Generalized Gamma family (built-in family 3) reduces to **Weibull** exactly when its
shape parameter $Q = 1$.  Fixing $Q = 1$ via `generalized_gamma_delay(Q = 1)` therefore
fits the same distribution — a useful sanity check for the custom implementation.

```{r fit-gengamma}
mod_gengamma <- model(
  likelihood = nb_likelihood(),
  epidemic   = ar1_epidemic(),
  delay      = generalized_gamma_delay(Q = 1)   # Q = 1 => Weibull
)

data_gengamma   <- prepare_data(mod_gengamma, m = m_mat, max_time = n_times)
priors_gengamma <- default_priors(mod_gengamma, data_gengamma)
fit_g <- fit(mod_gengamma, data_gengamma, priors_gengamma)
```

In the GenGamma (flexsurv) parametrisation with $Q = 1$:

$$F(d) = \mathrm{pgamma}\!\left(\exp\!\left(\frac{\log d - \mu}{\sigma}\right),\; 1,\; 1\right)
       = 1 - \exp\!\left(-e^{(\log d - \mu)/\sigma}\right)$$

Matching to the Weibull $1 - \exp(-(d/\lambda)^k)$ gives $\mu = \log\lambda$ and
$\sigma = 1/k$.

```{r gengamma-params}
mu_g   <- fit_g$reconstruct$delay_mu
sig_g  <- fit_g$reconstruct$delay_sigma

cat("Fitted GenGamma (Q=1), converted to Weibull (true: shape = 2, scale = 7):\n")
cat("  shape =", round(1 / sig_g, 3), "\n")
cat("  scale =", round(exp(mu_g), 3), "days\n")
```

## Comparing the two fits

Both approaches model the same distribution.  Their delay densities should overlap closely:

```{r compare-plot, fig.cap="Fitted delay distributions: custom Weibull vs. built-in GenGamma(Q=1). Because Q=1 is exactly the Weibull distribution, the two curves should nearly coincide."}
delay_grid <- seq(0.5, 30, by = 0.5)

# Custom Weibull PMF via pweibull (base R)
shape_w <- exp(theta_w[1])
scale_w <- exp(theta_w[2])
pmf_w <- stats::pweibull(delay_grid + 0.5, shape = shape_w, scale = scale_w) -
         stats::pweibull(pmax(0, delay_grid - 0.5), shape = shape_w, scale = scale_w)

# GenGamma(Q=1) is Weibull with shape = 1/sigma, scale = exp(mu)
shape_g <- 1 / sig_g
scale_g <- exp(mu_g)
pmf_g <- stats::pweibull(delay_grid + 0.5, shape = shape_g, scale = scale_g) -
         stats::pweibull(pmax(0, delay_grid - 0.5), shape = shape_g, scale = scale_g)

compare_df <- data.frame(
  delay   = rep(delay_grid, 2),
  density = c(pmf_w, pmf_g),
  model   = rep(c("Custom Weibull", "GenGamma (Q=1)"), each = length(delay_grid))
)

ggplot(compare_df, aes(x = delay, y = density, colour = model, linetype = model)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c("Custom Weibull" = "#2166AC", "GenGamma (Q=1)" = "#D6604D")) +
  labs(x = "Reporting delay (days)", y = "Probability mass",
       title = "Custom Weibull vs. built-in GenGamma (Q = 1)",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

The two curves overlap because $Q = 1$ in the GenGamma parametrisation is algebraically identical to the Weibull.  Any small difference reflects numerical precision and the different prior specifications.

## Delay distribution comparison

The two fitted delay densities should nearly coincide because they model the
same distribution:

```{r nowcast-compare, fig.height = 4, fig.cap="Fitted delay distributions: custom Weibull vs. built-in GenGamma(Q=1). The two curves overlap because Q=1 is exactly the Weibull distribution."}
delay_grid <- seq(0.5, 25, by = 0.5)

# Custom Weibull PMF
shape_w <- exp(theta_w[1])
scale_w <- exp(theta_w[2])
pmf_w2  <- stats::pweibull(delay_grid + 0.5, shape = shape_w, scale = scale_w) -
            stats::pweibull(pmax(0, delay_grid - 0.5), shape = shape_w, scale = scale_w)

# GenGamma(Q=1) converts to Weibull: shape = 1/sigma, scale = exp(mu)
pmf_g2  <- stats::pweibull(delay_grid + 0.5, shape = 1/sig_g, scale = exp(mu_g)) -
            stats::pweibull(pmax(0, delay_grid - 0.5), shape = 1/sig_g, scale = exp(mu_g))

# True distribution
pmf_true <- stats::pweibull(delay_grid + 0.5, shape = true_shape, scale = true_scale) -
             stats::pweibull(pmax(0, delay_grid - 0.5), shape = true_shape, scale = true_scale)

compare_df2 <- data.frame(
  delay   = rep(delay_grid, 3),
  density = c(pmf_w2, pmf_g2, pmf_true),
  model   = rep(c("Custom Weibull", "GenGamma (Q=1)", "True Weibull(2,7)"), each = length(delay_grid))
)

ggplot(compare_df2, aes(x = delay, y = density, colour = model, linetype = model)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(
    values = c("Custom Weibull" = "#2166AC", "GenGamma (Q=1)" = "#D6604D",
               "True Weibull(2,7)" = "black")) +
  scale_linetype_manual(
    values = c("Custom Weibull" = "solid", "GenGamma (Q=1)" = "dashed",
               "True Weibull(2,7)" = "dotted")) +
  labs(x = "Reporting delay (days)", y = "Probability mass",
       title = "Custom Weibull vs. GenGamma (Q=1) vs. truth",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## Tips for writing your own delay factory

1. **Parameterise on the unconstrained scale.** Use `exp(theta[i])` for strictly positive parameters, `plogis(theta[i])` for bounded-[0,1] parameters.  The `inits` argument should be on the same unconstrained scale.

2. **Guard against `log(0)`.** Add a small epsilon: `log(cdf_value + 1e-300)`.  Likewise for `log_survival`.

3. **`log_survival` for large delays.** The discretised delay log-likelihood uses `log_survival` for observations beyond the empirical median, where `1 - F` is close to zero and numerically stable.  For the Weibull, `log(1 - F(d)) = -(d/lambda)^k` is available in closed form.

4. **Validate before fitting.** `validate_custom_delay()` test-tapes the factory and checks that both `fn()` and `gr()` are finite at the init values.

5. **Fix parameters.** Pass a length-1 numeric in `priors` to fix that parameter:
   ```r
   custom_delay(factory, n_params = 2L,
                priors = list(normal_prior(0, 1), 2.0),  # scale fixed at 2
                ...)
   ```

---

# Custom Epidemic Processes

The framework is **epidemic-process agnostic**: the reporting-delay model only ever needs one thing from the epidemic side — the log expected-incidence trajectory `log_mean[t, s]` for each event-time `t` and stratum `s`.  How that trajectory is generated does not matter: it can come from a regression, a discrete recursion, an ODE model, or anything else.

`custom_process()` gives you a single seam to plug in any such generator.

## The interface: `intensity_fn(theta) → log_mean[T × S]`

You write a function that:

1. Takes a numeric parameter vector `theta`.
2. Returns a numeric **matrix** with `n_time` rows and `n_strata` columns containing `log` expected incidence (already on the natural log scale — the framework applies its own soft-cap and links to the likelihood on top).

The function must be written in RTMB-traceable arithmetic (same rules as for delay factories, plus one additional note about index assignment):

**AD-safe operations:** `+`, `-`, `*`, `/`, `exp`, `log`, `sqrt`, `abs`, `sum`, `cumsum`, scalar/matrix arithmetic, fixed-length `for` loops.

**Index assignment in loops:** if you assign into a pre-allocated vector inside a `for` loop (`lm[t] <- ...`), you must put `'[<-' <- RTMB::ADoverload('[<-')` at the very start of `intensity_fn`.  Alternatively, prefer **`cumsum()`** or `Reduce()` for accumulation — these avoid the issue entirely.

## Example 1 — Pure random walk

A pure random walk on log-incidence:

$$\log \lambda_t = \log \lambda_{t-1} + \sigma \cdot \varepsilon_t, \quad \varepsilon_t \sim \mathcal{N}(0,1)$$

This differs structurally from the built-in AR(1): there is no mean-reversion parameter $\phi$ — the trajectory drifts freely.  The parameter vector is `(log_mu0, log_sigma, eps_1, …, eps_T)`.

The key: we use `cumsum()` so no index assignment is needed.

```{r rw-setup}
# Simulate a random-walk epidemic
set.seed(2024)
n_times     <- 40L
true_mu0    <- 3.0    # log initial incidence
true_sigma  <- 0.15   # step size on log scale
log_lambda_true <- true_mu0 + cumsum(c(0, rnorm(n_times - 1L, 0, true_sigma)))
lambda_true <- exp(log_lambda_true)

# Generate observations with a LogNormal(log 5, 0.7) reporting delay
event_idx <- rep(seq_len(n_times), rpois(n_times, lambda_true))
delays_rw <- pmax(1L, round(rlnorm(length(event_idx), meanlog = log(5), sdlog = 0.7)))
m_rw <- cbind(event_idx, 1L, delays_rw)

cat("RW epidemic: mean incidence", round(mean(lambda_true), 1),
    "| range", paste(round(range(lambda_true), 1), collapse = "–"), "\n")
```

Now define `intensity_fn` using `cumsum()`:

```{r rw-process}
# n_times is captured from the environment above — no need to pass it
rw_intensity_fn <- function(theta) {
  log_mu0  <- theta[1]
  sigma_rw <- exp(theta[2])
  eps      <- theta[3:(2L + n_times)]          # n_times innovations
  log_mean <- log_mu0 + cumsum(sigma_rw * eps) # cumsum: no [<- needed
  matrix(log_mean, n_times, 1L)
}

# n_params = 2 structural + n_times innovations
rw_proc <- custom_process(
  intensity_fn = rw_intensity_fn,
  n_params     = 2L + n_times,
  priors       = c(
    list(normal_prior(3, 1)),          # log_mu0
    list(normal_prior(-2, 0.5)),       # log_sigma_rw (log scale)
    rep(list(std_normal_prior()), n_times)   # innovations eps_1 ... eps_T
  ),
  name        = "RandomWalk",
  param_names = c("log_mu0", "log_sigma", paste0("eps_", seq_len(n_times))),
  inits       = c(3, log(0.15), rep(0, n_times))
)
validate_custom_process(rw_proc)
```

The `n_params` must be set before fitting.  A natural way to find `n_times` without re-running the model: call `prepare_data()` once with a placeholder process and read `max_time` from the result.

### Fitting

```{r rw-fit}
mod_rw <- model(nb_likelihood(), rw_proc, lognormal_delay())
dat_rw <- prepare_data(mod_rw, m = m_rw, max_time = n_times)
pr_rw  <- default_priors(mod_rw, dat_rw)
fit_rw <- fit(mod_rw, dat_rw, pr_rw)
```

Compare against the HSGP (built-in flexible smoother) on the same data:

```{r hsgp-fit-rw}
mod_hsgp  <- model(nb_likelihood(), hsgp_epidemic(), lognormal_delay())
dat_hsgp  <- prepare_data(mod_hsgp, m = m_rw, max_time = n_times)
pr_hsgp   <- default_priors(mod_hsgp, dat_hsgp)
fit_hsgp  <- fit(mod_hsgp, dat_hsgp, pr_hsgp)
```

### Comparing the fits

```{r rw-compare-plot, fig.height=4, fig.cap="Random walk vs. HSGP on the same data. Both recover the underlying trend; the custom random walk uses a simpler structural assumption."}
# Reconstruct log_mean for both fits
lm_rw   <- fit_rw$reconstruct$mu_safe
lm_hsgp <- fit_hsgp$reconstruct$mu_safe

plot_df <- data.frame(
  t         = rep(seq_len(n_times), 3),
  log_mean  = c(log_lambda_true, lm_rw[, 1], lm_hsgp[, 1]),
  model     = rep(c("True RW", "Custom RW", "Built-in HSGP"), each = n_times)
)

ggplot(plot_df, aes(x = t, y = exp(log_mean), colour = model, linetype = model)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("True RW" = "black",
                                  "Custom RW" = "#2166AC",
                                  "Built-in HSGP" = "#D6604D")) +
  scale_linetype_manual(values = c("True RW" = "dotted",
                                    "Custom RW" = "solid",
                                    "Built-in HSGP" = "dashed")) +
  labs(x = "Event time", y = "Expected incidence (lambda)",
       title = "Custom random walk vs. built-in HSGP",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

## Example 2 — SIR ODE (hand-rolled Euler integration)

A compartmental SIR model:

$$\frac{dS}{dt} = -\beta S I / N, \quad \frac{dI}{dt} = \beta S I / N - \gamma I$$

with observed incidence $\lambda_t = \Delta S_t = \beta S_t I_t / N$ at each step.

We implement this as a **discrete-time Euler scheme** inside `intensity_fn`.  Because the loop uses index assignment, we add `'[<-' <- RTMB::ADoverload('[<-')` at the top of the function.

Parameters: `(log_R0, log_gamma, log_I0)` — all on the log scale so they remain positive.

```{r sir-data}
# Simulate SIR epidemic
set.seed(2024)
N_pop <- 10000L; R0_sir <- 2.5; gamma_sir <- 0.1; I0_sir <- 5L
S <- N_pop - I0_sir; I <- I0_sir
n_sir <- 60L
sir_incidence <- numeric(n_sir)
for (t in seq_len(n_sir)) {
  beta_val <- R0_sir * gamma_sir / N_pop
  new_inf  <- beta_val * S * I
  sir_incidence[t] <- max(new_inf, 0)
  S <- S - new_inf; I <- I + new_inf - gamma_sir * I
}
# Observed counts
sir_counts <- rpois(n_sir, sir_incidence)

# Build m matrix (Poisson delay)
event_idx_sir <- rep(seq_len(n_sir), sir_counts)
delays_sir <- pmax(1L, rpois(length(event_idx_sir), lambda = 4))
m_sir <- cbind(event_idx_sir, 1L, delays_sir)
cat("SIR peak at t =", which.max(sir_counts), "| max counts =", max(sir_counts), "\n")
```

```{r sir-process}
sir_intensity_fn <- function(theta) {
  # Must declare [<- overload at the top to enable loop-based vector building
  '[<-' <- RTMB::ADoverload('[<-')

  R0_val  <- exp(theta[1])
  gamma_r <- exp(theta[2])
  I0_val  <- exp(theta[3])
  N       <- 10000

  S_t <- N - I0_val; I_t <- I0_val
  incidence <- numeric(n_sir)   # pre-allocate; filled via [<- inside loop

  for (t in seq_len(n_sir)) {
    beta_t  <- R0_val * gamma_r / N
    new_inf <- beta_t * S_t * I_t
    incidence[t] <- new_inf
    S_t <- S_t - new_inf
    I_t <- I_t + new_inf - gamma_r * I_t
  }

  # pmax-safe guard: (x + |x|)/2 replaces pmax(x, 0) for AD types
  log_mean <- log((incidence + abs(incidence)) * 0.5 + 1e-8)
  matrix(log_mean, n_sir, 1L)
}

sir_proc <- custom_process(
  intensity_fn = sir_intensity_fn,
  n_params     = 3L,
  priors       = list(
    normal_prior(log(2.5), 0.5),   # log_R0
    normal_prior(log(0.1), 0.3),   # log_gamma
    normal_prior(log(5),   1.0)    # log_I0
  ),
  name        = "SIR_ODE",
  param_names = c("log_R0", "log_gamma", "log_I0"),
  inits       = c(log(2), log(0.1), log(5))
)
validate_custom_process(sir_proc)
```

### Fitting and comparison with the built-in SIR

```{r sir-fit}
mod_sir_custom <- model(nb_likelihood(), sir_proc, lognormal_delay())
dat_sir_custom <- prepare_data(mod_sir_custom, m = m_sir, max_time = n_sir)
pr_sir_custom  <- default_priors(mod_sir_custom, dat_sir_custom)
fit_sir_custom <- fit(mod_sir_custom, dat_sir_custom, pr_sir_custom)
```

The built-in `sir_epidemic()` uses a mathematically equivalent discrete-time SIR with an additional AR(1) trend on $\beta(t)$:

```{r sir-builtin-fit}
mod_sir_builtin <- model(
  nb_likelihood(),
  sir_epidemic(R0 = numeric(0), gamma = numeric(0),
               N_pop = 10000, use_beta_rw_trend = FALSE),
  lognormal_delay()
)
dat_sir_builtin <- prepare_data(mod_sir_builtin, m = m_sir, max_time = n_sir)
pr_sir_builtin  <- default_priors(mod_sir_builtin, dat_sir_builtin)
fit_sir_builtin <- fit(mod_sir_builtin, dat_sir_builtin, pr_sir_builtin)
```

Extract and compare the fitted SIR parameters:

```{r sir-params}
# Custom SIR
cp <- as.numeric(fit_sir_custom$parList$custom_process_params)
cat("Custom SIR:\n")
cat("  R0    =", round(exp(cp[1]), 3), "(true:", R0_sir, ")\n")
cat("  gamma =", round(exp(cp[2]), 3), "(true:", gamma_sir, ")\n")
cat("  I0    =", round(exp(cp[3]), 1), "(true:", I0_sir, ")\n\n")

# Built-in SIR
cat("Built-in SIR:\n")
cat("  R0    =", round(exp(fit_sir_builtin$parList$log_R0), 3), "(true:", R0_sir, ")\n")
gamma_builtin <- stats::plogis(as.numeric(fit_sir_builtin$parList$u_gamma))
cat("  gamma =", round(gamma_builtin, 3), "(true:", gamma_sir, ")\n")
```

Both should recover the true parameters; any small discrepancy comes from the AR(1) trend in the built-in version (which we disabled with `use_beta_rw_trend = FALSE`).

```{r sir-compare-plot, fig.height=4, fig.cap="Custom SIR ODE vs. built-in sir_epidemic() on identical data. Both recover the epidemic curve; the custom process gives identical fits with only three lines of ODE code."}
lm_sir_custom  <- fit_sir_custom$reconstruct$mu_safe
lm_sir_builtin <- fit_sir_builtin$reconstruct$mu_safe

sir_plot_df <- data.frame(
  t        = rep(seq_len(n_sir), 3),
  lambda   = c(sir_incidence, exp(lm_sir_custom[, 1]), exp(lm_sir_builtin[, 1])),
  model    = rep(c("True SIR", "Custom SIR ODE", "Built-in sir_epidemic()"), each = n_sir)
)

ggplot(sir_plot_df, aes(x = t, y = lambda, colour = model, linetype = model)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("True SIR" = "black",
                                  "Custom SIR ODE" = "#2166AC",
                                  "Built-in sir_epidemic()" = "#D6604D")) +
  scale_linetype_manual(values = c("True SIR" = "dotted",
                                    "Custom SIR ODE" = "solid",
                                    "Built-in sir_epidemic()" = "dashed")) +
  labs(x = "Event time", y = "Expected incidence",
       title = "Custom SIR ODE vs. built-in sir_epidemic()",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

## Tips for writing your own `intensity_fn`

1. **Return the full `log_mean`, not just a trend.** Unlike HSGP/AR1 (which return a trend added to an intercept), `custom_process` is responsible for the entire `log_mean[T × S]` matrix — include any baseline/intercept in your function.

2. **Parameterise on the unconstrained scale.** Use `exp(theta[i])` for positive parameters (R0, sigma, scale), `plogis(theta[i])` for probabilities.  Provide `inits` on the same unconstrained scale.

3. **Know your `n_time` before constructing the process.** The innovations in a random walk have length `n_time`.  Check this first with a helper call:
   ```r
   n_times <- prepare_data(model(nb_likelihood(), hsgp_epidemic(), lognormal_delay()),
                           m = your_m_matrix)$max_time
   ```

4. **Accumulation without `[<-`:** use `cumsum(sigma * eps)` for a random walk; `Reduce(f, 1:n, accumulate = TRUE)` for more complex recursions — both are AD-safe without any overloading.

5. **Accumulation with `[<-`:** if you need a pre-allocated vector in a `for` loop (e.g. an ODE), put `'[<-' <- RTMB::ADoverload('[<-')` at the **very first line** of `intensity_fn`.

6. **Guard `log` of near-zero values.** Incidence can approach zero in early/late epidemic phases.  Use `log((x + abs(x)) * 0.5 + 1e-8)` — the `(x + abs(x)) / 2` is the AD-safe `pmax(x, 0)`.

7. **Validate before fitting.** `validate_custom_process()` test-tapes your function and checks finite gradients.  Call it right after defining the process.

8. **RTMBode for production ODE models.** This vignette uses a hand-rolled Euler scheme.  For stiff systems or when adaptive step-size control matters, the `RTMBode` package provides a full Sundials/CVODES adjoint solver that integrates cleanly with RTMB.  Add it to your project with `install.packages("RTMBode")` and consult its documentation for the `ode()` interface.
