---
title: "Custom Delays and Epidemic Processes"
author: "Rodrigo Zepeda-Tello"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Delays and Epidemic Processes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.align  = "center",
  fig.height = 4,
  collapse   = TRUE,
  comment    = "#>"
)
```

`diseasenowcasting` ships four built-in reporting-delay families (LogNormal, Gamma, Generalized Gamma, Dirichlet) and three epidemic processes (HSGP, AR(1), SIR). This vignette shows how to go beyond the menu and supply **your own** delay distribution or **your own** epidemic model, by writing a short R function. No knowledge of the package internals is needed — only basic R and the model you have in mind.

## How custom components work (and the one rule you must follow)

To fit a model, the package needs the *gradient* (the vector of partial derivatives) of the likelihood with respect to your parameters. It obtains this gradient **automatically** from your function — a technique called *automatic differentiation*, provided by the [RTMB](https://github.com/kaskr/RTMB) package. You never compute a derivative yourself.

The price for this convenience is a single rule:

> **Your function must be written so RTMB can differentiate it.** In practice this means using only ordinary arithmetic and a short list of math functions, and avoiding anything that "branches" on the value of a parameter.

Concretely:

**Allowed:** `+`, `-`, `*`, `/`, `exp`, `log`, `sqrt`, `abs`, `sum`, `cumsum`, `pnorm`, `pgamma`, `lgamma`, matrix arithmetic, and `for` loops whose **length is fixed** (does not depend on a parameter).

**Not allowed:** `if` / `ifelse` that test the *value* of a parameter, `pmax` / `pmin` on parameters (use the identity `pmax(x, 0) = (x + abs(x)) / 2` instead), random number draws, calls to external solvers (`deSolve`, optimisers), or any non-smooth function.

Each constructor comes with a **validator** (`validate_custom_delay()`, `validate_custom_process()`) that checks this rule for you *before* you fit — always run it first; it turns a cryptic failure deep in the optimiser into a clear message.

## Setup

```{r load}
library(diseasenowcasting)
library(RTMB)        # REQUIRED for custom components — see note below
library(ggplot2)
set.seed(2025)
```

> **Why `library(RTMB)`?** The built-in models work with just `library(diseasenowcasting)`. But a *custom* function you write lives in your own workspace, and its arithmetic (`+`, `exp`, `cumsum`, …) only connects to RTMB's auto-differentiation machinery when RTMB is attached to your session. The package keeps RTMB out of the way by default (so it does not silently replace functions like `dnorm`), so **you must run `library(RTMB)` yourself whenever you use `custom_delay()` or `custom_process()`.** If you forget, the validator will remind you.

## Example 1 — Weibull delay via `custom_delay()`

**What you must provide.** A *CDF factory*: a function of a parameter vector `theta` that returns a list of three functions of a delay `d`:

| Element | Meaning | Formula for the Weibull |
|---|---|---|
| `cdf(d)` | $F(d) = \Pr(\text{delay} \le d)$ | $1 - \exp(-(d/\lambda)^k)$ |
| `log_cdf(d)` | $\log F(d)$ | $\log\!\left(1 - \exp(-(d/\lambda)^k)\right)$ |
| `log_survival(d)` | $\log(1 - F(d))$ | $-(d/\lambda)^k$ |

The package supplies `log_cdf` and `log_survival` separately (rather than just `cdf`) because each is numerically stable in a different regime — `log_survival` stays accurate far out in the tail where `1 - F(d)` underflows to zero.

The Weibull CDF is $F(d) = 1 - \exp(-(d/\lambda)^k)$ with shape $k$ and scale $\lambda$. We work on the **log scale** (`theta[1] = log k`, `theta[2] = log \lambda`) so the parameters are unconstrained — RTMB optimises over the whole real line, and `exp()` maps back to positive values:

```{r weibull-factory}
weibull_cdf_factory <- function(theta) {
  shape <- exp(theta[1])   # k > 0
  scale <- exp(theta[2])   # lambda > 0
  list(
    cdf          = function(d) 1 - exp(-(d / scale)^shape),
    log_cdf      = function(d) log(1 - exp(-(d / scale)^shape)),
    log_survival = function(d) -(d / scale)^shape
  )
}
```

We then wrap it in `custom_delay()`. Each argument:

- **`cdf_factory`** — the function above.
- **`n_params`** — how many entries `theta` has (here 2: log-shape and log-scale).
- **`priors`** — a list of length `n_params`. Each entry is **either** a prior (e.g. `normal_prior(0, 1)`) for a parameter to be *estimated*, **or** a single number to *fix* that parameter at a constant. The order must match `theta`.
- **`name`**, **`param_names`** — labels used in printing and diagnostics.
- **`inits`** — starting values for the optimiser, on the same (unconstrained) scale as `theta`.

```{r weibull-delay}
weibull_delay_spec <- custom_delay(
  cdf_factory = weibull_cdf_factory,
  n_params    = 2L,
  priors      = list(
    normal_prior(0, 1),          # log_shape ~ N(0,1)
    normal_prior(log(7), 1)      # log_scale ~ N(log 7, 1)
  ),
  name        = "Weibull",
  param_names = c("log_shape", "log_scale"),
  inits       = c(0, log(7))
)
```

Use `validate_custom_delay()` to confirm the factory is differentiable before fitting. It test-runs your factory through RTMB and checks the result and its gradient are finite, reporting a clear message on success or failure:

> **Note.** The check is deliberately conservative: a few genuinely AD-safe functions may not pass. If you are confident your factory follows the rule above, a failed check is not necessarily fatal — but for most users it catches real mistakes early.

```{r validate}
validate_custom_delay(weibull_delay_spec)
```

## Preparing the data

We generate synthetic case data with a known Weibull(shape = 2, scale = 7) delay so
the fitted parameters are verifiable. For these low-level examples we feed the model a plain matrix `m` with **three columns** — `(event_time, count, delay)` — one row per (event-time, delay) cell: how many cases occurred at `event_time` and were reported `delay` time-units later. (In normal use you would instead build a `tbl_now` from a line-list and call `nowcast()`; here we use the matrix API to keep the example self-contained.)

```{r data-prep}
n_times     <- 40L
true_shape  <- 2.0
true_scale  <- 7.0
true_lambda <- 80 * exp(0.04 * seq_len(n_times))  # growing epidemic

rows <- list()
for (t in seq_len(n_times)) {
  n_cases <- rpois(1, true_lambda[t])
  if (n_cases == 0L) next
  delays      <- pmin(ceiling(rweibull(n_cases, shape = true_shape, scale = true_scale)), 21L)
  report_t    <- t + delays
  keep        <- report_t <= n_times
  if (!any(keep)) next
  delays_kept <- delays[keep]
  agg <- tapply(rep(1L, sum(keep)), delays_kept, sum)
  for (d in names(agg)) {
    rows[[length(rows) + 1L]] <- c(event_time = t, count = agg[[d]], delay = as.integer(d))
  }
}
m_mat <- do.call(rbind, rows)  # columns: event_time, count, delay
```

## Fitting with the custom Weibull delay

We build a model, pass `weibull_delay_spec` as the delay, and call `fit()` directly
on the prepared data (no `tbl_now` required for the low-level API):

```{r fit-weibull}
mod_weibull <- model(
  likelihood = nb_likelihood(),
  epidemic   = ar1_epidemic(),
  delay      = weibull_delay_spec
)

data_weibull  <- prepare_data(mod_weibull, m = m_mat, max_time = n_times)
priors_weibull <- default_priors(mod_weibull, data_weibull)
fit_w <- fit(mod_weibull, data_weibull, priors_weibull)
```

The custom parameters live in `fit$parList$custom_delay_params`:

```{r weibull-params}
theta_w <- as.numeric(fit_w$parList$custom_delay_params)

cat("Fitted Weibull (true: shape = 2, scale = 7):\n")
cat("  shape =", round(exp(theta_w[1]), 3), "\n")
cat("  scale =", round(exp(theta_w[2]), 3), "days\n")
```

## Example 2 — Generalized Gamma with Q fixed to 1

The Generalized Gamma family (built-in family 3) reduces to **Weibull** exactly when its
shape parameter $Q = 1$.  Fixing $Q = 1$ via `generalized_gamma_delay(Q = 1)` therefore
fits the same distribution — a useful sanity check for the custom implementation.

```{r fit-gengamma}
mod_gengamma <- model(
  likelihood = nb_likelihood(),
  epidemic   = ar1_epidemic(),
  delay      = generalized_gamma_delay(Q = 1)   # Q = 1 => Weibull
)

data_gengamma   <- prepare_data(mod_gengamma, m = m_mat, max_time = n_times)
priors_gengamma <- default_priors(mod_gengamma, data_gengamma)
fit_g <- fit(mod_gengamma, data_gengamma, priors_gengamma)
```

In the GenGamma (flexsurv) parametrisation with $Q = 1$:

$$F(d) = \mathrm{pgamma}\!\left(\exp\!\left(\frac{\log d - \mu}{\sigma}\right),\; 1,\; 1\right)
       = 1 - \exp\!\left(-e^{(\log d - \mu)/\sigma}\right)$$

Matching to the Weibull $1 - \exp(-(d/\lambda)^k)$ gives $\mu = \log\lambda$ and
$\sigma = 1/k$.

```{r gengamma-params}
mu_g   <- fit_g$reconstruct$delay_mu
sig_g  <- fit_g$reconstruct$delay_sigma

cat("Fitted GenGamma (Q=1), converted to Weibull (true: shape = 2, scale = 7):\n")
cat("  shape =", round(1 / sig_g, 3), "\n")
cat("  scale =", round(exp(mu_g), 3), "days\n")
```

## Comparing the two fits

Both approaches model the same distribution.  Their delay densities should overlap closely:

```{r compare-plot, fig.cap="Fitted delay distributions: custom Weibull vs. built-in GenGamma(Q=1). Because Q=1 is exactly the Weibull distribution, the two curves should nearly coincide."}
delay_grid <- seq(0.5, 30, by = 0.5)

# Custom Weibull PMF via pweibull (base R)
shape_w <- exp(theta_w[1])
scale_w <- exp(theta_w[2])
pmf_w <- stats::pweibull(delay_grid + 0.5, shape = shape_w, scale = scale_w) -
         stats::pweibull(pmax(0, delay_grid - 0.5), shape = shape_w, scale = scale_w)

# GenGamma(Q=1) is Weibull with shape = 1/sigma, scale = exp(mu)
shape_g <- 1 / sig_g
scale_g <- exp(mu_g)
pmf_g <- stats::pweibull(delay_grid + 0.5, shape = shape_g, scale = scale_g) -
         stats::pweibull(pmax(0, delay_grid - 0.5), shape = shape_g, scale = scale_g)

compare_df <- data.frame(
  delay   = rep(delay_grid, 2),
  density = c(pmf_w, pmf_g),
  model   = rep(c("Custom Weibull", "GenGamma (Q=1)"), each = length(delay_grid))
)

ggplot(compare_df, aes(x = delay, y = density, colour = model, linetype = model)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c("Custom Weibull" = "#2166AC", "GenGamma (Q=1)" = "#D6604D")) +
  labs(x = "Reporting delay (days)", y = "Probability mass",
       title = "Custom Weibull vs. built-in GenGamma (Q = 1)",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

The two curves overlap because $Q = 1$ in the GenGamma parametrisation is algebraically identical to the Weibull.  Any small difference reflects numerical precision and the different prior specifications.

## Delay distribution comparison

The two fitted delay densities should nearly coincide because they model the
same distribution:

```{r nowcast-compare, fig.height = 4, fig.cap="Fitted delay distributions: custom Weibull vs. built-in GenGamma(Q=1). The two curves overlap because Q=1 is exactly the Weibull distribution."}
delay_grid <- seq(0.5, 25, by = 0.5)

# Custom Weibull PMF
shape_w <- exp(theta_w[1])
scale_w <- exp(theta_w[2])
pmf_w2  <- stats::pweibull(delay_grid + 0.5, shape = shape_w, scale = scale_w) -
            stats::pweibull(pmax(0, delay_grid - 0.5), shape = shape_w, scale = scale_w)

# GenGamma(Q=1) converts to Weibull: shape = 1/sigma, scale = exp(mu)
pmf_g2  <- stats::pweibull(delay_grid + 0.5, shape = 1/sig_g, scale = exp(mu_g)) -
            stats::pweibull(pmax(0, delay_grid - 0.5), shape = 1/sig_g, scale = exp(mu_g))

# True distribution
pmf_true <- stats::pweibull(delay_grid + 0.5, shape = true_shape, scale = true_scale) -
             stats::pweibull(pmax(0, delay_grid - 0.5), shape = true_shape, scale = true_scale)

compare_df2 <- data.frame(
  delay   = rep(delay_grid, 3),
  density = c(pmf_w2, pmf_g2, pmf_true),
  model   = rep(c("Custom Weibull", "GenGamma (Q=1)", "True Weibull(2,7)"), each = length(delay_grid))
)

ggplot(compare_df2, aes(x = delay, y = density, colour = model, linetype = model)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(
    values = c("Custom Weibull" = "#2166AC", "GenGamma (Q=1)" = "#D6604D",
               "True Weibull(2,7)" = "black")) +
  scale_linetype_manual(
    values = c("Custom Weibull" = "solid", "GenGamma (Q=1)" = "dashed",
               "True Weibull(2,7)" = "dotted")) +
  labs(x = "Reporting delay (days)", y = "Probability mass",
       title = "Custom Weibull vs. GenGamma (Q=1) vs. truth",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## Tips for writing your own delay factory

1. **Parameterise on the unconstrained scale.** Use `exp(theta[i])` for strictly positive parameters, `plogis(theta[i])` for bounded-[0,1] parameters.  The `inits` argument should be on the same unconstrained scale.

2. **Guard against `log(0)`.** Add a small epsilon: `log(cdf_value + 1e-300)`.  Likewise for `log_survival`.

3. **`log_survival` for large delays.** The discretised delay log-likelihood uses `log_survival` for observations beyond the empirical median, where `1 - F` is close to zero and numerically stable.  For the Weibull, `log(1 - F(d)) = -(d/lambda)^k` is available in closed form.

4. **Validate before fitting.** `validate_custom_delay()` test-tapes the factory and checks that both `fn()` and `gr()` are finite at the init values.

5. **Fix parameters.** Pass a length-1 numeric in `priors` to fix that parameter:
   ```r
   custom_delay(factory, n_params = 2L,
                priors = list(normal_prior(0, 1), 2.0),  # scale fixed at 2
                ...)
   ```

---

# Custom Epidemic Processes

The framework is **epidemic-process agnostic**: the reporting-delay model only ever needs one thing from the epidemic side — the log expected-incidence trajectory `log_mean[t, s]` for each event-time `t` and stratum `s`.  How that trajectory is generated does not matter: it can come from a regression, a discrete recursion, an ODE model, or anything else.

`custom_process()` gives you a single seam to plug in any such generator.

## The interface: `intensity_fn(theta) → log_mean[T × S]`

**What you must provide.** A single function — call it `intensity_fn` — that:

1. Takes a numeric parameter vector `theta`.
2. Returns a numeric **matrix** with one row per event-time (`n_time` rows) and one column per stratum (`n_strata` columns). Each entry is the **log of the expected number of cases** at that time and stratum, i.e. $\log \lambda_{t,s}$.

That is the *whole* contract. Everything downstream — the reporting-delay model, censoring, the count likelihood — is shared and handled for you. You only describe how the expected case curve is generated. (If you have a single, unstratified series, the matrix has one column: `matrix(log_lambda, n_time, 1L)`.)

Note: return the **entire** log-mean, including any baseline level — unlike the built-in HSGP/AR(1), which add a trend on top of a separate intercept, a custom process owns the full curve.

The function must follow the same AD rule as a delay factory, with one extra wrinkle about filling a vector inside a loop:

**AD-safe operations:** `+`, `-`, `*`, `/`, `exp`, `log`, `sqrt`, `abs`, `sum`, `cumsum`, scalar/matrix arithmetic, fixed-length `for` loops.

**Filling a vector inside a `for` loop.** If you build a result by assigning into a pre-allocated vector (`lambda[t] <- ...`), add this one line at the **top** of `intensity_fn`:
```r
`[<-` <- RTMB::ADoverload("[<-")
```
This makes index-assignment differentiable. If you can avoid the loop with **`cumsum()`** (as in the random walk below) you don't need it at all — prefer that when you can.

## Example 1 — Pure random walk

A pure random walk on log-incidence:

$$\log \lambda_t = \log \lambda_{t-1} + \sigma \cdot \varepsilon_t, \quad \varepsilon_t \sim \mathcal{N}(0,1)$$

This differs structurally from the built-in AR(1): there is no mean-reversion parameter $\phi$ — the trajectory drifts freely.  The parameter vector is `(log_mu0, log_sigma, eps_1, …, eps_T)`.

The key: we use `cumsum()` so no index assignment is needed.

```{r rw-setup}
# Simulate a random-walk epidemic
set.seed(2024)
n_times     <- 40L
true_mu0    <- 3.0    # log initial incidence
true_sigma  <- 0.15   # step size on log scale
log_lambda_true <- true_mu0 + cumsum(c(0, rnorm(n_times - 1L, 0, true_sigma)))
lambda_true <- exp(log_lambda_true)

# Generate observations with a LogNormal(log 5, 0.7) reporting delay
event_idx <- rep(seq_len(n_times), rpois(n_times, lambda_true))
delays_rw <- pmax(1L, round(rlnorm(length(event_idx), meanlog = log(5), sdlog = 0.7)))
m_rw <- cbind(event_idx, 1L, delays_rw)

cat("RW epidemic: mean incidence", round(mean(lambda_true), 1),
    "| range", paste(round(range(lambda_true), 1), collapse = "–"), "\n")
```

Now define `intensity_fn` using `cumsum()`:

```{r rw-process}
# n_times is captured from the environment above — no need to pass it
rw_intensity_fn <- function(theta) {
  log_mu0  <- theta[1]
  sigma_rw <- exp(theta[2])
  eps      <- theta[3:(2L + n_times)]          # n_times innovations
  log_mean <- log_mu0 + cumsum(sigma_rw * eps) # cumsum: no [<- needed
  matrix(log_mean, n_times, 1L)
}

# n_params = 2 structural + n_times innovations
rw_proc <- custom_process(
  intensity_fn = rw_intensity_fn,
  n_params     = 2L + n_times,
  priors       = c(
    list(normal_prior(3, 1)),          # log_mu0
    list(normal_prior(-2, 0.5)),       # log_sigma_rw (log scale)
    rep(list(std_normal_prior()), n_times)   # innovations eps_1 ... eps_T
  ),
  name        = "RandomWalk",
  param_names = c("log_mu0", "log_sigma", paste0("eps_", seq_len(n_times))),
  inits       = c(3, log(0.15), rep(0, n_times))
)
validate_custom_process(rw_proc)
```

The `custom_process()` arguments mirror `custom_delay()`:

- **`intensity_fn`** — the function above.
- **`n_params`** — the length of `theta`. Here it is `2 + n_times`: two structural parameters (`log_mu0`, `log_sigma`) **plus one innovation $\varepsilon_t$ per time point**. The random-walk path *is* its innovations, so the model estimates `n_time` of them.
- **`priors`** — a list of length `n_params`, in the same order as `theta`; each entry is a prior (to estimate) or a number (to fix). Here the two structural parameters get informative priors and every innovation gets `std_normal_prior()` (the standard random-walk assumption $\varepsilon_t \sim \mathcal N(0,1)$).
- **`name`**, **`param_names`**, **`inits`** — labels and optimiser starting values, as before.

> **Getting `n_times` right.** Because the innovation count depends on the length of your series, you need `n_time` *before* building the process. The simplest way is to let `prepare_data()` tell you, using any built-in process as a placeholder:
>
> ```r
> n_times <- prepare_data(
>   model(nb_likelihood(), hsgp_epidemic(), lognormal_delay()),
>   m = your_data
> )$max_time
> ```
>
> Then build `custom_process()` with `n_params = 2L + n_times` and fit on the same data.

### Fitting

```{r rw-fit}
mod_rw <- model(nb_likelihood(), rw_proc, lognormal_delay())
dat_rw <- prepare_data(mod_rw, m = m_rw, max_time = n_times)
pr_rw  <- default_priors(mod_rw, dat_rw)
fit_rw <- fit(mod_rw, dat_rw, pr_rw)
```

Compare against the HSGP (built-in flexible smoother) on the same data:

```{r hsgp-fit-rw}
mod_hsgp  <- model(nb_likelihood(), hsgp_epidemic(), lognormal_delay())
dat_hsgp  <- prepare_data(mod_hsgp, m = m_rw, max_time = n_times)
pr_hsgp   <- default_priors(mod_hsgp, dat_hsgp)
fit_hsgp  <- fit(mod_hsgp, dat_hsgp, pr_hsgp)
```

### Comparing the fits

```{r rw-compare-plot, fig.height=4, fig.cap="Random walk vs. HSGP on the same data. Both recover the underlying trend; the custom random walk uses a simpler structural assumption."}
# Reconstruct log_mean for both fits
lm_rw   <- fit_rw$reconstruct$mu_safe
lm_hsgp <- fit_hsgp$reconstruct$mu_safe

plot_df <- data.frame(
  t         = rep(seq_len(n_times), 3),
  log_mean  = c(log_lambda_true, lm_rw[, 1], lm_hsgp[, 1]),
  model     = rep(c("True RW", "Custom RW", "Built-in HSGP"), each = n_times)
)

ggplot(plot_df, aes(x = t, y = exp(log_mean), colour = model, linetype = model)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("True RW" = "black",
                                  "Custom RW" = "#2166AC",
                                  "Built-in HSGP" = "#D6604D")) +
  scale_linetype_manual(values = c("True RW" = "dotted",
                                    "Custom RW" = "solid",
                                    "Built-in HSGP" = "dashed")) +
  labs(x = "Event time", y = "Expected incidence (lambda)",
       title = "Custom random walk vs. built-in HSGP",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

## Example 2 — SIR ODE (hand-rolled Euler integration)

A compartmental SIR model:

$$\frac{dS}{dt} = -\beta S I / N, \quad \frac{dI}{dt} = \beta S I / N - \gamma I$$

with observed incidence $\lambda_t = \Delta S_t = \beta S_t I_t / N$ at each step.

We implement this as a **discrete-time Euler scheme** inside `intensity_fn`.  Because the loop uses index assignment, we add `'[<-' <- RTMB::ADoverload('[<-')` at the top of the function.

Parameters: `(log_R0, log_gamma, log_I0)` — all on the log scale so they remain positive.

```{r sir-data}
# Simulate SIR epidemic
set.seed(2024)
N_pop <- 10000L; R0_sir <- 2.5; gamma_sir <- 0.1; I0_sir <- 5L
S <- N_pop - I0_sir; I <- I0_sir
n_sir <- 60L
sir_incidence <- numeric(n_sir)
for (t in seq_len(n_sir)) {
  beta_val <- R0_sir * gamma_sir / N_pop
  new_inf  <- beta_val * S * I
  sir_incidence[t] <- max(new_inf, 0)
  S <- S - new_inf; I <- I + new_inf - gamma_sir * I
}
# Observed counts
sir_counts <- rpois(n_sir, sir_incidence)

# Build m matrix (Poisson delay)
event_idx_sir <- rep(seq_len(n_sir), sir_counts)
delays_sir <- pmax(1L, rpois(length(event_idx_sir), lambda = 4))
m_sir <- cbind(event_idx_sir, 1L, delays_sir)
cat("SIR peak at t =", which.max(sir_counts), "| max counts =", max(sir_counts), "\n")
```

```{r sir-process}
sir_intensity_fn <- function(theta) {
  # Must declare [<- overload at the top to enable loop-based vector building
  '[<-' <- RTMB::ADoverload('[<-')

  R0_val  <- exp(theta[1])
  gamma_r <- exp(theta[2])
  I0_val  <- exp(theta[3])
  N       <- 10000

  S_t <- N - I0_val; I_t <- I0_val
  incidence <- numeric(n_sir)   # pre-allocate; filled via [<- inside loop

  for (t in seq_len(n_sir)) {
    beta_t  <- R0_val * gamma_r / N
    new_inf <- beta_t * S_t * I_t
    incidence[t] <- new_inf
    S_t <- S_t - new_inf
    I_t <- I_t + new_inf - gamma_r * I_t
  }

  # pmax-safe guard: (x + |x|)/2 replaces pmax(x, 0) for AD types
  log_mean <- log((incidence + abs(incidence)) * 0.5 + 1e-8)
  matrix(log_mean, n_sir, 1L)
}

sir_proc <- custom_process(
  intensity_fn = sir_intensity_fn,
  n_params     = 3L,
  priors       = list(
    normal_prior(log(2.5), 0.5),   # log_R0
    normal_prior(log(0.1), 0.3),   # log_gamma
    normal_prior(log(5),   1.0)    # log_I0
  ),
  name        = "SIR_ODE",
  param_names = c("log_R0", "log_gamma", "log_I0"),
  inits       = c(log(2), log(0.1), log(5))
)
validate_custom_process(sir_proc)
```

### Fitting and comparison with the built-in SIR

```{r sir-fit}
mod_sir_custom <- model(nb_likelihood(), sir_proc, lognormal_delay())
dat_sir_custom <- prepare_data(mod_sir_custom, m = m_sir, max_time = n_sir)
pr_sir_custom  <- default_priors(mod_sir_custom, dat_sir_custom)
fit_sir_custom <- fit(mod_sir_custom, dat_sir_custom, pr_sir_custom)
```

The built-in `sir_epidemic()` uses a mathematically equivalent discrete-time SIR with an additional AR(1) trend on $\beta(t)$:

```{r sir-builtin-fit}
mod_sir_builtin <- model(
  nb_likelihood(),
  sir_epidemic(R0 = numeric(0), gamma = numeric(0),
               N_pop = 10000, use_beta_rw_trend = FALSE),
  lognormal_delay()
)
dat_sir_builtin <- prepare_data(mod_sir_builtin, m = m_sir, max_time = n_sir)
pr_sir_builtin  <- default_priors(mod_sir_builtin, dat_sir_builtin)
fit_sir_builtin <- fit(mod_sir_builtin, dat_sir_builtin, pr_sir_builtin)
```

Extract and compare the fitted SIR parameters:

```{r sir-params}
# Custom SIR
cp <- as.numeric(fit_sir_custom$parList$custom_process_params)
cat("Custom SIR:\n")
cat("  R0    =", round(exp(cp[1]), 3), "(true:", R0_sir, ")\n")
cat("  gamma =", round(exp(cp[2]), 3), "(true:", gamma_sir, ")\n")
cat("  I0    =", round(exp(cp[3]), 1), "(true:", I0_sir, ")\n\n")

# Built-in SIR
cat("Built-in SIR:\n")
cat("  R0    =", round(exp(fit_sir_builtin$parList$log_R0), 3), "(true:", R0_sir, ")\n")
gamma_builtin <- stats::plogis(as.numeric(fit_sir_builtin$parList$u_gamma))
cat("  gamma =", round(gamma_builtin, 3), "(true:", gamma_sir, ")\n")
```

Both fits land near the truth, but notice that $R_0$ and $\gamma$ individually are only *weakly* identified from a single growing epidemic: what the early curve really pins down is their **product** $\beta = R_0\,\gamma$ (the transmission rate that sets the growth speed). So you may see, say, $R_0$ a little high and $\gamma$ a little low while $\beta = R_0\gamma$ stays close to the true $`r round(R0_sir * gamma_sir, 3)`$. This is a property of the SIR model and the data, not of the custom-process machinery — the built-in `sir_epidemic()` shows the same behaviour. The fitted **incidence curves**, which are what the data actually constrain, match closely:

```{r sir-compare-plot, fig.height=4, fig.cap="Custom SIR ODE vs. built-in sir_epidemic() on identical data. Both recover the epidemic curve; the custom process gives identical fits with only three lines of ODE code."}
lm_sir_custom  <- fit_sir_custom$reconstruct$mu_safe
lm_sir_builtin <- fit_sir_builtin$reconstruct$mu_safe

sir_plot_df <- data.frame(
  t        = rep(seq_len(n_sir), 3),
  lambda   = c(sir_incidence, exp(lm_sir_custom[, 1]), exp(lm_sir_builtin[, 1])),
  model    = rep(c("True SIR", "Custom SIR ODE", "Built-in sir_epidemic()"), each = n_sir)
)

ggplot(sir_plot_df, aes(x = t, y = lambda, colour = model, linetype = model)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("True SIR" = "black",
                                  "Custom SIR ODE" = "#2166AC",
                                  "Built-in sir_epidemic()" = "#D6604D")) +
  scale_linetype_manual(values = c("True SIR" = "dotted",
                                    "Custom SIR ODE" = "solid",
                                    "Built-in sir_epidemic()" = "dashed")) +
  labs(x = "Event time", y = "Expected incidence",
       title = "Custom SIR ODE vs. built-in sir_epidemic()",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

---

## Tips for writing your own `intensity_fn`

1. **Return the full `log_mean`, not just a trend.** Unlike HSGP/AR1 (which return a trend added to an intercept), `custom_process` is responsible for the entire `log_mean[T × S]` matrix — include any baseline/intercept in your function.

2. **Parameterise on the unconstrained scale.** Use `exp(theta[i])` for positive parameters (R0, sigma, scale), `plogis(theta[i])` for probabilities.  Provide `inits` on the same unconstrained scale.

3. **Know your `n_time` before constructing the process.** The innovations in a random walk have length `n_time`.  Check this first with a helper call:
   ```r
   n_times <- prepare_data(model(nb_likelihood(), hsgp_epidemic(), lognormal_delay()),
                           m = your_m_matrix)$max_time
   ```

4. **Accumulation without `[<-`:** use `cumsum(sigma * eps)` for a random walk; `Reduce(f, 1:n, accumulate = TRUE)` for more complex recursions — both are AD-safe without any overloading.

5. **Accumulation with `[<-`:** if you need a pre-allocated vector in a `for` loop (e.g. an ODE), put `'[<-' <- RTMB::ADoverload('[<-')` at the **very first line** of `intensity_fn`.

6. **Guard `log` of near-zero values.** Incidence can approach zero in early/late epidemic phases.  Use `log((x + abs(x)) * 0.5 + 1e-8)` — the `(x + abs(x)) / 2` is the AD-safe `pmax(x, 0)`.

7. **Validate before fitting.** `validate_custom_process()` test-tapes your function and checks finite gradients.  Call it right after defining the process.

8. **RTMBode for production ODE models.** This vignette uses a hand-rolled Euler scheme.  For stiff systems or when adaptive step-size control matters, the `RTMBode` package provides a full Sundials/CVODES adjoint solver that integrates cleanly with RTMB.  Add it to your project with `install.packages("RTMBode")` and consult its documentation for the `ode()` interface.
