---
title: "Custom Delay Distributions"
author: "Rodrigo Zepeda-Tello"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Delay Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.align  = "center",
  fig.height = 4,
  collapse   = TRUE,
  comment    = "#>"
)
```

`diseasenowcasting` ships four built-in reporting-delay families (LogNormal, Gamma, Generalized Gamma, Dirichlet). With `custom_delay()` you can plug in **any** parametric distribution whose CDF you can write in RTMB-traceable R.

## Requirements: RTMB-traceable CDF factories

The only constraint comes from RTMB's automatic differentiation (AD) tape: every arithmetic operation inside `cdf_factory` must be AD-safe.

**Allowed:** `+ - * /`, `exp`, `log`, `sqrt`, `pnorm`, `pgamma`, `lgamma`, `abs`, `sum`, scalar multiplication, fixed-length `for` loops (length must not depend on parameter values).

**Not allowed inside the tape:** `if`/`ifelse` on parameter values, `pmax`/`pmin` on AD types (replace `pmax(x, 0)` with `(x + abs(x)) / 2`), external ODE solvers, any non-differentiable function.

## Setup

```{r load}
library(diseasenowcasting)
library(ggplot2)
set.seed(2025)
```

## Example 1 — Weibull delay via `custom_delay()`

The Weibull CDF is $F(d) = 1 - \exp(-(d/\lambda)^k)$.  We parameterise on the log scale so both parameters are unconstrained:

```{r weibull-factory}
weibull_cdf_factory <- function(theta) {
  shape <- exp(theta[1])   # k > 0
  scale <- exp(theta[2])   # lambda > 0
  list(
    cdf          = function(d) 1 - exp(-(d / scale)^shape),
    log_cdf      = function(d) log(1 - exp(-(d / scale)^shape) + 1e-300),
    log_survival = function(d) -(d / scale)^shape
  )
}
```

We then wrap it in `custom_delay()`:

```{r weibull-delay}
weibull_delay_spec <- custom_delay(
  cdf_factory = weibull_cdf_factory,
  n_params    = 2L,
  priors      = list(
    normal_prior(0, 1),          # log_shape ~ N(0,1)
    normal_prior(log(7), 1)      # log_scale ~ N(log 7, 1)
  ),
  name        = "Weibull",
  param_names = c("log_shape", "log_scale"),
  inits       = c(0, log(7))
)
```

Use `validate_custom_delay()` to confirm the factory is RTMB-traceable before fitting:

```{r validate}
validate_custom_delay(weibull_delay_spec)
```

## Preparing the data

We generate synthetic case data with a known Weibull(shape = 2, scale = 7) delay so
the fitted parameters are verifiable:

```{r data-prep}
n_times     <- 40L
true_shape  <- 2.0
true_scale  <- 7.0
true_lambda <- 80 * exp(0.04 * seq_len(n_times))  # growing epidemic

rows <- list()
for (t in seq_len(n_times)) {
  n_cases <- rpois(1, true_lambda[t])
  if (n_cases == 0L) next
  delays      <- pmin(ceiling(rweibull(n_cases, shape = true_shape, scale = true_scale)), 21L)
  report_t    <- t + delays
  keep        <- report_t <= n_times
  if (!any(keep)) next
  delays_kept <- delays[keep]
  agg <- tapply(rep(1L, sum(keep)), delays_kept, sum)
  for (d in names(agg)) {
    rows[[length(rows) + 1L]] <- c(event_time = t, count = agg[[d]], delay = as.integer(d))
  }
}
m_mat <- do.call(rbind, rows)  # columns: event_time, count, delay
```

## Fitting with the custom Weibull delay

We build a model, pass `weibull_delay_spec` as the delay, and call `fit()` directly
on the prepared data (no `tbl_now` required for the low-level API):

```{r fit-weibull}
mod_weibull <- model(
  likelihood = nb_likelihood(),
  epidemic   = ar1_epidemic(),
  delay      = weibull_delay_spec
)

data_weibull  <- prepare_data(mod_weibull, m = m_mat, max_time = n_times)
priors_weibull <- default_priors(mod_weibull, data_weibull)
fit_w <- fit(mod_weibull, data_weibull, priors_weibull)
```

The custom parameters live in `fit$parList$custom_delay_params`:

```{r weibull-params}
theta_w <- as.numeric(fit_w$parList$custom_delay_params)

cat("Fitted Weibull (true: shape = 2, scale = 7):\n")
cat("  shape =", round(exp(theta_w[1]), 3), "\n")
cat("  scale =", round(exp(theta_w[2]), 3), "days\n")
```

## Example 2 — Generalized Gamma with Q fixed to 1

The Generalized Gamma family (built-in family 3) reduces to **Weibull** exactly when its
shape parameter $Q = 1$.  Fixing $Q = 1$ via `generalized_gamma_delay(Q = 1)` therefore
fits the same distribution — a useful sanity check for the custom implementation.

```{r fit-gengamma}
mod_gengamma <- model(
  likelihood = nb_likelihood(),
  epidemic   = ar1_epidemic(),
  delay      = generalized_gamma_delay(Q = 1)   # Q = 1 => Weibull
)

data_gengamma   <- prepare_data(mod_gengamma, m = m_mat, max_time = n_times)
priors_gengamma <- default_priors(mod_gengamma, data_gengamma)
fit_g <- fit(mod_gengamma, data_gengamma, priors_gengamma)
```

In the GenGamma (flexsurv) parametrisation with $Q = 1$:

$$F(d) = \mathrm{pgamma}\!\left(\exp\!\left(\frac{\log d - \mu}{\sigma}\right),\; 1,\; 1\right)
       = 1 - \exp\!\left(-e^{(\log d - \mu)/\sigma}\right)$$

Matching to the Weibull $1 - \exp(-(d/\lambda)^k)$ gives $\mu = \log\lambda$ and
$\sigma = 1/k$.

```{r gengamma-params}
mu_g   <- fit_g$reconstruct$delay_mu
sig_g  <- fit_g$reconstruct$delay_sigma

cat("Fitted GenGamma (Q=1), converted to Weibull (true: shape = 2, scale = 7):\n")
cat("  shape =", round(1 / sig_g, 3), "\n")
cat("  scale =", round(exp(mu_g), 3), "days\n")
```

## Comparing the two fits

Both approaches model the same distribution.  Their delay densities should overlap closely:

```{r compare-plot, fig.cap="Fitted delay distributions: custom Weibull vs. built-in GenGamma(Q=1). Because Q=1 is exactly the Weibull distribution, the two curves should nearly coincide."}
delay_grid <- seq(0.5, 30, by = 0.5)

# Custom Weibull PMF via pweibull (base R)
shape_w <- exp(theta_w[1])
scale_w <- exp(theta_w[2])
pmf_w <- stats::pweibull(delay_grid + 0.5, shape = shape_w, scale = scale_w) -
         stats::pweibull(pmax(0, delay_grid - 0.5), shape = shape_w, scale = scale_w)

# GenGamma(Q=1) is Weibull with shape = 1/sigma, scale = exp(mu)
shape_g <- 1 / sig_g
scale_g <- exp(mu_g)
pmf_g <- stats::pweibull(delay_grid + 0.5, shape = shape_g, scale = scale_g) -
         stats::pweibull(pmax(0, delay_grid - 0.5), shape = shape_g, scale = scale_g)

compare_df <- data.frame(
  delay   = rep(delay_grid, 2),
  density = c(pmf_w, pmf_g),
  model   = rep(c("Custom Weibull", "GenGamma (Q=1)"), each = length(delay_grid))
)

ggplot(compare_df, aes(x = delay, y = density, colour = model, linetype = model)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c("Custom Weibull" = "#2166AC", "GenGamma (Q=1)" = "#D6604D")) +
  labs(x = "Reporting delay (days)", y = "Probability mass",
       title = "Custom Weibull vs. built-in GenGamma (Q = 1)",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

The two curves overlap because $Q = 1$ in the GenGamma parametrisation is algebraically identical to the Weibull.  Any small difference reflects numerical precision and the different prior specifications.

## Delay distribution comparison

The two fitted delay densities should nearly coincide because they model the
same distribution:

```{r nowcast-compare, fig.height = 4, fig.cap="Fitted delay distributions: custom Weibull vs. built-in GenGamma(Q=1). The two curves overlap because Q=1 is exactly the Weibull distribution."}
delay_grid <- seq(0.5, 25, by = 0.5)

# Custom Weibull PMF
shape_w <- exp(theta_w[1])
scale_w <- exp(theta_w[2])
pmf_w2  <- stats::pweibull(delay_grid + 0.5, shape = shape_w, scale = scale_w) -
            stats::pweibull(pmax(0, delay_grid - 0.5), shape = shape_w, scale = scale_w)

# GenGamma(Q=1) converts to Weibull: shape = 1/sigma, scale = exp(mu)
pmf_g2  <- stats::pweibull(delay_grid + 0.5, shape = 1/sig_g, scale = exp(mu_g)) -
            stats::pweibull(pmax(0, delay_grid - 0.5), shape = 1/sig_g, scale = exp(mu_g))

# True distribution
pmf_true <- stats::pweibull(delay_grid + 0.5, shape = true_shape, scale = true_scale) -
             stats::pweibull(pmax(0, delay_grid - 0.5), shape = true_shape, scale = true_scale)

compare_df2 <- data.frame(
  delay   = rep(delay_grid, 3),
  density = c(pmf_w2, pmf_g2, pmf_true),
  model   = rep(c("Custom Weibull", "GenGamma (Q=1)", "True Weibull(2,7)"), each = length(delay_grid))
)

ggplot(compare_df2, aes(x = delay, y = density, colour = model, linetype = model)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(
    values = c("Custom Weibull" = "#2166AC", "GenGamma (Q=1)" = "#D6604D",
               "True Weibull(2,7)" = "black")) +
  scale_linetype_manual(
    values = c("Custom Weibull" = "solid", "GenGamma (Q=1)" = "dashed",
               "True Weibull(2,7)" = "dotted")) +
  labs(x = "Reporting delay (days)", y = "Probability mass",
       title = "Custom Weibull vs. GenGamma (Q=1) vs. truth",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## Tips for writing your own factory

1. **Parameterise on the unconstrained scale.** Use `exp(theta[i])` for strictly positive parameters, `plogis(theta[i])` for bounded-[0,1] parameters.  The `inits` argument should be on the same unconstrained scale.

2. **Guard against `log(0)`.** Add a small epsilon: `log(cdf_value + 1e-300)`.  Likewise for `log_survival`.

3. **`log_survival` for large delays.** The discretised delay log-likelihood uses `log_survival` for observations beyond the empirical median, where `1 - F` is close to zero and numerically stable.  For the Weibull, `log(1 - F(d)) = -(d/lambda)^k` is available in closed form.

4. **Validate before fitting.** `validate_custom_delay()` test-tapes the factory and checks that both `fn()` and `gr()` are finite at the init values.

5. **Fix parameters.** Pass a length-1 numeric in `priors` to fix that parameter:
   ```r
   custom_delay(factory, n_params = 2L,
                priors = list(normal_prior(0, 1), 2.0),  # scale fixed at 2
                ...)
   ```
