---
title: "Custom Delays and Epidemic Processes"
author: "Rodrigo Zepeda-Tello"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Delays and Epidemic Processes}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.align  = "center",
  fig.height = 4,
  collapse   = TRUE,
  comment    = "#>"
)
# Code-folding for plotting chunks (chunks tagged `fold = TRUE`):
#   * HTML  -> the source is wrapped in a collapsed <details> block (closed by
#     default), so readers compare the plots without wading through ggplot code;
#   * PDF    -> the plotting source is omitted entirely (only the figure shows).
# The CUSTOM code (factories / intensity functions) is never folded — that is
# the part that matters.
local({
  default_source <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    out <- default_source(x, options)
    if (!isTRUE(options$fold)) return(out)
    if (knitr::is_html_output())
      paste0("<details><summary>Show plotting code</summary>\n\n", out, "\n\n</details>")
    else ""
  })
})
```

`diseasenowcasting` is a **model-agnostic nowcasting framework**. A nowcast here is built from three interchangeable pieces — a **likelihood**, an **epidemic process**, and a **reporting-delay distribution** — and you can either pick a built-in for each or **supply your own**. Most tools (e.g. `epinowcast`, `baselinenowcast`) give you a fixed model; here the model is yours to assemble or extend.

This vignette shows the two extension points, each on a **real dataset**:

- a **custom delay distribution** (a Weibull), fit to weekly **dengue** surveillance (`tbl.now::denguedat`); and
- a **custom epidemic process** (a hand-written SIR model), fit to the 2022 **mpox** outbreak (`tbl.now::mpoxdat`).

In both cases the only code that really matters is the short function *you* write; everything else (delay censoring, the count likelihood, strata, drawing the nowcast) is shared machinery you get for free.

## The one rule, and why `library(RTMB)` is needed

To fit, the package needs the gradient of the likelihood with respect to your parameters. It gets this **automatically** (automatic differentiation, via [RTMB](https://github.com/kaskr/RTMB)) — you never differentiate anything by hand. The price is a single rule:

> **Write your function in plain arithmetic that RTMB can differentiate.** Use `+`, `-`, `*`, `/`, `exp`, `log`, `sqrt`, `abs`, `sum`, `cumsum`, `pnorm`, `pgamma`, and fixed-length `for` loops. Avoid `if`/`ifelse` on a *parameter value*, `pmax`/`pmin` on parameters (use `(x + abs(x))/2` for `pmax(x, 0)`), random draws, and external solvers.

Each extension point ships a **validator** (`validate_custom_delay()`, `validate_custom_epidemic()`) — run it first; it tapes your function and reports a clear message instead of a cryptic failure inside the optimiser.

```{r load}
library(diseasenowcasting)
library(tbl.now)
library(RTMB)        # REQUIRED for custom components (see note below)
library(ggplot2)
```

> **Why `library(RTMB)`?** Built-in models work with just `library(diseasenowcasting)`. A *custom* function you write lives in your own workspace, and its arithmetic only connects to RTMB's machinery when RTMB is attached. The package keeps RTMB out of the way by default, so **attach it yourself whenever you use `custom_delay()` or `custom_epidemic()`.** If you forget, the validator reminds you.

---

# Part 1 — A custom delay: Weibull, on dengue

## The data

`denguedat` is a weekly dengue line-list from Colombia (`onset_week`, `report_week`). We take one season and build a `tbl_now`, which carries the two time indices the nowcast needs:

```{r dengue-data}
dengue_season <- subset(denguedat,
                        onset_week >= as.Date("1990-06-01") &
                        onset_week <= as.Date("1991-06-01"))
dengue_tn <- tbl_now(dengue_season,
                     event_date  = onset_week,
                     report_date = report_week,
                     data_type   = "linelist",
                     verbose     = FALSE)
```

## The custom part — a Weibull CDF factory

**This is the only code that matters here.** A custom delay is a *factory*: a function of a parameter vector `theta` returning three functions of a delay `d`.

| Element | Meaning | Weibull form |
|---|---|---|
| `cdf(d)` | $F(d) = \Pr(\text{delay} \le d)$ | $1 - \exp(-(d/\lambda)^k)$ |
| `log_cdf(d)` | $\log F(d)$ | $\log\!\big(1 - \exp(-(d/\lambda)^k)\big)$ |
| `log_survival(d)` | $\log(1 - F(d))$ | $-(d/\lambda)^k$ |

`log_cdf` and `log_survival` are supplied separately because each stays numerically stable in a different regime (`log_survival` keeps its accuracy far out in the tail). We work on the **log scale** (`theta[1] = log k`, `theta[2] = log λ`) so the optimiser is unconstrained.

```{r weibull-factory}
weibull_cdf_factory <- function(theta) {
  shape <- exp(theta[1])   # k > 0
  scale <- exp(theta[2])   # lambda > 0
  list(
    cdf          = function(d) 1 - exp(-(d / scale)^shape),
    log_cdf      = function(d) log(1 - exp(-(d / scale)^shape) + 1e-300),
    log_survival = function(d) -(d / scale)^shape
  )
}
```

Wrap it in `custom_delay()`. The arguments:

- **`cdf_factory`** — the function above.
- **`priors`** — a list with one entry per parameter, in `theta` order. Each entry is **either** a prior (a free parameter, estimated) **or** a single number (fixed). The number of parameters is read off this list, so there is no `n_params` argument.
- **`name`**, **`param_names`**, **`inits`** — labels and (unconstrained-scale) starting values.

```{r weibull-delay}
weibull_delay <- custom_delay(
  cdf_factory = weibull_cdf_factory,
  priors      = list(normal_prior(0, 1), normal_prior(log(3), 1)),
  name        = "Weibull",
  param_names = c("log_shape", "log_scale"),
  inits       = c(0, log(3))
)
validate_custom_delay(weibull_delay)
```

## Fit it (and a built-in to compare against)

The Generalized Gamma reduces **exactly** to a Weibull when its shape parameter `Q = 1`, so `generalized_gamma_delay(Q = 1)` is a perfect built-in cross-check for our custom factory:

```{r dengue-fit}
nc_weibull  <- nowcast(dengue_tn,
                       model(nb_likelihood(), ar1_epidemic(), weibull_delay),
                       type = "one_stage", temporal_effects = "none", n_draws = 300)

nc_gengamma <- nowcast(dengue_tn,
                       model(nb_likelihood(), ar1_epidemic(), generalized_gamma_delay(Q = 1)),
                       type = "one_stage", temporal_effects = "none", n_draws = 300)
```

The fitted custom parameters live in `fit$parList$custom_delay_params`:

```{r dengue-params}
theta_w <- as.numeric(nc_weibull@fits[[1]]$parList$custom_delay_params)
cat("Custom Weibull delay:  shape =", round(exp(theta_w[1]), 2),
    " scale =", round(exp(theta_w[2]), 2), "weeks\n")
# GenGamma(Q=1) is a Weibull with shape = 1/sigma, scale = exp(mu):
mu_g  <- nc_gengamma@fits[[1]]$reconstruct$delay_mu
sig_g <- nc_gengamma@fits[[1]]$reconstruct$delay_sigma
cat("Built-in GenGamma(Q=1): shape =", round(1 / sig_g, 2),
    " scale =", round(exp(mu_g), 2), "weeks\n")
```

The two should nearly coincide — confirming the custom factory matches a known distribution:

```{r dengue-compare-plot, fig.cap="Fitted reporting-delay distribution: custom Weibull vs. the built-in GenGamma(Q=1), which is algebraically the same Weibull.", fold=TRUE}
delay_grid <- seq(0.5, 20, by = 0.5)
pmf <- function(shape, scale)
  stats::pweibull(delay_grid + 0.5, shape, scale) -
  stats::pweibull(pmax(0, delay_grid - 0.5), shape, scale)

compare_df <- rbind(
  data.frame(delay = delay_grid, p = pmf(exp(theta_w[1]), exp(theta_w[2])), model = "Custom Weibull"),
  data.frame(delay = delay_grid, p = pmf(1 / sig_g, exp(mu_g)),            model = "GenGamma (Q=1)")
)
ggplot(compare_df, aes(delay, p, colour = model, linetype = model)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c("Custom Weibull" = "#2166AC", "GenGamma (Q=1)" = "#D6604D")) +
  labs(x = "Reporting delay (weeks)", y = "Probability mass",
       title = "Custom Weibull vs. built-in GenGamma(Q = 1)", colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) + theme(legend.position = "bottom")
```

And the nowcast itself, using the custom delay:

```{r dengue-nowcast-plot, fig.cap="Dengue nowcast with the custom Weibull delay (median + 50/90% intervals).", fold=TRUE}
autoplot(predict(nc_weibull, seed = 1))
```

---

# Part 2 — A custom epidemic: an SIR model, on mpox

The epidemic side is just as pluggable. A custom epidemic is **one function** `intensity_fn(theta)` returning the log expected-incidence matrix `log_mean[n_time × n_strata]` — the *entire* curve, baseline included. How you generate it (regression, random walk, ODE, …) is up to you.

## The data

`mpoxdat` is daily count data from the 2022 US mpox outbreak. We pool the race strata into one series and build a `tbl_now`:

```{r mpox-data}
mpox_tn <- tbl_now(mpoxdat,
                   event_date  = dx_date,
                   report_date = dx_report_date,
                   case_count  = n,
                   data_type   = "count-incidence",
                   verbose     = FALSE,
                   now         = as.Date("2022/09/01"))
```

The SIR loop runs over every event-time, so we need `max_time` (the number of event-times the model spans) **before** writing the function. `infer_max_time()` reads it off the data:

```{r mpox-ntime}
max_time <- infer_max_time(mpox_tn)
```

## The custom part — a hand-written SIR

**This is the code that matters.** A discrete-time SIR: susceptibles `S` and infectious `I` evolve day by day, and the modelled incidence is the new infections `β·S·I/N`. The parameters `(R0, γ, I0)` are on the log scale. Because we fill `incidence` inside a `for` loop, we add the one-line `[<-` overload at the top so RTMB can differentiate the assignment.

```{r sir-fn}
sir_intensity_fn <- function(theta) {
  #Assigning in a for loop a vector requires this:
  `[<-`     <- RTMB::ADoverload("[<-")         
  
  #We define R0, gamma and I0 as the parameters we are adjusting:
  R0        <- exp(theta[1]); gamma <- exp(theta[2]); I0 <- exp(theta[3])
  
  #The total size of the epidemic is constant:
  N         <- 1e5
  
  #And we define the initial values for S and I
  S         <- N - I0; I <- I0
  
  #We define beta (for force of infection)
  beta      <- R0 * gamma / N
  
  #Empty vector. Has to be instantiated in RTMB
  incidence <- numeric(max_time)

  #Only loops that don't depend on parameters are allowed.
  #Here we loop through a constant (time) which is valid
  for (t in seq_len(max_time)) {

    new_inf      <-  beta * S * I
    incidence[t] <- new_inf

    #Rest of the model
    S <- S - new_inf
    I <- I + new_inf - gamma * I
  }
  # log of pmax(incidence, 0), AD-safe, guarding log(0):
  matrix(log((incidence + abs(incidence)) * 0.5 + 1e-8), max_time, 1L)
}
```

Wrap it in `custom_epidemic()` (same argument shape as `custom_delay()`; the three parameters are read off the `priors` list, so there is no `n_params` argument):

```{r sir-epi}
custom_sir_epidemic <- custom_epidemic(
  intensity_fn = sir_intensity_fn,
  priors       = list(normal_prior(log(1.5), 0.4),   # log R0
                      normal_prior(log(0.1), 0.3),   # log gamma
                      normal_prior(log(5),   1.0)),  # log I0
  name         = "SIR",
  param_names  = c("log_R0", "log_gamma", "log_I0"),
  inits        = c(log(1.5), log(0.1), log(5))
)
validate_custom_epidemic(custom_sir_epidemic)
```

## Fit it

```{r mpox-fit}
nc_sir_custom  <- nowcast(mpox_tn,
                          model(nb_likelihood(), custom_sir_epidemic, lognormal_delay()),
                          temporal_effects = "none", n_draws = 300)
```

And observe the results:

```{r mpox-nowcast-plot, fig.cap="mpox nowcast with the custom SIR epidemic (median + 50/90% intervals).", fold=TRUE}
autoplot(nc_sir_custom)
```


---

# Part 3 (advanced) — An SIR with a proper ODE

> **Advanced.** Part 2 stepped the SIR forward **one Euler step per day** — simple, but a coarse approximation of the underlying differential equations. Here we integrate the *continuous* SIR ODE properly, with a **4th-order Runge–Kutta (RK4)** scheme and several sub-steps per day. The whole integrator is ordinary arithmetic, so it tapes and differentiates exactly like everything else — there is no special "ODE mode", you just write the solver.

The SIR system is

$$\frac{dS}{dt} = -\beta\,S\,I, \qquad \frac{dI}{dt} = \beta\,S\,I - \gamma\,I, \qquad \beta = \frac{R_0\,\gamma}{N},$$

and the daily incidence is the number of new infections on that day — i.e. the drop in `S`. We advance the state with `substeps` RK4 steps of size `dt = 1/substeps` per day and accumulate the new infections. Defining the right-hand sides `rhs_S` / `rhs_I` as small helper functions keeps the RK4 stage readable; RTMB traces straight through them.

```{r sir-ode-fn}
sir_ode_fn <- function(theta) {
  `[<-` <- RTMB::ADoverload("[<-")            # needed for `incidence[t] <- ...`
  R0 <- exp(theta[1]); gamma <- exp(theta[2]); I0 <- exp(theta[3])
  N  <- 1e5; beta <- R0 * gamma / N

  rhs_S <- function(S, I) -beta * S * I            # dS/dt
  rhs_I <- function(S, I)  beta * S * I - gamma * I # dI/dt

  substeps <- 4L; dt <- 1 / substeps               # 4 RK4 steps per day
  S <- N - I0; I <- I0
  incidence <- numeric(max_time)
  for (t in seq_len(max_time)) {
    new_today <- 0
    for (k in seq_len(substeps)) {                 # RK4 integration within the day
      k1S <- rhs_S(S, I);                       k1I <- rhs_I(S, I)
      k2S <- rhs_S(S + dt/2*k1S, I + dt/2*k1I); k2I <- rhs_I(S + dt/2*k1S, I + dt/2*k1I)
      k3S <- rhs_S(S + dt/2*k2S, I + dt/2*k2I); k3I <- rhs_I(S + dt/2*k2S, I + dt/2*k2I)
      k4S <- rhs_S(S + dt*k3S, I + dt*k3I);     k4I <- rhs_I(S + dt*k3S, I + dt*k3I)
      S_next <- S + dt/6 * (k1S + 2*k2S + 2*k3S + k4S)
      I_next <- I + dt/6 * (k1I + 2*k2I + 2*k3I + k4I)
      new_today <- new_today + (S - S_next)        # new infections this sub-step
      S <- S_next; I <- I_next
    }
    incidence[t] <- new_today
  }
  matrix(log((incidence + abs(incidence)) * 0.5 + 1e-8), max_time, 1L)
}
```

The wrapping and fitting are exactly as before:

```{r sir-ode-fit}
sir_ode_epidemic <- custom_epidemic(
  intensity_fn = sir_ode_fn,
  priors       = list(normal_prior(log(1.5), 0.4),
                      normal_prior(log(0.1), 0.3),
                      normal_prior(log(5),   1.0)),
  name         = "SIR-ODE",
  param_names  = c("log_R0", "log_gamma", "log_I0"),
  inits        = c(log(1.5), log(0.1), log(5))
)
validate_custom_epidemic(sir_ode_epidemic)

nc_sir_ode <- nowcast(mpox_tn,
                      model(nb_likelihood(), sir_ode_epidemic, lognormal_delay()),
                      temporal_effects = "none", n_draws = 300)
```

The RK4 ODE and the one-step Euler version give very similar fitted curves here (the mpox epidemic is slow enough that a daily step is already accurate), but the RK4 integrator is the one you would trust for fast or stiff dynamics:

```{r sir-ode-compare-plot, fig.cap="Fitted mpox incidence: the one-step Euler SIR (Part 2) vs. the RK4-integrated ODE (Part 3).", fold=TRUE}
curve_df <- rbind(
  data.frame(t = seq_len(max_time), y = exp(nc_sir_custom@fits[[1]]$reconstruct$mu_safe[, 1]),
             model = "Euler (1 step/day)"),
  data.frame(t = seq_len(max_time), y = exp(nc_sir_ode@fits[[1]]$reconstruct$mu_safe[, 1]),
             model = "RK4 ODE (4 steps/day)")
)
ggplot(curve_df, aes(t, y, colour = model, linetype = model)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Euler (1 step/day)" = "#D6604D", "RK4 ODE (4 steps/day)" = "#2166AC")) +
  labs(x = "Event day", y = "Incidence",
       title = "One-step Euler vs. RK4-integrated SIR ODE", colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) + theme(legend.position = "bottom")
```

---

## Tips for writing your own component

1. **Parameterise on the unconstrained scale.** Use `exp(theta[i])` for positive quantities, `plogis(theta[i])` for probabilities; give `inits` on that same scale.
2. **A custom epidemic owns the whole `log_mean`.** Include any baseline level — nothing (no intercept, no covariates) is added on top.
3. **Know `max_time` first** if your function loops over time. `infer_max_time(tn)` gives the number of event-times the model spans.
4. **Accumulate without `[<-` when you can.** `cumsum(sigma * eps)` builds a random walk with no index assignment (so no overload needed). Reach for `` `[<-` <- RTMB::ADoverload("[<-") `` only when a loop must fill a pre-allocated vector (e.g. an ODE).
5. **Guard `log` of near-zero values** with `log((x + abs(x)) * 0.5 + 1e-8)` — the `(x + abs(x))/2` is an AD-safe `pmax(x, 0)`.
6. **Validate before fitting** — `validate_custom_delay()` / `validate_custom_epidemic()` catch non-differentiable code early.
7. **For production ODEs**, the `RTMBode` package offers an adaptive/stiff solver that plugs into RTMB; the hand-rolled Euler loop above keeps this vignette dependency-free.
