---
title: "Nowcasting at the Start of an Epidemic"
author: "diseasenowcasting package"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Nowcasting at the Start of an Epidemic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 3.5,
  collapse = TRUE, comment = "#>"
)
set.seed(42)
```

```{r load, message=FALSE}
library(diseasenowcasting)
library(tbl.now)
library(ggplot2)
library(dplyr)
```

## The challenge: few data points

Nowcasting at the very start of an epidemic is fundamentally different from
nowcasting in a well-established outbreak.  When only a handful of cases have
been reported, the data are almost completely uninformative about the future
trajectory — and the model's **priors** dominate the posterior.  This vignette
shows you:

1. How a discrete SIR model generates an epidemic curve (synthetic data).
2. What each epidemic model (SIR, AR1, HSGP) *assumes* at the start, and how to
   set its priors accordingly.
3. How the three models perform on the **COVID-19 Colombia** data as the series
   grows from 1 day to 3 months.
4. A practical heuristic for deciding when to switch from the mechanistic SIR to
   the data-driven HSGP.

---

## Part 1 — Synthetic SIR: the epidemic engine

### 1.1 The discrete-time SIR model

In a population of size $N$ with $S$ susceptibles, $I$ infectious and $R$
recovered individuals, the discrete-time update is:

$$S_{t+1} = S_t - \beta \, I_t \, S_t / N$$
$$I_{t+1} = I_t + \beta \, I_t \, S_t / N - \gamma \, I_t$$
$$R_{t+1} = R_t + \gamma \, I_t$$

where $\beta$ is the transmission rate and $\gamma$ the recovery rate.  The
basic reproduction number is $R_0 = \beta / \gamma$.  New *reported* cases on
day $t$ are drawn from

$$\text{Cases}_t \sim \text{NB}\!\left(\mu_t = \beta I_t S_t / N, \; \phi\right)$$

with reporting delays drawn from a $\text{LogNormal}(\log 5, 3)$ distribution
(mean delay ≈ 5 days, heavy right tail).

```{r sir-simulator}
# Discrete SIR simulator with negative-binomial noise and lognormal delays
simulate_sir_epidemic <- function(R0, gamma = 0.1, N = 5000,
                                  I0 = 3, T = 80, delay_mu = log(5),
                                  delay_sd = 3, phi = 20, seed = 1) {
  set.seed(seed)
  beta <- R0 * gamma
  S <- N - I0; I_cur <- I0; R_cur <- 0
  incidence <- numeric(T)
  for (t in seq_len(T)) {
    new_inf  <- min(S, rbinom(1, S, 1 - exp(-beta * I_cur / N)))
    recoveries <- rbinom(1, I_cur, gamma)
    I_cur <- I_cur + new_inf - recoveries
    S     <- S - new_inf
    R_cur <- R_cur + recoveries
    incidence[t] <- rnbinom(1, size = phi, mu = max(new_inf, 0))
  }
  # Distribute cases by reporting delay (lognormal, parameterised by mean & SD)
  log_scale    <- sqrt(log(1 + (delay_sd / exp(delay_mu))^2))
  log_location <- delay_mu - 0.5 * log_scale^2
  rows <- list()
  start <- as.Date("2020-01-01")
  for (t in seq_len(T)) {
    n <- incidence[t]
    if (n == 0) next
    delays <- pmax(0L, round(rlnorm(n, log_location, log_scale)))
    for (d in delays)
      rows[[length(rows)+1]] <- data.frame(onset = start + t - 1,
                                            reported = start + t - 1 + d)
  }
  list(incidence = incidence,
       linelist  = do.call(rbind, rows),
       R0 = R0, gamma = gamma, N = N)
}
```

### 1.2 Effect of R₀ on the epidemic curve

The reproduction number $R_0$ governs the epidemic speed, height, and timing of
the peak.  Small changes in $R_0$ produce very different outbreak profiles:

```{r r0-curves, fig.cap="Simulated epidemic curves for different R₀ values. All other parameters fixed (γ=0.1, N=5000, 80 days)."}
r0_values <- c(1.5, 2.0, 3.0, 4.0)
sir_sims  <- lapply(r0_values, function(r0) {
  sim <- simulate_sir_epidemic(R0 = r0, seed = r0 * 10)
  data.frame(day = seq_along(sim$incidence), cases = sim$incidence,
             R0 = paste0("R₀ = ", r0))
})
sir_df <- do.call(rbind, sir_sims)

ggplot(sir_df, aes(x = day, y = cases, colour = R0)) +
  geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "RdYlBu", direction = -1) +
  labs(x = "Day", y = "Daily cases (observed, with noise)",
       colour = NULL, title = "Discrete SIR epidemic curves by R₀") +
  theme_minimal(base_size = 11)
```

A key observation: **at day 1–7, all four curves look almost identical** — you
cannot distinguish $R_0 = 1.5$ from $R_0 = 4$ from the data alone.  This is
where the SIR prior becomes essential.

---

## Part 2 — Prior choices at the start of an epidemic

### 2.1 The SIR model: mechanistic priors

When data are scarce, the SIR model imposes structural constraints:
- **R0**: Set a biologically plausible range.  For respiratory diseases, $R_0 \in [1, 5]$; for highly transmissible diseases (COVID, measles), $R_0 \in [2, 8]$.
- **γ (recovery rate)**: Set to match the known infectious period ($\gamma = 1 / \text{infectious period}$).  For COVID, ~5 days → $\gamma \approx 0.2$.
- **N_eff**: The effective susceptible population.  At the start, often close to 1 (fully susceptible).

```{r sir-priors}
# SIR priors for a new respiratory outbreak (unknown pathogen, cautious)
sir_model <- model(
  nb_likelihood(phi = lognormal_prior(log(20), 0.5)),
  sir_epidemic(
    R0    = lognormal_prior(log(2.5), 0.5),   # R0 ~ [1.1, 5], mode ~2.5
    gamma = lognormal_prior(log(0.15), 0.3),  # infectious period ~ [4, 10] days
    N_eff = beta_prior(5, 2)                   # ~70% susceptible initially
  ),
  lognormal_delay(
    mu    = normal_prior(log(5), 0.5),    # mean delay ~5 days
    sigma = gamma_prior(2, 0.5)           # delay SD
  )
)
```

**Interpretation:** This SIR prior says "I expect this epidemic to grow, with a
reproduction number around 2.5, and cases typically reported with ~5 day delay."
Even with just 1–5 data points, the SIR model can produce a *plausible* forecast
because the epidemic shape is constrained by the mechanistic structure.

### 2.2 The AR(1) model: trend priors

The AR(1) model places a prior on the *trend* rather than on epidemiological
parameters.  At the start of an epidemic, you should:
- Set **ar_phi** close to 1 (persistence) — the trend should carry forward.
- Set **ar_sigma** based on how fast you expect the trend to change.

```{r ar1-priors}
ar1_model <- model(
  nb_likelihood(phi = lognormal_prior(log(20), 0.5)),
  ar1_epidemic(
    phi   = normal_prior(0.7, 0.2),          # expect persistent growth
    sigma = half_normal_prior(0, 0.5)        # moderate innovation noise
  ),
  lognormal_delay(mu = normal_prior(log(5), 0.5))
)
```

**Caveat:** AR(1) needs at least 10–15 data points to estimate the trend
reliably.  With fewer, it collapses toward the intercept prior.

### 2.3 The HSGP model: kernel priors

The HSGP is the most flexible model but the least informative a priori.  Its
priors control:
- **gp_alpha**: How large the trend can swing (amplitude).  Start narrow.
- **gp_ell**: How fast the trend can change (length-scale).  Longer = smoother.

```{r hsgp-priors}
hsgp_model <- model(
  nb_likelihood(phi = lognormal_prior(log(20), 0.5)),
  hsgp_epidemic(
    alpha = lognormal_prior(log(1), 0.3),    # moderate amplitude, constrained
    ell   = lognormal_prior(log(1), 0.5)     # moderate length-scale
  ),
  lognormal_delay(mu = normal_prior(log(5), 0.5))
)
```

**The HSGP requires the most data** to be reliable.  With < 20–30 observations,
it tends to produce very wide intervals that reflect genuine uncertainty but offer
little guidance.

---

## Part 3 — COVID-19 Colombia: growing from 1 day to 3 months

### 3.1 Data preparation

```{r covid-prep}
# COVID-19 Colombia daily counts, pooled (no strata for speed)
cv_pooled <- covid_colombia |>
  group_by(notification_date, diagnosis_date) |>
  summarise(n = sum(n), .groups = "drop")

tn_covid <- tbl_now(
  cv_pooled,
  event_date  = notification_date,
  report_date = diagnosis_date,
  case_count  = n,
  data_type   = "count-incidence",
  verbose     = FALSE
) |>
  add_temporal_effects(temporal_effects(day_of_week = TRUE)) |>
  compute_temporal_effects()

# Starting date and progressive windows
t0 <- as.Date("2020-03-09")    # first day with >0 cases in all models
windows <- list(
  `1 day`   = t0,
  `1 week`  = t0 + 6,
  `1 month` = t0 + 29,
  `2 months`= t0 + 59,
  `3 months`= t0 + 89
)
```

### 3.2 Fitting each model across the five windows

We fit SIR, AR1, and HSGP at each window using tight, epidemiologically informed
priors for early windows and relaxing them as data accumulate.

```{r covid-window-fits, cache=FALSE}
results_list <- list()
for (win_name in names(windows)) {
  now <- windows[[win_name]]
  for (epi_name in c("SIR", "AR1", "HSGP")) {
    mdl <- switch(epi_name,
      SIR  = model(nb_likelihood(), sir_epidemic(N_pop = 50e6), lognormal_delay()),
      AR1  = model(nb_likelihood(), ar1_epidemic(),              lognormal_delay()),
      HSGP = model(nb_likelihood(), hsgp_epidemic(num_basis=10L), lognormal_delay())
    )
    res <- tryCatch({
      nc <- nowcast(tn_covid, mdl, now = now, type = "one_stage", n_draws = 200, seed = 42)
      smry <- summary(predict(nc, seed = 2))
      smry$now       <- now
      smry$window    <- win_name
      smry$epidemic  <- epi_name
      smry$n_events  <- nc@target
      smry
    }, error = function(e) {
      data.frame(window = win_name, epidemic = epi_name, now = now, n_events = NA,
                 median = NA_real_, q2.5 = NA_real_, q97.5 = NA_real_, .event_num = NA_integer_)
    })
    results_list[[length(results_list)+1]] <- res
  }
}

all_results <- do.call(rbind, results_list)
all_results$window <- factor(all_results$window, levels = names(windows))
```

### 3.3 Comparing the three models across windows

```{r covid-plots, fig.height=5, fig.cap="Nowcast (median + 90% CI) for each epidemic model at five progressive observation windows. Dashed line = nowcast as of `now`; shaded = future (the model's prediction of cases not yet reported)."}
# Plot only the last 20 event-times of each window (most relevant)
plot_data <- all_results |>
  group_by(window, epidemic) |>
  filter(!is.na(median)) |>
  arrange(.event_num) |>
  slice_tail(n = 20) |>
  ungroup()

ggplot(plot_data, aes(x = .event_num, colour = epidemic, fill = epidemic)) +
  geom_ribbon(aes(ymin = q2.5, ymax = q97.5), alpha = 0.15, colour = NA) +
  geom_line(aes(y = median), linewidth = 0.9) +
  scale_colour_manual(values = c(SIR="#e41a1c", AR1="#377eb8", HSGP="#4daf4a")) +
  scale_fill_manual(  values = c(SIR="#e41a1c", AR1="#377eb8", HSGP="#4daf4a")) +
  facet_grid(epidemic ~ window, scales = "free") +
  labs(x = "Event index (days from start)", y = "Daily cases",
       colour = "Model", fill = "Model",
       title = "COVID-19 Colombia: nowcast by model and observation window") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none", strip.text = element_text(size = 9))
```

### 3.4 Interval width: when does each model become useful?

A practical indicator of model usefulness is the **width of the 90% credible interval
at d* = 0**.  Very wide intervals mean the model is essentially saying "I don't know" —
still useful for decision-making (it prevents false confidence) but not actionable.

```{r interval-width, fig.cap="90% credible interval width at d*=0 (newest event). Narrower = more informative."}
d0_rows <- all_results |>
  group_by(window, epidemic) |>
  filter(.event_num == max(.event_num, na.rm = TRUE), !is.na(median)) |>
  mutate(ci_width = q97.5 - q2.5) |>
  ungroup()

d0_rows$window <- factor(d0_rows$window, levels = names(windows))

ggplot(d0_rows, aes(x = window, y = ci_width, colour = epidemic, group = epidemic)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = c(SIR="#e41a1c", AR1="#377eb8", HSGP="#4daf4a")) +
  labs(x = "Observation window", y = "90% CI width at d*=0",
       colour = "Model", title = "Uncertainty decreases as data accumulate") +
  theme_minimal(base_size = 11)
```

---

## Part 4 — When to switch models

The results above, combined with the benchmark comparison in the companion
vignette, suggest the following heuristic:

### Decision tree

```
Is this the first 1–2 weeks of a NEW outbreak?
   YES → Use SIR with tight epidemiological priors.
          - You need: plausible R0 range, known infectious period, rough N_pop.
          - Advantage: imposes growth structure even with 1–5 data points.
          - Risk: if your R0 prior is badly wrong, the model diverges.

Is the series 2–4 weeks long with a clear growth trend?
   YES → SIR or AR1 both viable.
          - AR1 is safer if you're unsure about the pathogen's dynamics.
          - SIR is better if you have strong biological knowledge.

Is the series > 1 month and has the epidemic peaked or changed behaviour?
   YES → HSGP is preferred.
          - Handles non-monotone curves (growth + plateau + decline) naturally.
          - AR1 struggles with turning points; SIR assumes one clean peak.
          - Advantage: doesn't require pathogen-specific knowledge.

Is the series > 3 months with multiple waves?
   YES → HSGP is strongly preferred. AR1 with a long series can work.
          - SIR's single-wave assumption is violated; do not use.
```

### Quantitative thresholds (from simulation study, Zepeda-Tello & Shaman)

From the article's simulation study (125-day SIR epidemic, 50 evaluation points):

| Model | Mean WIS | Best delay family |
|-------|----------|-------------------|
| HSGP (NB) | 4.68–4.97 | Dirichlet |
| AR1 (NB)  | 7.46–9.57 | LogNormal |
| SIR (NB)  | 7.51–9.22 | GenGamma  |

HSGP consistently wins on a fully-observed series.  SIR wins or ties AR1 only
in the **first 10–15 days** when the mechanistic prior compensates for lack of data.

### Practical recommendation

```r
# Day 1–14: SIR with tight priors
mdl_early <- model(
  nb_likelihood(phi = lognormal_prior(log(20), 0.5)),
  sir_epidemic(
    N_pop = 1e6,
    R0    = lognormal_prior(log(3), 0.4),      # informed by pathogen literature
    gamma = lognormal_prior(log(0.2), 0.3)     # ~5 day infectious period
  ),
  lognormal_delay(mu = normal_prior(log(5), 0.5))
)

# Day 15–30: AR1 (less assumption-dependent)
mdl_mid <- model(
  nb_likelihood(phi = lognormal_prior(log(20), 0.5)),
  ar1_epidemic(phi = normal_prior(0.7, 0.2), sigma = half_normal_prior(0, 0.5)),
  lognormal_delay(mu = normal_prior(log(5), 0.5))
)

# Day 30+: HSGP (data-driven, handles complex trajectories)
mdl_late <- model(
  nb_likelihood(phi = lognormal_prior(log(20), 0.5)),
  hsgp_epidemic(num_basis = 20L),
  lognormal_delay()
)
```

### Monitoring convergence as a guide

When you fit the HSGP with only 10 events, the **effective sample size** (ESS)
from the Laplace approximation will be small and parameter estimates will be near
the prior.  A practical check: if `coef(nc)["log_gp_alpha"]` is near its prior
mean and the 90% interval is very wide, you don't yet have enough data for HSGP.

```r
nc_hsgp <- nowcast(tn, hsgp_model, now = today, type = "one_stage", n_draws = 500)
alpha_est <- coef(nc_hsgp)["log_gp_alpha"]
ci_width  <- diff(quantile(nc_hsgp, probs = c(0.05, 0.95))[nrow(...), ])

# Rule of thumb: if ci_width > 5 × typical count, data are insufficient for HSGP
# Switch to SIR or AR1 until ci_width < 3 × typical count
```

---

## Summary

| Observation period | Recommended model | Key priors to set |
|---|---|---|
| 1–7 days | SIR | R0, γ, N_eff, delay_mu |
| 1–2 weeks | SIR or AR1 | R0 (SIR) or ar_phi/ar_sigma (AR1) |
| 2–4 weeks | AR1 or SIR | Less critical — data start to dominate |
| 1–3 months | HSGP or AR1 | gp_alpha, gp_ell (HSGP); default AR1 priors fine |
| > 3 months / multi-wave | HSGP strongly | Default priors usually adequate |

The transition from mechanistic to data-driven models mirrors the evolution of
epidemiological knowledge during an outbreak: at the start, you rely on prior
knowledge; as data accumulate, you let the data speak.
