---
title: "Forecasting"
author: "diseasenowcasting team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4, fig.align = "center"
)
```

A nowcast estimates the cases that have already happened but are not yet all
reported. A forecast asks about cases that have not happened yet. For
`diseasenowcasting` the two are the same computation: after `now` nothing has
been reported, so those event times add nothing to the likelihood, and the
epidemic there is just the fitted process carried forward. `forecast()` does
exactly that with the draws of an existing fit, so there is no refit and the
nowcast and forecast form one joint distribution.

```{r libraries}
library(tbl.now)
library(diseasenowcasting)
library(dplyr)
library(ggplot2)
```

## Forecasting next week's dengue cases

We nowcast weekly dengue in Puerto Rico as of 5 September 1994 with an
exponential-smoothing epidemic.

```{r dengue-fit}
data(denguedat)
dengue <- tbl_now(
  filter(denguedat, onset_week >= as.Date("1992-01-06")),
  event_date = onset_week, report_date = report_week,
  data_type = "linelist", verbose = FALSE
)

fit <- nowcast(
  dengue,
  model(nb_likelihood(), ets_epidemic(), lognormal_delay()),
  now = as.Date("1994-09-05"), temporal_effects = "none",
  n_draws = 1000, seed = 1
)
```

`forecast()` extends it `h` weeks. The result is an ordinary `tbl_nowcast`
with an extra `.horizon` column: `0` at `now`, negative for the nowcast,
`1, ..., h` for the forecast.

```{r dengue-forecast}
fc <- forecast(fit, h = 2, seed = 2)

# Median and 95% interval, one row per date
interval <- function(predictions) {
  predictions |>
    group_by(across(-c(.quantile_level, .value))) |>
    summarise(lower  = .value[.quantile_level == 0.025],
              median = .value[.quantile_level == 0.5],
              upper  = .value[.quantile_level == 0.975], .groups = "drop")
}
interval(fc@predictions) |> filter(.horizon >= -1)
```

The uncertainty widens past `now` for two reasons: the epidemic can move in the
new week, and nothing of it has been reported yet.

```{r dengue-plot}
interval(fc@predictions) |>
  filter(.horizon >= -15) |>
  ggplot(aes(onset_week)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = .horizon > 0), alpha = 0.3) +
  geom_line(aes(y = median, colour = .horizon > 0)) +
  geom_vline(xintercept = as.Date("1994-09-05"), linetype = "dashed") +
  scale_fill_manual(values = unname(dn_palette()[c("reported", "accent")]),
                    labels = c("nowcast", "forecast"), name = NULL) +
  scale_colour_manual(values = unname(dn_palette()[c("reported", "accent")]),
                      labels = c("nowcast", "forecast"), name = NULL) +
  labs(x = "Onset week", y = "Cases") +
  theme_diseasenowcasting()
```

Every recursive epidemic process forecasts any horizon: `ar1_epidemic()`,
`arima_epidemic()`, `ets_epidemic()`, `sts_epidemic()`, the random walks and
`sir_epidemic()`. `hsgp_epidemic()` extrapolates its basis only up to the edge
of its domain, and a `custom_epidemic()` cannot be extended. Temporal effects
are recomputed for the new dates; any other event covariate needs its future
values in `new_data`.

## Which cases? Revisions

With a revision process, reports are later confirmed or retracted, and a
forecast can count any part of them. In `covid_us`, a probable case is never
laboratory-confirmed, so the revision process is confirmation-only.

```{r covid-fit}
data(covid_us)
covid <- covid_us |>
  filter(onset_dt >= as.Date("2020-03-01"), onset_dt <= as.Date("2020-07-31")) |>
  mutate(
    revision_type = if_else(current_status == "Laboratory-confirmed case",
                            "confirmed", "pending"),
    cdc_report_dt = if_else(revision_type == "pending", as.Date(NA), cdc_report_dt)
  ) |>
  count(onset_dt, pos_spec_dt, cdc_report_dt, revision_type, wt = n) |>
  tbl_now(event_date = onset_dt, report_date = pos_spec_dt,
          revision_date = cdc_report_dt, revision_type = revision_type,
          case_count = n, data_type = "count-incidence", verbose = FALSE) |>
  aggregate_time_units("weeks", verbose = FALSE)

covid_fit <- nowcast(
  covid, model(nb_likelihood(), ets_epidemic(), lognormal_delay()),
  now = as.Date("2020-07-05"), temporal_effects = "none",
  n_draws = 1000, seed = 1
)
```

`category` chooses what is counted. The default is the fit's own estimand:
here, the reports that will eventually be confirmed.

```{r covid-categories}
next_week <- function(category) {
  forecast(covid_fit, h = 1, category = category,
           include_nowcast = FALSE, seed = 3)@predictions |>
    interval() |>
    mutate(category = category)
}
bind_rows(lapply(c("confirmed", "pending", "overall"), next_week))
```

The categories are *eventual* statuses: `"overall"` counts every report,
`"confirmed"` and `"retracted"` those that end up that way, and `"pending"`
those that never do. Each mode only defines the categories it can observe;
asking for `"retracted"` here is an error, because this data never records a
retraction. See `?forecast.diseasenowcasting_nowcast` for the full table.
