---
title: "Handling Outlier Delays with Censoring"
author: "Rodrigo Zepeda-Tello"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling Outlier Delays with Censoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 4,
  collapse = TRUE, comment = "#>"
)
```

```{r load}
set.seed(248675)
library(diseasenowcasting)
library(tbl.now)
library(dplyr)
library(ggplot2)
```

## The problem: an extreme delay

Real surveillance data occasionally contains reports with extreme reporting delays. This can be due to typos, healthcare-system hurdles or other issues not related to the disease's natural evolution.  In
Colombia's COVID-19 data (`covid_colombia`) the bulk of reports arrive within a
week or two, but a handful take **more than 100 days**:

```{r delay-hist1}
data(covid_colombia)

tbl_covid <- covid_colombia |> 
  tbl_now(event_date = notification_date,
          case_count  = n,
              data_type   = "count-incidence",
          report_date = diagnosis_date,
          t_effects = temporal_effects(day_of_week = TRUE))

summary(tbl_covid$.delay)
```

```{r delay-hist2, echo=FALSE, fig.cap="_Reporting-delay distribution of COVID-19 Colombia (extremes exagerated for illustration purposes). A few reports arrive hundreds of days late._"}
tbl_covid |> 
  filter(n > 0) |> 
  ggplot() +
  geom_histogram(aes(x = .delay, weight = n), fill = "#A8BFA9", binwidth = 5) +
  geom_histogram(aes(x = .delay, weight = n), fill = "#DA6529", binwidth = 5,
                 data = tbl_covid |> filter(.delay > 102)) +
  annotate(
    "segment",
    x = 220, xend = 110,
    y = 1.e6, yend = 1000,
    arrow = arrow(length = unit(0.25, "cm"))
  ) +
  annotate(
    "text",
    x = 220,
    y = 1.e6,
    label = "Extreme delays of\n100 days or more",
    hjust = 0
  ) +
  theme_diseasenowcasting() +
  scale_y_sqrt(labels = scales::comma_format()) +
  labs(x = "Delay (days)", y = "Case count", title = "COVID-19 cases in Colombia")
```



When a parametric delay model (log-normal, gamma, ...) is fit to data containing
such an outlier, the extreme value **drags the estimated delay distribution to
the right**.  The model then believes delays are longer than they really are,
thus inflating the most recent nowcasts.

The `diseasenowcasting` framework offers a fix. It treats such reports as
**right-censored**.  Instead of telling the model "this case had delay exactly
330", it tells it only "this case arrived *by* delay 330" (_i.e._ its delay of 330 is an
**upper bound** for the true delay). 

In what follows we explain how to use the model to automatically detect abnormal
delays and how to inform the model so that predictions are improved. 

## 1) Fit a model

The first step for a model to learn about *extreme delays* is to have an 
initial model with historical data so that it learns what the usual 
delay distribution is. In this case we'll work  with an early-pandemic window and fit a nowcast. To 
play out the "new data arrives" story we first fit on the reports available
at an early date:

```{r filter-asof}
#Initial data
initial_tbl <- tbl_covid |> 
  filter(notification_date <= as.Date("2020-08-31") & diagnosis_date <= as.Date("2020-08-31")) |> 
  change_now() #Update the "now" of the nowcast to the latest date
```


Fit a nowcast to this data

```{r fit-asof}
initial_ncast <- nowcast(initial_tbl)
```

## 2) Update the model

We can then get new data:

```{r filert-asof2}
new_data_tbl <- tbl_covid |> 
  filter(notification_date <= as.Date("2020-09-01") & diagnosis_date <= as.Date("2020-09-01")) |> 
  change_now()
```

And,  `update()` the model. This will automatically score the new report against 
the old fit and **warn** that something is amiss:

```{r surprise, warning=TRUE}
nc_updated <- update(initial_ncast, new_data_tbl)
```

The warning tells us *exactly* what was unexpected (reporting delays far longer
than usual).  The full table is available via `extreme_values()`:

```{r surprise-table}
extreme_values(nc_updated)
```

The `mean_tail_prob` expressess the probability of observing such a value, The `cdf_prob` the probability of lying below that value. Variable `level` shows the level of certainty to qualify something as an outlier (default = `0.99`) and can be modified in `update(..., level = 0.95)`. 

## 3) Censor the outliers and re-fit

We follow the warning's advice: we flag as censored every report whose delay exceeds a sensible
bound (here 99 days as reported by `extreme_values()`). The function `censor_delays_above()` works  by setting `is_censored = TRUE` in the `tbl_now` for
reports greater than the `max_delay`. Extreme delays are thus turned into upper bounds. The `nowcast()` then reads
the `is_censored` flag automatically.

```{r censor, eval = FALSE}
new_data_tbl_censored <- censor_delays_above(new_data_tbl, max_delay = 99)

#Adds column `.is_censored`:
new_data_tbl_censored
```

```{r, echo = FALSE}
new_data_tbl_censored <- censor_delays_above(new_data_tbl, max_delay = 99)

new_data_tbl_censored |> dplyr::select(.is_censored, dplyr::everything())
```


We refit:

```{r refit}
nc_updated_censored <- update(initial_ncast, new_data_tbl_censored)
```

The fitted values change once the outliers are no longer taken literally.

```{r delay-compare}
#Previous
coef(nc_updated)

#Updated
coef(nc_updated_censored)
```

Which also affects predictions:

```{r delay-compare-predict}
#Previous
pred_previous <- predict(nc_updated) 
summary(pred_previous) |> tail(6)

#Updated
pred_censored <- predict(nc_updated_censored)
summary(pred_censored) |> tail(6)
```



## 4) Does it nowcast better?  Backtest

Finally we check that controlling the extreme values actually *improves accuracy*.  We
backtest the same model on the plain data (`new_data_tbl`) and on the censored 
data (`new_data_tbl_censored`) across a set of dates, scoring the most recent nowcast ($d^* = 0$) against the eventual truth with the Weighted Interval Score (WIS; lower is better) and coverage (closer to the expected coverage the better).

```{r backtest-compare}
eval_dates <- as.Date(c("2020-04-15", "2020-05-01", "2020-05-15", "2020-06-01"))

bt_plain <- backtest(new_data_tbl, dates = eval_dates)
bt_cens  <- backtest(new_data_tbl_censored, dates = eval_dates)

rbind(
  plain    = score(bt_plain, report = FALSE)[, c("wis", "coverage_50", "coverage_90")],
  censored = score(bt_cens,  report = FALSE)[, c("wis", "coverage_50", "coverage_90")]
)
```

Censoring the outlier delays result in a **lower (better) WIS** and a better coverage:
the delay distribution is no longer pulled to the right by extreme values. **This
shows that censoring the extreme values improves the models.**

## Summary -- fit -> update -> identify outliers -> censor -> refit loop

In general the workflow is: 

1. Fit a nowcast.

2. New data arrives -> `update()` scores it and **warns** about potential 
outliers.

3. Manually identify which ones correspond to outliers and which correspond
to true values. This has to be done by a human as no automated system will know
when something flagged as noise is real. 

4. Use `censor_delays_above(tn, bound)` turning the delays into an upper bound. Or
modify the `tbl_now` directly (column `is_censored`)

5. Re-fit -> the delay distribution is no longer distorted.

