---
title: "Handling Outlier Delays with Censoring"
author: "Rodrigo Zepeda-Tello"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling Outlier Delays with Censoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 4,
  collapse = TRUE, comment = "#>"
)
```

```{r load}
set.seed(248675)
library(diseasenowcasting)
library(tbl.now)
library(dplyr)
library(ggplot2)
```

## The problem: an extreme delay

Real surveillance data occasionally contains reports with extreme reporting delays. This can be due to typos, healthcare-system hurdles or other issues not related to the disease's natural evolution.  In
Colombia's COVID-19 data (`covid_colombia`) the bulk of reports arrive within a
week or two, but a handful take **more than 300 days**:

```{r delay-hist1}
data(covid_colombia)

tbl_covid <- covid_colombia |> 
  tbl_now(event_date = notification_date,
          case_count  = n,
              data_type   = "count-incidence",
          report_date = diagnosis_date,
          t_effects = temporal_effects(day_of_week = TRUE))

summary(tbl_covid$.delay)
```

```{r delay-hist2, echo=FALSE, fig.cap="Reporting-delay distribution of COVID-19 Colombia (extremes exagerated for illustration purposes). A few reports arrive hundreds of days late}
tbl_covid |> 
  filter(n > 0) |> 
  ggplot() +
  geom_histogram(aes(x = .delay, weight = n), fill = "#A8BFA9", binwidth = 5) +
  geom_histogram(aes(x = .delay, weight = n), fill = "#DA6529", binwidth = 5,
                 data = tbl_covid |> filter(.delay > 102)) +
  annotate(
    "segment",
    x = 220, xend = 110,
    y = 1.e6, yend = 1000,
    arrow = arrow(length = unit(0.25, "cm"))
  ) +
  annotate(
    "text",
    x = 220,
    y = 1.e6,
    label = "Extreme delays of\n100 days or more",
    hjust = 0
  ) +
  theme_diseasenowcasting() +
  scale_y_sqrt(labels = scales::comma_format()) +
  labs(x = "Delay", y = "Case count", title = "COVID-19 cases in Colombia")
```



When a parametric delay model (log-normal, gamma, ...) is fit to data containing
such an outlier, the extreme value **drags the estimated delay distribution to
the right**.  The model then believes delays are longer than they really are,
thus inflating the most recent nowcasts.

The `diseasenowcasting` framrwork offers a fix. It treats such reports as
**right-censored**.  Instead of telling the model "this case had delay exactly
330", it tells it only "this case arrived *by* delay 330" (_i.e._ its delay is an
**upper bound**). 

In what follows we explain how to use the model to automatically detect abnormal
delays and how to inform the model so that predictions are improved. 

## 1) Fit a model

The first step for a model to learn about *extreme delays* is to have an 
initial model with historical data so that it learns what the usual 
delay distribution is. 

In this case we'll work  with an early-pandemic window and fit a nowcast. To 
play out the "new data arrives" story we first fit on the reports available
**as of** an early date by specifying a **now**:

```{r fit-asof}
initial_tbl <- tbl_covid |> 
  filter(notification_date <= as.Date("2020-08-31") & diagnosis_date <= as.Date("2020-08-31"))
```


Fit a nowcast

```{r fit-asof}
initial_ncast <- nowcast(initial_tbl)
```


And then get new data:

```{r fit-asof}
new_data_tbl <- tbl_covid |> 
  filter(notification_date <= as.Date("2020-09-01") & diagnosis_date <= as.Date("2020-09-01"))
```

Finally,  `update()` will automatically score the new report against 
the old fit and **warn** that something is amiss:

```{r surprise, warning=TRUE}
nc_updated <- update(initial_ncast, new_data_tbl)
```

The warning tells us *exactly* what was unexpected (a reporting delay far longer
than the model expects) and suggests the remedy.  The full table is available
via `surprise_result()`:

```{r surprise-table}
sr <- surprise_result(nc_updated)
subset(sr$delay_surprise, is_surprising)
```

## Step 2 -- censor the outlier and re-fit

We follow the warning's advice: flag every report whose delay exceeds a sensible
bound (here 45 days, comfortably above the 99th percentile of ~50 days) as
censored.  `censor_delays_above()` does this -- it sets `is_censored = TRUE` for
those reports, turning their delay into an upper bound.  `nowcast()` then reads
the `is_censored` flag automatically.

```{r censor}
tn_censored <- censor_delays_above(tn, max_delay = 45)
```

```{r refit}
nc_plain    <- nowcast(tn,          mdl, now = as.Date("2020-05-01"),
                       type = "one_stage", n_draws = 500,
                       temporal_effects = "none", seed = 1)
nc_censored <- nowcast(tn_censored, mdl, now = as.Date("2020-05-01"),
                       type = "one_stage", n_draws = 500,
                       temporal_effects = "none", seed = 1)
```

The fitted delay mean drops once the outliers are no longer taken literally:

```{r delay-compare}
data.frame(
  fit       = c("plain", "censored"),
  delay_mu  = round(c(coef(nc_plain)["delay_mu"], coef(nc_censored)["delay_mu"]), 3),
  delay_sd  = round(c(coef(nc_plain)["delay_sigma"], coef(nc_censored)["delay_sigma"]), 2)
)
```

## Step 3 -- does it nowcast better?  A backtest

Finally we check that controlling the surprise actually *improves accuracy*.  We
backtest the same model on the plain data and on the censored data across a set
of as-of dates, scoring the d\* = 0 nowcast against the eventual truth with the
Weighted Interval Score (WIS; lower is better).

```{r backtest-compare}
eval_dates <- as.Date(c("2020-04-15", "2020-05-01", "2020-05-15", "2020-06-01"))

bt_plain <- backtest(tn,          mdl, dates = eval_dates, type = "one_stage",
                     n_draws = 300, max_delay = Inf, temporal_effects = "none", seed = 1)
bt_cens  <- backtest(tn_censored, mdl, dates = eval_dates, type = "one_stage",
                     n_draws = 300, max_delay = Inf, temporal_effects = "none", seed = 1)

rbind(
  plain    = score(bt_plain, report = FALSE)[, c("wis", "coverage_90")],
  censored = score(bt_cens,  report = FALSE)[, c("wis", "coverage_90")]
)
```

Censoring the handful of outlier delays gives a **lower (better) WIS** without
hurting coverage: the delay distribution is no longer pulled to the right by a
few extreme values, so the most recent nowcasts are better calibrated.

## Summary -- the surprise -> censor -> refit loop

```
1. Fit a nowcast.
2. New data arrives -> update() scores it and WARNS about surprises.
3. A "delay too long" warning flags an outlier report.
4. censor_delays_above(tn, bound) turns its delay into an upper bound.
5. Re-fit -> the delay distribution is no longer distorted -> better WIS.
```

This is the practical payoff of modelling the reporting delay as a stochastic
process with proper censoring rather than a nuisance to be cleaned away: the
model can *tell you* when an observation does not fit, and the censoring
machinery lets you respond without discarding data.

> **Note.** The improvement is largest for the **one-stage** fit, whose joint
> likelihood is directly distorted by outlier delays.  The **two-stage** cascade
> estimates the delay from a recent window first and is already fairly robust to
> such outliers, so it benefits less.

## References

- Zepeda-Tello, R. & Shaman, J. (2025). Nowcasting from an epidemic's
  reporting-delay: a censored regression approach. *Preprint*.
- Gunther, F. et al. (2021). Nowcasting the COVID-19 pandemic in Bavaria.
  *Biometrical Journal*.
