---
title: "Introduction to diseasenowcasting: Real-Time Epidemic Nowcasting"
author: "Rodrigo Zepeda-Tello"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to diseasenowcasting: Real-Time Epidemic Nowcasting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 8,
  fig.height = 4,
  collapse = TRUE,
  comment  = "#>"
)
```

`diseasenowcasting` is an R package for nowcasting time series of epidemiological cases. Epidemiologic surveillance tools usually have an intrinsic delay between the **true date of an event** (`event_date`) and the **report date for that event** (`report_date`). Some examples include the true date being symptom onset or testing time and the report date corresponds to when the case was registered in the system. 
`diseasenowcasting` uses censored Bayesian structural time series models (via R's Template Model Builder [`RTMB`](https://cran.r-project.org/web/packages/RTMB/index.html)) to infer the cases that have not yet been reported thus providing a prediction of the final number of cases.


## Your data: the `tbl_now` format

`diseasenowcasting` works with data organised as a `tbl_now` object from the companion
[`tbl.now`](https://rodrigozepeda.github.io/tbl.now/) package.  A `tbl_now` is simply a data frame that has been annotated
with the roles of its columns: 
+ *event date* when the event happened (e.g. symptom onset)
+ *report date*  when the event was reported (e.g. date entered into the database)
+ *strata* (optional) columns that defined all the strata (e.g. sex and region)
+ *now*  (optional) the date until which to nowcast (assumes all events and reports before the now have been observed)


```{r load}
library(diseasenowcasting)
library(tbl.now)
library(ggplot2)
set.seed(27653)
```

As a quick example, here is how to build a `tbl_now` using surveillance data 
for dengue in Puerto Rico:

```{r tblnow-demo}
data(denguedat)
```

```{r, echo = FALSE}
head(denguedat, 6)
```


Which we can transform to a `tbl_now` by specifying the event and report dates as well as the `data_type`. A `tbl.now` allows the user to give both linelist and count-data.  

```{r}
dengue_tbl <- tbl_now(
  denguedat,
  event_date  = onset_week,    # symptom onset date
  report_date = report_week,   # when the record was reported
  strata      = gender,        # remove if no strata or add as vector all strata (e.g. strata = c(gender, age, region))
  data_type   = "linelist",    # another option is "count-incidence"  if data is aggregated
  now         =  as.Date("1991-01-01")
)
dengue_tbl
```

Once your data is a `tbl_now`, a single call to `nowcast()` does the rest.

## Example 1 -- Dengue fever (weekly, Puerto Rico)

We fit a stratified nowcast to illustrate the basic workflow. 

```{r dengue-fit}
nc_dengue <- nowcast(dengue_tbl, n_draws = 1000)
nc_dengue
```

The fitted model can be visualized with [autoplot()]:

```{r dengue-plot}
autoplot(nc_dengue) 
```

The shaded bars show the median while the errorbar has the 90 % credible intervals.  

Values can be obtained via `predict()` and `summary()`:

```{r dengue-results}
# Full posterior-predictive nowcast at every event-time
pred_dengue <- predict(nc_dengue)

#This creates a summary of mean and quantiles
summary(pred_dengue) |> tail(6)   # last 6 weeks
```


Additionally the `nowcast_diagnostic()` shows the fitted distribution for the delay, the smoothed epidemic process as well as the aggregated nowcast (for the sum of all strata):

```{r dengue-diagnostic}
nowcast_diagnostic(nc_dengue) 
```


## Example 2 -- Mpox (daily, USA, stratified by race)

The `mpoxdat` dataset (also in `tbl.now`) covers the 2022 mpox outbreak in New York City
with daily case counts stratified by race.  We use the
**two-stage** nowcast: Stage 1 estimates the delay distribution from a recent
window of data; Stage 2 uses multiple imputations of that delay to fit the
epidemic, correctly propagating delay uncertainty into the final estimates.

```{r mpox-prep}
data(mpoxdat)

mpox_tbl <- tbl_now(
  mpoxdat,
  event_date  = dx_date,
  report_date = dx_report_date,
  case_count  = n,
  data_type   = "count-incidence",
  now         =  as.Date("2022-08-15")
) 
```

A simple plot of the data shows that we should be taking into account daily effects. 

```{r mpoxplot}
ggplot(mpox_tbl) +
  geom_col(aes(x = dx_date, y = n), fill = "#5F7E62") +
  labs(x = "Event date", y = "Case count") +
  theme_diseasenowcasting() 
```


Although the model automatically tries to do it, we can also set it with `add_temporal_effects()`:

```{r}
mpox_tbl <- mpox_tbl |> 
  add_temporal_effects(temporal_effects(day_of_week = TRUE))
```

You can see that the `tbl_now` indicates its computation:

```{r, echo = FALSE}
mpox_tbl
```

One can also choose between several likelihoods, epidemic processess and delay distributions
and feed it into the `model()`. Here we use a Suceptible-Infected-Recovered model (SIR) with
a delay that follows a lognormal distribution:

```{r mpox-fit}
#Models can be modified via the model() 
mpox_model <- model(likelihood = nb_likelihood(), epidemic = sir_epidemic(), delay = lognormal_delay())

nc_mpox <- nowcast(mpox_tbl, model = mpox_model)

autoplot(nc_mpox) 
```


## Example 3 -- Comparing models with a backtest

A **backtest** reruns nowcasts at multiple historical dates and scores them
against the eventually-observed totals.  This lets you objectively compare
model families before committing to one for real-time monitoring.

The `backtest()` fits one nowcast per (`date` x `model`) cell. Those cells run 
in parallel through the [future](https://future.futureverse.org/) framework.  
By default they run sequentially; to use several CPU cores, 
set a parallel plan *before* calling `backtest()`:

```{r backtest-plan, eval=FALSE}
library(future)
plan(multisession, workers = 4)   # 4 parallel R sessions
# ... run backtest() ...
plan(sequential)                  # restore serial execution when done
```

In what follows we run a backtest in sequential mode however we
strongly recommend using as many workers in a multisession plan
as possible:

```{r backtest}
# Compare HSGP (flexible GP trend) vs AR1 (autoregressive trend) 
# and SIR (susceptible, infected, recovered) on mpox
models_to_compare <- list(
  model(nb_likelihood(), hsgp_epidemic(), lognormal_delay()),
  model(nb_likelihood(), ar1_epidemic(),  lognormal_delay()),
  model(nb_likelihood(), sir_epidemic(),  lognormal_delay())
)

backtest_mpox <- backtest(
  mpox_tbl,
  models  = models_to_compare,
  n_dates = 5, #Test 3 dates at random from the mpox data
  seed    = 42
)
```

```{r score}
# Rank by Weighted Interval Score (WIS) -- lower is better
score(backtest_mpox, metric = "wis", report = TRUE)
```

The `score()` output shows WIS, absolute percentage error (APE), and empirical
coverage at 50 % and 90 %.  A well-calibrated nowcast should have coverage close
to these levels.

## Next steps

This vignette covered the basics: building a `tbl_now`, fitting a nowcast with
`nowcast()`, inspecting results with `predict()` / `summary()` / `autoplot()`,
and comparing models with `backtest()` / `score()`.

A few directions to explore next:

<!--- TODO: FILL HERE REFERENCING THE OTHER VIGNETTES. HAVE A MODEL VIGNETTE FOR PUBLIC HEALTH PRACTICIONERS--->

See `?nowcast`, `?backtest`, `?score`, and `?model` for full documentation, and
the `SKILL.md` file in the package root for an AI-oriented API reference.
