---
title: "Introduction to diseasenowcasting: Real-Time Epidemic Nowcasting"
author: "diseasenowcasting team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to diseasenowcasting: Real-Time Epidemic Nowcasting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 7,
  fig.align='center',
  fig.height = 4,
  collapse = TRUE,
  comment  = "#>"
)
```

`diseasenowcasting` is an R package for nowcasting time series of epidemiological cases. Epidemiologic surveillance tools usually have an intrinsic delay between the **true date of an event** (`event_date`) and the **report date for that event** (`report_date`). Some examples include the true date being symptom onset or testing time and the report date corresponds to when the case was registered in the system.
`diseasenowcasting` uses censored Bayesian models (via R's Template Model Builder [`RTMB`](https://cran.r-project.org/web/packages/RTMB/index.html)) to infer the cases that have not yet been reported thus providing a prediction of the final number of casees.

## Your data: the `tbl_now` format

`diseasenowcasting` works with data organised as a `tbl_now` object from the companion
[`tbl.now`](https://rodrigozepeda.github.io/tbl.now/) package.  A `tbl_now` is simply a data frame that has been annotated
with the roles of its columns: 

+ *event date* when the event happened (e.g. symptom onset)

+ *report date*  when the event was reported (e.g. date entered into the database)

+ *strata* (optional) columns that defined all the strata (e.g. sex and region)

+ *now*  (optional) the date until which to nowcast (assumes all events and reports before 
the now have been observed and missing observations correspond to no observations - i.e. 
if one day there were not cases the missingness can be translated into zero cases)


```{r load}
library(diseasenowcasting)
library(tbl.now)
library(ggplot2)
set.seed(27653)
```

As a quick example, here is how to build a `tbl_now` using the following surveillance data 
for dengue in Puerto Rico:

```{r tblnow-demo}
data(denguedat)
```

```{r, echo = FALSE}
head(denguedat, 6)
```


We can transform the `data.frame` to a `tbl_now` by specifying the event and report dates (`onset` and `report` weeks respectively) as well as the `data_type` and the strata (in this case, `gender`). 

```{r}
dengue_tbl <- tbl_now(
  denguedat,
  event_date  = onset_week,    # symptom onset date
  report_date = report_week,   # when the record was reported
  data_type   = "linelist",    # another option is "count-incidence"  if data is aggregated
  now         =  as.Date("1991-01-01") #When is the now of the nowcast
)
dengue_tbl
```

Once your data is a `tbl_now`, a single call to `nowcast()` does the rest.

> For more information about `tbl_now` check [the package's website](https://rodrigozepeda.github.io/tbl.now/index.html). 

## Example 1 -- Dengue fever (setting up a stratified nowcast)

We fit a nowcast stratified by gender to illustrate the basic workflow. First we add
the column `gender` as strata to the `tbl_now`:

```{r denguestrata}
dengue_tbl <-  dengue_tbl |>  add_strata(gender)
```

Notice that the `tbl_now` automatically prints the `strata` specification below:

```{r}
dengue_tbl
```

We can also add temporal effects for example a weekly seasonality (52 seasons) as well
as a holiday effect using the `almanac` package:

```{r, eval=requireNamespace("almanac", quietly = TRUE)}
library(almanac)

#Specify 52 seasons (weekly) and holidays from the US
t_effects  <- temporal_effects(seasons = 52, holidays = cal_us_federal())

#Add the temporal effects
dengue_tbl <- dengue_tbl |> 
  add_temporal_effects(t_effects)

```


Finally we fit the model:

```{r dengue-fit}
nc_dengue <- nowcast(dengue_tbl)
```

The fitted model can be visualized with `autoplot()`. Note that the nowcast was already stratified by the strata specified in the `tbl_now`:

```{r dengue-plot, fig.cap="*Nowcast for dengue example. The shaded bars show the median while the errorbar has the 90% credible intervals.*"}
autoplot(nc_dengue) 
```

Values can be obtained via `predict()` and `summary()`:

```{r dengue-results, eval = FALSE}
# Full posterior-predictive nowcast at every event-time
pred_dengue <- predict(nc_dengue)

#This creates a summary of mean and quantiles
summary(pred_dengue) 
```



```{r dengue-results3, echo = FALSE}
# Full posterior-predictive nowcast at every event-time
pred_dengue <- predict(nc_dengue)

#This creates a summary of mean and quantiles
summary(pred_dengue) |> tail(6)
```


Additionally the `nowcast_diagnostic()` shows the fitted distribution for the delay, the smoothed epidemic process as well as the aggregated nowcast (for the sum of all strata):

```{r dengue-diagnostic, fig.height=10}
nowcast_diagnostic(nc_dengue) 
```


## Example 2 -- Mpox (modifying the nowcast model)

The `mpoxdat` dataset (also in `tbl.now`) covers the 2022 mpox outbreak in New York City
with daily case counts stratified by race.  

```{r mpox-prep}
data(mpoxdat)

mpox_tbl <- tbl_now(
  mpoxdat,
  event_date  = dx_date,
  report_date = dx_report_date,
  case_count  = n,
  data_type   = "count-incidence",
  now         =  as.Date("2022-08-15")
) 
```

A simple plot of the data shows that we should be taking into account day-of-the-week effects:

```{r mpoxplot, fig.height=8}
autoplot(mpox_tbl)
```


We can also set it again with `add_temporal_effects()`:

```{r}
mpox_tbl <- mpox_tbl |> 
  add_temporal_effects(temporal_effects(day_of_week = TRUE))
```

You can see that the `tbl_now` indicates its computation:

```{r, echo = FALSE}
mpox_tbl
```

One can also choose between several likelihoods, epidemic processess and delay distributions
and feed it into the `model()`. Here we use a Susceptible-Infected-Recovered model (SIR) with
a delay that follows a lognormal distribution:

```{r mpox-fit}
#Models can be modified via the model() 
mpox_model <- model(likelihood = nb_likelihood(),   #Negative binomial
                    epidemic   = sir_epidemic(),    #SIR model
                    delay      = lognormal_delay()) #Delay distribution

#We can then fit the  model
nc_mpox <- nowcast(mpox_tbl, model = mpox_model)

#And show the nowcast
autoplot(nc_mpox) 
```


## Example 3 -- Comparing models with a backtest

A **backtest** reruns nowcasts at multiple historical dates and scores them
against the eventually-observed totals.  This lets you compare between
models before committing to one for real-time monitoring. 

>The `backtest()`  function fits one nowcast per `date` and `model` cell. Those cells run 
in parallel through the [future](https://future.futureverse.org/) framework.  
By default they run sequentially; to use several CPU cores, 
set a [parallel plan](https://future.futureverse.org/reference/plan.html) *before* calling `backtest()`:

```{r backtest-plan, eval=FALSE}
library(future)
plan(multisession, workers = 4)   # 4 parallel R sessions
# ... run backtest() ...
plan(sequential)                  # restore serial execution when done
```

In what follows we run a backtest in sequential mode however we
strongly recommend using as many workers in a multisession plan
as possible:

```{r backtest}
# Compare HSGP (flexible GP trend) vs AR1 (autoregressive trend) 
# and SIR (susceptible, infected, recovered) on mpox
models_to_compare <- list(
  model(nb_likelihood(), hsgp_epidemic(), lognormal_delay()),
  model(nb_likelihood(), ar1_epidemic(),  lognormal_delay()),
  model(nb_likelihood(), sir_epidemic(),  lognormal_delay())
)

#Uncomment this line to use several of your cores
#plan(multisession, workers = 4) 

backtest_mpox <- backtest(
  mpox_tbl,
  models  = models_to_compare,
  n_dates = 5 #Test 3 dates at random from the mpox data
)

#This closes the plan multisession opened above
#plan(sequential)   
```

We can then plot the Weighted Interval Score (WIS) and interval coverage:

```{r scoreplot}
# Plot scoring metrics
autoplot(backtest_mpox)
```

Or obtain it via `score()`:

```{r score}
# Rank by Weighted Interval Score (WIS) -- lower is better
score(backtest_mpox)
```


The `score()` output shows WIS, absolute percentage error (APE), and empirical
coverage at 50 % and 90 %.  A well-calibrated nowcast should have the lowest
WIS and coverage close to these levels. 

> In this specific test we would choose the SIR for
having the lowest WIS at essentially the same coverage as HGSP. Note however that
for the tutorial  we only used 5 historical dates which is too low to reach
a definite conslusion. 

## Next steps

This vignette covered the basics: building a `tbl_now`, fitting a nowcast with
`nowcast()`, inspecting results with `predict()` / `summary()` / `autoplot()`,
and comparing models with `backtest()` / `score()`.

Depending on what you want to do next, check out the following
vignettes:

- **[Nowcasting at the Start of an Epidemic](Nowcasting_at_the_start_of_an_Epidemic.html)**
  —  A worked, end-to-end case study of monitoring an outbreak in real time: 
  choosing a model when data are scarce, reading the nowcast as the epidemic grows, 
  and experimenting with the prior to
  encode what you already believe about the epidemic before much data arrives.

- **[Understanding Priors in diseasenowcasting](Understanding_Priors.html)**
  — *Make the model say what you mean.* Shows the package's default priors, how
  to tighten or loosen them, and how the prior trades off against the data. Uses
  the prior-predictive tools (`nowcast(..., prior_only = TRUE)`) to *see* what a
  prior implies before fitting.
  
- **[Custom delays and epidemic processes](https://rodrigozepeda.github.io/diseasenowcasting/articles/Custom_delays_and_processes.html)**
  — Two examples on how to set **your own delays and epidemic processes**. Includes
  how to use ordinary differential equation models. 


- **[Handling Outlier Delays with Censoring](Handling_Outlier_Delays_with_Censoring.html)**
  — *Robustness to reporting glitches.* How the censored likelihood copes with
  unusually long reporting delays, and how to flag extreme 
  delays in your surveillance stream.
  
- **[Using alongside an LLM](LLM_Usage.html)**
  — *Use AI.* How to use the [`SKILL.md`](https://github.com/RodrigoZepeda/diseasenowcasting/blob/master/SKILL.md)
  to teach a Large Language Model how to you develop your nowcasts with `diseasenowcasting`. 

- **[Benchmark (diseasenowcasting vs NobBS and epinowcast)](Benchmark.html)**
  — *How does it compare?* A reproducible backtest comparing `diseasenowcasting`
  against the `NobBS` and `epinowcast` packages.

- **[Mathematical Foundations of diseasenowcasting](Mathematics.html)**
  — *Under the hood.* The censored likelihood, the epidemic processes
  (HSGP, AR(1), SIR), the delay families, and the Laplace-approximation inference
  that powers `RTMB`.

See `?nowcast`, `?backtest`, `?score`, and `?model` for full documentation.

