---
title: "Introduction to dcast3: Real-Time Epidemic Nowcasting"
author: "dcast3 package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dcast3: Real-Time Epidemic Nowcasting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 8,
  fig.height = 4,
  collapse = TRUE,
  comment  = "#>"
)
```

## What is nowcasting — and why does it matter?

When a new infectious disease case occurs, it almost never appears in official
surveillance data right away.  There is always a **reporting delay**: the time
between a person getting sick (or being tested positive) and that case being
entered into the national database.  This delay can range from a few days to
several weeks, depending on the disease and the health system.  The practical
consequence is that the most recent counts in any surveillance report are
*systematically too low* — not because cases didn't happen, but because they
haven't been reported yet.

**Nowcasting** is the statistical task of estimating the *true* number of cases
that have already occurred, correcting for this undercount.  A good nowcast
answers the question: "Given what we can see today, what is the most likely
epidemic size right now?"  This is essential for real-time public-health
decision-making — triggering interventions, allocating resources, and
communicating risk to the public — because acting on the raw, incomplete data
can lead to systematically delayed or wrong decisions.

The `dcast3` package implements a principled Bayesian nowcasting approach.  It
models the reporting delay as a stochastic process and estimates the epidemic
curve jointly with the delay distribution, so that uncertainty about both is
propagated into every estimate.  The result is a full probability distribution
over the current epidemic size, not just a point estimate.

---

## Your data: the `tbl_now` format

`dcast3` works with data organised as a `tbl_now` object from the companion
`tbl.now` package.  A `tbl_now` is simply a data frame that has been annotated
with the roles of its columns: which column is the *event date* (e.g. symptom
onset), which is the *report date* (e.g. date entered into the database), and
— optionally — which columns define *strata* (e.g. sex, region).

```{r load}
library(dcast3)
library(tbl.now)
library(ggplot2)
```

As a quick example, here is how to build a `tbl_now` from the Colombia COVID-19
data bundled with `dcast3`:

```{r tblnow-demo}
# covid_colombia ships with dcast3: 37,600 rows of daily counts by sex
head(covid_colombia, 4)

tn_covid <- tbl_now(
  covid_colombia,
  event_date  = notification_date,   # when the patient was notified / onset
  report_date = diagnosis_date,      # when the record entered the database
  strata      = sex,
  case_count  = n,
  data_type   = "count-incidence",
  verbose     = FALSE
)
tn_covid
```

Once your data is a `tbl_now`, a single call to `nowcast()` does the rest.

---

## Example 1 — Dengue fever (weekly, Puerto Rico)

`tbl.now` ships a weekly dengue linelist from Puerto Rico (`denguedat`).  We fit
an unstratified nowcast at a single point in time to illustrate the basic
workflow. 

```{r dengue-prep}
# denguedat: onset_week, report_week, gender
dn <- tbl_now(
  denguedat,
  event_date  = onset_week,
  report_date = report_week,
  data_type   = "linelist",
  verbose     = FALSE,
  now         =  as.Date("1991-01-01")
)
```

```{r dengue-fit}
mdl_dengue <- model(
  nb_likelihood(),      # negative-binomial counts (handles overdispersion)
  hsgp_epidemic(),      # Hilbert-space GP trend (flexible, smooth)
  lognormal_delay()     # log-normal reporting delay
)

nc_dengue <- nowcast(
  dn, mdl_dengue,
  type    = "two_stage",
  n_draws = 300,
  seed    = 42
)
nc_dengue
```

```{r dengue-results}
# Full posterior-predictive nowcast at every event-time
pred_dengue <- predict(nc_dengue, seed = 42)
summary(pred_dengue) |> tail(6)   # last 6 weeks
```

```{r dengue-plot}
autoplot(pred_dengue) +
  labs(title = "Dengue nowcast — Puerto Rico",
       subtitle = paste("as of", get_now(dn)),
       x = "Week index", y = "Cases") 
```

The shaded bands show the 50 % and 90 % credible intervals.  The most recent
weeks (rightmost points) carry the widest intervals — this is the nowcast
correcting for the fact that the latest data are the most incomplete.

---

## Example 2 — Mpox (daily, USA, stratified by race)

The `mpoxdat` dataset (also in `tbl.now`) covers the 2022 mpox outbreak in New York City
with daily case counts stratified by race.  We use the
**two-stage** nowcast: Stage 1 estimates the delay distribution from a recent
window of data; Stage 2 uses multiple imputations of that delay to fit the
epidemic, correctly propagating delay uncertainty into the final estimates.

```{r mpox-prep}
mp <- tbl_now(
  as.data.frame(mpoxdat),
  event_date  = dx_date,
  report_date = dx_report_date,
  strata      = race,
  case_count  = n,
  data_type   = "count-incidence",
  verbose     = FALSE
) |>
  add_temporal_effects(temporal_effects(day_of_week = TRUE)) |>
  compute_temporal_effects()

mp_now <- as.Date("2022-09-15")
```

```{r mpox-fit}
mdl_mpox <- model(
  nb_likelihood(),
  hsgp_epidemic(),
  lognormal_delay()
)

nc_mpox <- nowcast(
  mp, mdl_mpox,
  now     = mp_now,
  type    = "two_stage",   # delay uncertainty propagated via multiple imputation
  K       = 5,             # 5 delay imputations (increase to 25 in production)
  n_draws = 300,
  seed    = 42
)
nc_mpox
```

```{r mpox-results}
pred_mpox <- predict(nc_mpox, seed = 42)
autoplot(pred_mpox) +
  labs(title = "Mpox nowcast — United States (total across strata)",
       subtitle = paste("as of", format(mp_now), "| two-stage, stratified by race"),
       x = "Day index", y = "Cases")
```

Because the delay is shared across all race strata but each stratum has its own
epidemic trend, the model borrows information about the delay from the full
dataset while fitting independent curves per stratum.  The `predict()` output
reports the *total* (summed over strata), which is what matters for overall
epidemic monitoring.

---

## Example 3 — COVID-19 (daily, Colombia, stratified by sex)

Now we use the Colombia COVID-19 data bundled with `dcast3`.  We fit an
unstratified one-stage nowcast to a 90-day window in the early pandemic.

```{r covid-prep}
tn_covid_fit <- tbl_now(
  covid_colombia,
  event_date  = notification_date,
  report_date = diagnosis_date,
  strata      = sex,
  case_count  = n,
  data_type   = "count-incidence",
  verbose     = FALSE
) |>
  add_temporal_effects(temporal_effects(day_of_week = TRUE)) |>
  compute_temporal_effects()

cv_now <- as.Date("2020-06-01")
```

```{r covid-fit}
mdl_covid <- model(
  nb_likelihood(),
  hsgp_epidemic(),
  lognormal_delay()
)

nc_covid <- nowcast(
  tn_covid_fit, mdl_covid,
  now     = cv_now,
  type    = "one_stage",
  n_draws = 300,
  seed    = 42
)
nc_covid
```

```{r covid-results}
coef(nc_covid)   # delay mean/SD + epidemic parameters

pred_covid <- predict(nc_covid, seed = 42)
autoplot(pred_covid) +
  labs(title = "COVID-19 nowcast — Colombia (total, female + male)",
       subtitle = paste("as of", format(cv_now), "| stratified by sex"),
       x = "Day index", y = "Cases")
```

---

## Example 4 — Comparing models with a backtest

A **backtest** reruns nowcasts at multiple historical dates and scores them
against the eventually-observed totals.  This lets you objectively compare
model families before committing to one for real-time monitoring.

```{r backtest}
# Compare HSGP (flexible GP trend) vs AR1 (autoregressive trend) on dengue
models_to_compare <- list(
  model(nb_likelihood(), hsgp_epidemic(), lognormal_delay()),
  model(nb_likelihood(), ar1_epidemic(),  lognormal_delay())
)

bt <- backtest(
  dn,
  models  = models_to_compare,
  dates   = as.Date(c("2010-01-04", "2010-03-29", "2010-06-21")),
  type    = "one_stage",
  n_draws = 300,
  K       = 5,
  seed    = 42
)
```

```{r score}
# Rank by Weighted Interval Score (WIS) — lower is better
score(bt, metric = "wis", report = TRUE)
```

```{r backtest-plot}
autoplot(bt) +
  labs(title = "Dengue backtest — HSGP vs AR1",
       subtitle = "Ribbons = 50 % and 90 % credible intervals; points = final observed")
```

The `score()` output shows WIS, absolute percentage error (APE), and empirical
coverage at 50 % and 90 %.  A well-calibrated nowcast should have coverage close
to these nominal levels.

---

## Next steps

This vignette covered the basics: building a `tbl_now`, fitting a nowcast with
`nowcast()`, inspecting results with `predict()` / `summary()` / `autoplot()`,
and comparing models with `backtest()` / `score()`.

A few directions to explore next:

- **Delay families** — swap `lognormal_delay()` for `gamma_delay()`,
  `generalized_gamma_delay()`, or the non-parametric `dirichlet_delay()`.
- **Epidemic models** — try `sir_epidemic()` for a mechanistic compartmental
  model, or `ar1_epidemic()` when computational speed matters.
- **Two-stage cascade** — set `type = "two_stage"` with `K = 25` for the full
  multiple-imputation pipeline that propagates delay uncertainty into the final
  epidemic estimate (recommended for production use).
- **Custom priors** — pass a named list to the `phi` argument of `nowcast()` or
  supply a `priors` list to `fit()` directly.
- **Warm-start updates** — as new data arrive, `update(nc, new_rows)` refits
  from the previous estimates rather than cold, saving substantial computation.
- **Per-stratum results** — the fit object stores `[T × S]` lambda matrices
  accessible via `nc@fits[[1]]$lambda`; the `M_strata` array in the draw list
  gives per-stratum predictive counts.

See `?nowcast`, `?backtest`, `?score`, and `?model` for full documentation, and
the `SKILL.md` file in the package root for an AI-oriented API reference.
