---
title: "Nowcasting at the Start of an Epidemic"
author: "Rodrigo Zepeda-Tello"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Nowcasting at the Start of an Epidemic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 3.5,
  collapse = TRUE, comment = "#>"
)
set.seed(42)
```

```{r load, message=FALSE}
library(diseasenowcasting)
library(tbl.now)
library(ggplot2)
library(dplyr)
```

## TL;DR

1. Use SIR model with strong priors for 1-2 weeks. 
2. Use default AR model for 3-4 weeks.
3. Use HGSP for anything over 1 month. 

## The challenge: few data points

Nowcasting at the very start of an epidemic is different from
nowcasting during a well-established outbreak.  When only a handful of cases have
been reported, the data are almost completely uninformative about the future
trajectory and the model's **priors** dominate.  

This vignette shows you:

1. How to set each models priors (SIR, AR1, HSGP)  at the start of an epidemic.
2. How the three models perform on the **COVID-19 Colombia** data as the series
   grows from 1 day to 3 months.
3. A practical heuristic for deciding when to switch across the models.


## 1) Prior choices at the start of an epidemic

### 1.1 The SIR model: mechanistic priors

When data are scarce, the susceptible-infected-recovered (SIR) model imposes structural constraints:

* **R0 (basic reproductive number)**: The [basic reproductive number](https://en.wikipedia.org/wiki/Basic_reproduction_number), represents the
expected number of cases generated by one initial infectious case in a population. Alongside societal behaviour, it is partially related to the disease's biology and can be obtained from the literature, from epidemiological case studies or scientifically informed assumptions. 

* **gamma (recovery rate)**: the recovery rate is the inverse of the [infectious period](https://en.wikipedia.org/wiki/Infectious_period) ($\gamma = 1 / \text{infectious period}$). Where the infectious period can be thought of as how long a host maintains the capacity to transmit the disease to others. 

> **Note** For consistency, the units have to be identical to the units in the data. That is if data is weekly then the infectious period used to calculate the recovery rate should also be in weeks.

*  **N_eff**: The effective susceptible population.  That is, the proportion of individuals that can get infected. At the beginning of an outbreak in a completely susceptible population this value is close to 1. If part of the population is vaccinated this should be close to `1 - proportion vaccinated`. A closely related value that can be used initially is the expected [attack rate](https://en.wikipedia.org/wiki/Attack_rate). 

* **N_pop**: The total size of the population. That is, how many people are there (including the non-susceptible). 


```{r sir-priors}
# SIR priors for a new respiratory outbreak (unknown pathogen, cautious)
sir_model <- model(
  likelihood = nb_likelihood(phi = lognormal_prior(log(20), 0.5)),
  epidemic   = sir_epidemic(
    R0    = lognormal_prior(log(2.5),  0.1), # R0 ~ 2.5. Sd very tight. 
    gamma = lognormal_prior(log(1/14), 0.1), # infectious period ~ 14 days. Sd very tight. 
    N_eff = beta_prior(5, 2),                  # ~70% susceptible initially
    N_pop = 50000000                           # ~population of Colombia
  ),
  delay = lognormal_delay(
    mu    = normal_prior(log(5), 0.5),    # mean delay ~5 days
    sigma = gamma_prior(2, 0.5)           # delay SD
  )
)
```

Note that we are specifying the values of the sir epidemic process with a lot of certainty (`sdlog = 0.1`)
as strong priors are required to avoid broad prediction intervals.  Even with just 1-5 data points, 
the SIR model can produce a *plausible* forecast because the epidemic shape is constrained by the mechanistic structure.

### 1.2 The AR(1) model: trend priors

The AR(1) model places a prior on the *trend* rather than on epidemiological
parameters. 

* **ar_phi** reflects how much yesterday's values influence today's (initially set close to 1 to use yesterday as today's prediction).

* **ar_sigma** represents how fast you expect the trend to change.

```{r ar1-priors}
ar1_model <- model(
  nb_likelihood(phi = lognormal_prior(log(20), 0.5)),
  ar1_epidemic(
    phi   = normal_prior(0.9, 0.1),          # expect persistent growth
    sigma = half_normal_prior(0, 0.1)        # low innovation noise
  ),
  lognormal_delay(mu = normal_prior(log(5), 0.5))
)
```

> **Caveat:** AR(1) needs at least 10-15 data points to estimate the trend
reliably.  With fewer, it collapses toward the intercept prior.

### 1.3 The HSGP model: kernel priors

The HSGP is the most flexible model but the least informative _a priori_.  Its
priors control:

* **gp_alpha**: How large the trend can swing (amplitude).  Start narrow.

*  **gp_ell**: How fast the trend can change (length-scale).  Longer = smoother.

```{r hsgp-priors}
hsgp_model <- model(
  nb_likelihood(phi = lognormal_prior(log(20), 0.5)),
  hsgp_epidemic(
    alpha = lognormal_prior(log(1), 0.3),    # moderate amplitude, constrained
    ell   = lognormal_prior(log(1), 0.5)     # moderate length-scale
  ),
  lognormal_delay(mu = normal_prior(log(5), 0.5))
)
```

> **The HSGP requires the most data** to be reliable.  With < 20-30 observations,
it tends to produce very wide intervals that offer little guidance.

### 1.4 Experiment with the prior: prior-predictive simulation

How do you know whether a prior is sensible?  **Simulate from it before fitting.**
`nowcast(..., prior_only = TRUE)` ignores the data likelihood and draws the
epidemic parameters from their priors, returning the *prior-predictive* epidemic
curves.  This lets you literally see what a prior implies, and is the best way to
calibrate the strong priors you need early in an outbreak.

We only need a `tbl_now` to fix the time horizon (here 60 days):

```{r prior-grid}
grid_dates <- as.Date("2020-01-01") + 0:59
skeleton   <- tbl_now(
  data.frame(onset = grid_dates, reported = grid_dates),
  event_date = onset, report_date = reported,
  data_type = "linelist", verbose = FALSE
)

# Helper: prior-predictive median + 90% band of the latent incidence.
prior_band <- function(mod, label, n_draws = 300) {
  nc <- nowcast(skeleton, mod, prior_only = TRUE, n_draws = n_draws, seed = 1)
  q  <- quantile(nc, probs = c(0.05, 0.5, 0.95))
  data.frame(day = seq_len(nrow(q)) - 1L,
             lo = q[, 1], med = q[, 2], hi = q[, 3], scenario = label)
}
```

**SIR -- moving `R0`.**  The reproduction number controls how fast and how big
the epidemic is.  Compare three priors:

```{r prior-sir, fig.cap="Prior-predictive epidemic from the SIR model under three R0 priors (median + 90% band)."}
sir_for <- function(r0) model(
  nb_likelihood(),
  sir_epidemic(R0 = lognormal_prior(log(r0), 0.05), gamma = lognormal_prior(log(0.1), 0.1),
               N_pop = 5000),
  lognormal_delay(mu = normal_prior(log(5), 0.5))
)
sir_df <- rbind(prior_band(sir_for(1.5), "R0 = 1.5"),
                prior_band(sir_for(2.5), "R0 = 2.5"),
                prior_band(sir_for(4.0), "R0 = 4.0"))

ggplot(sir_df, aes(day, med, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  labs(x = "Day", y = "Prior-predictive incidence", colour = NULL, fill = NULL,
       title = "SIR: a higher R0 prior -> faster, larger epidemic") +
  theme_minimal(base_size = 11)
```

**AR(1) -- moving `phi`.**  The autocorrelation sets how *persistent* the trend
is.  A `phi` near 1 lets the trajectory wander far from the baseline; a `phi`
near 0 keeps it tightly mean-reverting:

```{r prior-ar1, fig.cap="Prior-predictive incidence from the AR(1) model under two phi priors."}
ar_for <- function(p) model(
  nb_likelihood(),
  ar1_epidemic(phi = normal_prior(p, 0.02), sigma = half_normal_prior(0, 0.5)),
  lognormal_delay(mu = normal_prior(log(5), 0.5))
)
ar_df <- rbind(prior_band(ar_for(0.10), "phi = 0.10 (mean-reverting)"),
               prior_band(ar_for(0.95), "phi = 0.95 (persistent)"))

ggplot(ar_df, aes(day, med, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  labs(x = "Day", y = "Prior-predictive incidence", colour = NULL, fill = NULL,
       title = "AR(1): a higher phi prior -> wider, more persistent excursions") +
  theme_minimal(base_size = 11)
```

**HSGP -- moving `alpha`.**  The GP amplitude controls how large the smooth
deviations from the baseline can be:

```{r prior-hsgp, fig.cap="Prior-predictive incidence from the HSGP model under two amplitude priors."}
hsgp_for <- function(a) model(
  nb_likelihood(),
  hsgp_epidemic(alpha = lognormal_prior(log(a), 0.1), ell = lognormal_prior(log(1), 0.3)),
  lognormal_delay(mu = normal_prior(log(5), 0.5))
)
hsgp_df <- rbind(prior_band(hsgp_for(0.2), "alpha = 0.2 (gentle)"),
                 prior_band(hsgp_for(1.5), "alpha = 1.5 (flexible)"))

ggplot(hsgp_df, aes(day, med, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  labs(x = "Day", y = "Prior-predictive incidence", colour = NULL, fill = NULL,
       title = "HSGP: a larger amplitude prior -> larger swings around the baseline") +
  theme_minimal(base_size = 11)
```

The take-away: **experiment with the prior** via `prior_only = TRUE` until the
simulated curves look like epidemics you consider plausible, *then* fit to data.

## 2) COVID-19 Colombia: example from 1 day to 3 months

### 2.1 Data preparation

For this example we use the `covid_colombia` dataset. This contains case counts for all COVID-19 cases registered in Colombia by notification and diagnosis dates. 

```{r covid-prep}
tn_covid <- tbl_now(
  covid_colombia,
  event_date  = notification_date,
  report_date = diagnosis_date,
  case_count  = n,
  data_type   = "count-incidence",
  t_effects   = temporal_effects(day_of_week = TRUE))
```

We will also get the final reported cases to compare against later:

```{r}
#Also get the final reports
final_report <- tn_covid |> 
  get_latest_reported_cases() 
```

### 2.2 Fitting each model across the five windows

We fit SIR, AR1, and HSGP at each window using tight, epidemiologically informed
priors for early windows and relaxing them as data accumulate.

```{r covid-window-fits, cache=FALSE}
# Starting date and progressive windows
t0 <- as.Date("2020-03-09")    # first day with >0 cases in all models
windows <- list(
  `1 week`   = t0,
  `2 weeks`  = t0 + 6,
  `3 weeks`  = t0 + 13,
  `4 weeks`  = t0 + 20,
  `1 month`  = t0 + 29,
  `2 months`  = t0 + 59,
  `3 months` = t0 + 89
)

results_list <- list()
for (win_name in names(windows)) {
  now <- windows[[win_name]]
  for (epi_name in c("SIR", "AR1", "HSGP")) {
    mdl <- switch(epi_name,
      SIR  = sir_model,
      AR1  = ar1_model,
      HSGP = hsgp_model
    )
    ncast          <- nowcast(tn_covid, mdl, now = now)
    df_predictions <- summary(predict(ncast))
    
    #Add some information
    df_predictions <- df_predictions |> 
      mutate(now = !!now) |> 
      mutate(window = !!win_name) |> 
      mutate(epidemic = !!epi_name) |> 
      left_join(final_report, by = join_by(.event_num)) |> 
      collect()
    
    #Return to list
    results_list[[length(results_list)+1]] <- df_predictions |> 
      filter(event_date >= now - 20)
  }
}

#Join all the models
all_results <- dplyr::bind_rows(results_list)

#Print the results
all_results <- all_results |> 
  mutate(window = factor(all_results$window, levels = names(windows)))
```

### 2.3 Comparing the three models across windows

Thw following shows all the nowcasted values alongside the final *truth*. Initially
the SIR outperforms AR(1) and HGSP. 

```{r covid-plots, fig.height=16, fig.cap="Nowcast (median + 90% CI) for each epidemic model at five progressive observation windows. Dashed line = nowcast as of `now`; shaded = future (the model's prediction of cases not yet reported)."}
# Plot only the last 20 event-times of each window (most relevant)
plot_data <- all_results |>
  group_by(window, epidemic) |>
  filter(!is.na(median)) |>
  arrange(.event_num) |>
  slice_tail(n = 20) |>
  ungroup()

ggplot(plot_data, aes(x = .event_num)) +
  geom_ribbon(aes(ymin = q2.5, ymax = q97.5, colour = epidemic, fill = epidemic), 
              alpha = 0.25, colour = NA) +
  geom_line(aes(y = median, colour = epidemic, fill = epidemic), linewidth = 0.9) +
  geom_point(aes(y = n, colour = "Truth"), linewidth = 0.9, shape = 17) +
  scale_colour_manual(
    values = c(SIR="#262626", AR1="#E8956A", HSGP="#A8BFA9", Truth = "#DA6529")
  ) +
  scale_fill_manual(
    values = c(SIR="#262626", AR1="#E8956A", HSGP="#A8BFA9", Truth = "#DA6529")
  ) +
  facet_wrap(window ~ epidemic, scales = "free", ncol = 3) +
  labs(x = "Event index (days from start)", y = "Daily cases",
       colour = "Model", fill = "Model",
       title = "COVID-19 Colombia: nowcast by model and observation window") +
  theme_diseasenowcasting()
```


### 2.4 Monitoring the fit with `ess()`

When you fit a flexible model (HSGP or AR1) with too few events, the Gaussian
(Laplace) approximation that `diseasenowcasting` uses becomes unreliable: the
posterior is skewed or barely moves from the priors.  The `ess()` function
quantifies this with an **importance-sampling effective sample size** over the
model's hyperparameters.  A high ESS (close to `n_draws`) means the
approximation is trustworthy; a low ESS warns you to interpret the intervals
with caution -- and, early in an epidemic, to fall back to the more constrained
SIR model.

```{r ess-demo}
# Fit the HSGP to an early window and check the approximation quality
early_tbl <- tn_covid |>
  dplyr::filter(notification_date <= as.Date("2020-03-20"))

nc_early <- nowcast(early_tbl, hsgp_model, now = as.Date("2020-03-20"),
                    type = "one_stage", n_draws = 300)

ess(nc_early, n_draws = 500)   # warns if the ESS is low
```

> Rule of thumb: if `ess()` warns (ESS / `n_draws` below ~10%), the data are not
> yet informative enough for HSGP or AR1 -- switch to SIR with tight priors until
> the ESS recovers as more data accumulate.

## 3) Conclusion: When to fit what?

| Observation period | Recommended model | Key priors to set |
|---|---|---|
| 1-2 weeks | SIR  | R0, gamma, N_eff, N_pop |
| 3-4 weeks | AR1  | ar_phi, ar_sigma |
| > 1 month | HSGP | default priors usually adequate |

The transition from mechanistic to data-driven models should mirror the evolution of
epidemiological knowledge during an outbreak: at the start, you rely on prior
knowledge; as data accumulate, you let the data speak.
