---
title: "Nowcasting at the Start of an Epidemic"
author: "diseasenowcasting team"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Nowcasting at the Start of an Epidemic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message  = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 3.5,
  fig.align='center',
  collapse = TRUE, comment = "#>"
)
set.seed(42)
```

```{r load, message=FALSE}
library(diseasenowcasting)
library(tbl.now)
library(ggplot2)
library(dplyr)
```

## TL;DR

1. Use SIR model with strong priors for 1-2 weeks (if data is daily) or 1-3 months (if data is weekly). 
Use `prior_only` to experiment with the priors for plausible curves. 
  
2. Use default AR model for 3-4 weeks (if data is daily) or  4-7 months (if data is weekly).

3. Use AR/HGSP for anything between 1-3 months (if data is daily) or 7 months to 1 year (if data is weekly). 

4. Use HGSP for long epidemic measurements 3+ months (if data is daily) or 1 year (if data is weekly).

## The challenge: few data points

Nowcasting at the very start of an epidemic is different from
nowcasting during a well-established outbreak.  When only a handful of cases have
been reported, the data are almost completely uninformative about the future
trajectory and the model's **priors** dominate.  

This vignette shows you:

1. How to set each models priors (SIR, AR1, HSGP)  at the start of an epidemic.
2. How the three models perform on the **COVID-19 Colombia** data as the series
   grows from 1 day to 3 months.
3. A practical heuristic for deciding when to switch across the models.

### 0) Prelude: Data preparation

For this example we use the `covid_colombia` dataset. This contains case counts for all COVID-19 cases registered in Colombia by notification and diagnosis dates. 

```{r covid-prep}
tn_covid <- tbl_now(
  covid_colombia,
  event_date  = notification_date,
  report_date = diagnosis_date,
  case_count  = n,
  data_type   = "count-incidence",
  t_effects   = temporal_effects(day_of_week = TRUE))
```

We will also get the final reported cases to compare against later:

```{r}
#Also get the final reports
final_report <- tn_covid |> 
  get_latest_reported_cases() 
```

## 1) Prior choices at the start of an epidemic

### 1.1 The SIR model: mechanistic priors

When data are scarce, the mechanistic susceptible-infected-recovered (SIR) model is best as it imposes structural constraints:

* **R0 (basic reproductive number)**: The [basic reproductive number](https://en.wikipedia.org/wiki/Basic_reproduction_number), represents the
expected number of cases generated by one initial infectious case in a population. Alongside societal behaviour, it is related to the disease's biology and can be approximated from the literature, from epidemiological case studies or scientifically informed assumptions. 

* **gamma (recovery rate)**: the recovery rate is the inverse of the [infectious period](https://en.wikipedia.org/wiki/Infectious_period) ($\gamma = 1 / \text{infectious period}$). The infectious period can be thought of as how long a host maintains the capacity to transmit the disease to others. This can also be approximated from the literature or scientific assumptions. 

> **Note** For consistency, the units have to be identical to the units in the data. That is if data is weekly then the infectious period used to calculate the recovery rate should also be in weeks.

*  **N_eff**: The effective susceptible population.  That is, the proportion of individuals that can get infected. At the beginning of an outbreak in a completely susceptible population this value is close to 1. If part of the population is vaccinated this should be close to `1 - proportion vaccinated`. A closely related value that can also be used initially is the expected [attack rate](https://en.wikipedia.org/wiki/Attack_rate). 

* **N_pop**: The total size of the population. That is, how many people are there (including the non-susceptible). 


```{r sir-priors}
# SIR priors for a COVID-19 outbreak 
# Assumptions reflect data known in 2020 
sir_model <- model(
  likelihood = nb_likelihood(),
  epidemic   = sir_epidemic(
    R0    = lognormal_prior(log(2.5),  0.1), # R0 ~ 2.5. Sd very tight. 
    gamma = lognormal_prior(log(1/14), 0.1), # infectious ~ 14 days / tight sd
    N_eff = beta_prior(5, 2),                # ~70% susceptible initially
    N_pop = 50000000                         # ~population of Colombia
  ),
  delay = lognormal_delay(
    mu    = normal_prior(log(5), 0.5),    # mean report delay ~5 days
    sigma = gamma_prior(2, 0.5)           # delay SD
  )
)
```

Note that we are specifying the values of the SIR epidemic process with a lot of certainty (`sdlog = 0.1`)
as strong priors are required to avoid broad prediction intervals.  Even with just 1-5 data points, 
the SIR model can produce a *plausible* forecast because the epidemic shape is constrained by the mechanistic structure.

### 1.2 The AR(1) model: trend priors

The AR(1) model places a prior on the *trend* rather than on epidemiological
parameters. 

* **ar_phi** reflects how much yesterday's values influence today's (initially set close to 1 to use yesterday as today's prediction).

* **ar_sigma** represents how fast you expect the trend to change.

```{r ar1-priors}
ar1_model <- model(
  nb_likelihood(),
  ar1_epidemic(
    phi   = normal_prior(0.9, 0.1),       # expect persistent growth
    sigma = exponential_prior(100)        # low innovation noise
  ),
  delay = lognormal_delay(
    mu    = normal_prior(log(5), 0.5),    # mean delay ~5 days
    sigma = gamma_prior(2, 0.5)           # delay SD
  )
)
```

> **Caveat:** AR(1) needs at least 10-15 data points to estimate the trend
reliably.  With fewer, it collapses toward the intercept prior.

### 1.3 The HSGP model: kernel priors

The HSGP is the most flexible model but the least informative _a priori_.  Its
priors control:

* **gp_alpha**: How large the trend can swing (amplitude).  Start narrow.

*  **gp_ell**: How fast the trend can change (length-scale).  Longer = smoother.

```{r hsgp-priors}
hsgp_model <- model(
  nb_likelihood(),
  hsgp_epidemic(
    alpha = half_normal_prior(0, 1),
    ell   = inv_gamma_prior(3, 1) 
  ),
  delay = lognormal_delay(
    mu    = normal_prior(log(5), 0.5),    # mean delay ~5 days
    sigma = gamma_prior(2, 0.5)           # delay SD
  )
)
```

> **The HSGP requires the most data** to be reliable.  With < 20-30 observations,
it tends to produce very wide intervals that offer little guidance.

## 2) Experiments with the prior: prior-predictive simulation

To know that a prior is sensible you can  **simulate from it before fitting via the
`prior_only` option. Setting `nowcast(..., prior_only = TRUE)` ignores the data likelihood and draws the
epidemic parameters solely from the priors, returning the *prior-predictive* epidemic
curves.  This lets you see what a prior implies, and is the best way to
calibrate the strong priors you need early in an outbreak.

The function call is very simple you just set up a nowcast with the option
`prior_only = TRUE`:

```{r, eval = FALSE}
nowcast(data = tn_covid, prior_only = TRUE) 
```

The following example show how slightly moving different parameters
generates different prior processess:

### 2.1) SIR -- moving `R0`.  

The reproduction numbe `R0` controls how fast and how big the epidemic is. 
Here we loop through three different `R0` values showing how a higher `R0` 
translates into faster growth. 

```{r prior-sir, fig.cap="*Prior-predictive epidemic from the SIR model under three R0 priors (median + 90% band).*"}
#Loop through different values for R0
model_list <- list(); k <- 0
for (R0val in c(1.5, 2.5, 4.0)){
  k  <- k + 1
  
  #Create a nowcast
  ncast <- nowcast(data  = tn_covid, 
                   model = model(
                     epidemic = sir_epidemic(
                       R0    = lognormal_prior(log(R0val), 0.05), #<- Change
                       gamma = lognormal_prior(log(1/14), 0.1), 
                       N_eff = beta_prior(5, 2),                
                       N_pop = 50000000 
                     )
                   ), 
                   now     = as.Date("2020/05/01"),
                   n_draws = 300,
                   prior_only = TRUE) 
  
  #Get the quantiles
  q  <- quantile(ncast, probs = c(0.05, 0.5, 0.95))
  model_list[[k]] <- data.frame(day = seq_len(nrow(q)) - 1L, 
                                lo = q[, 1], med = q[, 2], hi = q[, 3],
                                scenario = paste0("R0 = ", R0val))
  
}

sir_df <- dplyr::bind_rows(model_list)

ggplot(sir_df, aes(day, med, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  labs(x = "Day", y = "Prior-predictive incidence", colour = NULL, fill = NULL,
       title = "SIR: a higher R0 prior -> faster growth") +
  scale_color_manual(values = unname(dn_palette(3))) +
  scale_fill_manual(values = unname(dn_palette(3))) +
  theme_diseasenowcasting()
```

> The take-away: **experiment with the prior** via `prior_only = TRUE` until the
simulated curves look like epidemics you consider plausible, *then* fit to data.

## 2) COVID-19 Colombia: example from 1 day to 3 months

### 2.2 Fitting each model across the five windows

We fit SIR, AR1, and HSGP at each window using tight, epidemiologically informed
priors for early windows and relaxing them as data accumulate.

```{r covid-window-fits, cache=FALSE}
# Starting date and progressive windows
t0 <- as.Date("2020-03-09")    # first day with >0 cases in all models
windows <- list(
  `1 week`   = t0,
  `2 weeks`  = t0 + 6,
  `3 weeks`  = t0 + 13,
  `4 weeks`  = t0 + 20,
  `1 month`  = t0 + 29,
  `2 months`  = t0 + 59,
  `3 months` = t0 + 89
)

results_list <- list()
for (win_name in names(windows)) {
  now <- windows[[win_name]]
  for (epi_name in c("SIR", "AR1", "HSGP")) {
    mdl <- switch(epi_name,
      SIR  = sir_model,
      AR1  = ar1_model,
      HSGP = hsgp_model
    )
    ncast          <- nowcast(tn_covid, mdl, now = now)
    df_predictions <- summary(predict(ncast))
    
    #Add some information
    df_predictions <- df_predictions |> 
      mutate(now = !!now) |> 
      mutate(window = !!win_name) |> 
      mutate(epidemic = !!epi_name) |> 
      left_join(final_report, by = join_by(.event_num)) |> 
      collect()
    
    #Return to list
    results_list[[length(results_list)+1]] <- df_predictions |> 
      filter(event_date >= now - 20)
  }
}

#Join all the models
all_results <- dplyr::bind_rows(results_list)

#Print the results
all_results <- all_results |> 
  mutate(window = factor(all_results$window, levels = names(windows)))
```

### 2.3 Comparing the three models across windows

Thw following shows all the nowcasted values alongside the final *truth*. Initially
the SIR outperforms AR(1) and HGSP. As more data arrives HGSP takes over while
the SIR model eventually blows up. 

```{r covid-plots, fig.height=16, fig.cap="*Nowcast (median + 90% CI) for each epidemic model at five progressive observation windows.*"}
# Plot only the last 20 event-times of each window (most relevant)
plot_data <- all_results |>
  group_by(window, epidemic) |>
  filter(!is.na(median)) |>
  arrange(.event_num) |>
  slice_tail(n = 20) |>
  ungroup() |> 
  mutate(days_to_nowcast = event_date - now)

ggplot(plot_data, aes(x = days_to_nowcast)) +
  geom_ribbon(aes(ymin = q2.5, ymax = q97.5, colour = epidemic, 
                  fill = epidemic),  alpha = 0.25, colour = NA) +
  geom_line(aes(y = median, colour = epidemic, fill = epidemic), 
            linewidth = 0.9) +
  geom_point(aes(y = n, colour = "Truth"), linewidth = 0.9, shape = 17) +
  scale_colour_manual(
    values = c(SIR="#262626", AR1="#E8956A", HSGP="#A8BFA9", Truth = "#DA6529")
  ) +
  scale_fill_manual(
    values = c(SIR="#262626", AR1="#E8956A", HSGP="#A8BFA9", Truth = "#DA6529")
  ) +
  facet_wrap(window ~ epidemic, ncol = 3, scales = "free") +
  labs(x = "Days before the nowcast", y = "Daily cases",
       colour = "Model", fill = "Model",
       title = "COVID-19 Colombia: nowcast by model and observation window") +
  theme_diseasenowcasting()
```


### 2.4 A rule of thumb

If the interval is extremely wide (say interval width > 5 times the usual case-count)
then data is insufficient for HGSP or AR and you should switch to SIR. If
this happens in SIR the priors are probably too wide and need
to be tightened. 


## 3) Conclusion: When to fit what?

| Daily data observation period | Weekly data observation period | Recommended model | Key priors to set |
|---|---|---|---|
| 1-2 weeks | 1-3 months | SIR  | R0, gamma, N_eff, N_pop |
| 3-4 weeks | 4-7 months | SIR / AR1 / HGSP  | ar_phi, ar_sigma |
| 1-3 months | 7-12 months | AR1 / HSGP | default priors usually adequate |
| 3 months or more | 1 year or more | HSGP | default priors usually adequate |

The transition from mechanistic to data-driven models should mirror the evolution of
epidemiological knowledge during an outbreak: at the start, you rely on prior
knowledge; as data accumulate, you let the data speak.
