---
title: "Understanding Priors in diseasenowcasting"
author: "diseasenowcasting package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding Priors in diseasenowcasting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message   = FALSE,
  warning   = FALSE,
  fig.width = 9,
  fig.height = 4,
  collapse  = TRUE,
  comment   = "#>"
)
```

```{r load, include=FALSE}
library(diseasenowcasting)
library(ggplot2)

# Load pre-computed sensitivity data (avoids re-running fits at install time)
ps_file <- system.file("extdata", "prior_sensitivity.rds", package = "diseasenowcasting")
ps <- if (file.exists(ps_file)) readRDS(ps_file) else NULL

# Helper: filter and plot a prior comparison
plot_prior_effect <- function(prior_name, ylab = "Nowcast (median, 50/90% CI)",
                              title = NULL, n_recent = 30) {
  if (is.null(ps)) {
    return(ggplot() + annotate("text", x=1, y=1, label="Pre-computed data not found.\nRun devel/precompute_prior_sensitivity.R first.") + theme_void())
  }
  dat <- ps[ps$prior_name == prior_name, ]
  if (nrow(dat) == 0) return(ggplot() + theme_void() + labs(title = paste("No data for", prior_name)))

  # Keep only the most recent events per fit
  dat <- do.call(rbind, lapply(split(dat, list(dat$disease, dat$scenario)), function(block) {
    block <- block[order(block$.event_num, decreasing = TRUE), ]
    head(block, n_recent)
  }))

  dat$scenario <- factor(dat$scenario, levels = c("tight", "default", "loose"))
  dat$disease  <- factor(dat$disease,  levels = c("dengue", "mpox", "covid"))

  ggplot(dat, ggplot2::aes(x = .event_num, colour = scenario, fill = scenario)) +
    ggplot2::geom_ribbon(ggplot2::aes(ymin = q2.5, ymax = q97.5), alpha = 0.12, colour = NA) +
    ggplot2::geom_ribbon(ggplot2::aes(ymin = q25,  ymax = q75),  alpha = 0.20, colour = NA) +
    ggplot2::geom_line(ggplot2::aes(y = median), linewidth = 0.8) +
    ggplot2::scale_colour_manual(values = c(tight="#e41a1c", default="#377eb8", loose="#4daf4a")) +
    ggplot2::scale_fill_manual(  values = c(tight="#e41a1c", default="#377eb8", loose="#4daf4a")) +
    ggplot2::facet_wrap(~disease, scales = "free_y", nrow = 1) +
    ggplot2::labs(x = "Event index (most recent on the right)",
                  y = ylab, colour = "Prior scenario", fill = "Prior scenario",
                  title = title %||% paste("Effect of", prior_name, "prior")) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(legend.position = "bottom")
}
```

## What are priors, and why do they matter?

In Bayesian statistics, a **prior** is a probability distribution that encodes
your beliefs about a parameter *before* seeing the data.  When the data are
informative (many cases, long reporting history), the prior has little impact —
the data overwhelm it.  When the data are sparse or noisy (early epidemic,
small populations), the prior can substantially shape the estimates.

In `diseasenowcasting` every estimated parameter has a prior: the delay distribution, the
epidemic trend, the overdispersion, and the GP kernel hyperparameters.  This
vignette shows what happens to the nowcast when you move each prior, using
pre-computed results on three real diseases.

**Tip:** the "default" scenario in every plot uses the package defaults.
The "tight" scenario uses a very concentrated prior (strong belief in a specific
value); the "loose" scenario uses a diffuse prior (little prior information).

All plots below were generated with 300 posterior draws at a single as-of date
and are stored in `inst/extdata/prior_sensitivity.rds`.

---

## 1. HSGP epidemic priors

The **Hilbert-space Gaussian process** (`hsgp_epidemic()`) places a smooth
prior on the epidemic trend $f(t)$ via two hyperparameters:

- **`gp_alpha`** — the *amplitude* (marginal standard deviation) of the GP.  A
  larger amplitude allows bigger swings in the trend.
- **`gp_ell`** — the *length-scale*.  A larger length-scale forces the trend to
  vary more slowly (smoother); a smaller one allows rapid changes.

Both are given log-Normal priors by default:

```r
gp_alpha ~ LogNormal(log(1), 0.5)   # amplitude
gp_ell   ~ LogNormal(log(1), 0.5)   # length-scale
```

### 1.1 GP amplitude (`gp_alpha`)

```{r hsgp-alpha, fig.cap="Effect of the GP amplitude prior. Tight prior constrains the curve's overall height; loose prior allows more extreme epidemic sizes."}
plot_prior_effect("gp_alpha", title = "GP amplitude prior (gp_alpha)")
```

**What to watch for:** A tight amplitude prior (red) shrinks the curve toward
the prior mean and produces narrower credible intervals — useful when you are
confident the epidemic is not unusually large.  A loose prior (green) allows the
model to fit very high or very low incidence, producing wider intervals at the
most recent event times.

### 1.2 GP length-scale (`gp_ell`)

```{r hsgp-ell, fig.cap="Effect of the GP length-scale prior. A small length-scale (tight prior concentrated near small values) allows rapid changes; a large length-scale forces a smoother curve."}
plot_prior_effect("gp_ell", title = "GP length-scale prior (gp_ell)")
```

**What to watch for:** With a tight prior on a *small* length-scale the curve
can turn sharply, capturing rapid epidemic accelerations.  With a loose prior
(or concentrated on a large value) the curve is forced to be smooth, which may
miss short-term peaks but reduces noise.

---

## 2. AR(1) epidemic priors

The autoregressive epidemic (`ar1_epidemic()`) models the log-incidence as an
AR(1) process.  The two priors are:

- **`ar_phi`** — autocorrelation coefficient (how strongly the trend
  at time $t$ depends on time $t-1$).
- **`ar_sigma`** — standard deviation of the AR innovations (how much the trend
  can jump each step).

```{r ar1-priors, fig.cap="Effect of the AR1 prior settings. Tight ar_phi/sigma priors make the trend smooth and mean-reverting; loose priors allow larger, more persistent swings."}
plot_prior_effect("ar_phi/ar_sigma", title = "AR(1) autocorrelation and innovation SD priors")
```

**What to watch for:** A tight prior (concentrated near $\phi = 0$, small
$\sigma$) forces the trend toward a smooth, quickly mean-reverting process
with narrow intervals.  A loose prior (large $\sigma$) allows large jumps and
persistent trends, producing wider intervals.

---

## 3. SIR epidemic priors

The mechanistic SIR model (`sir_epidemic()`) is parameterised by:

- **`R0`** — the basic reproduction number.
- **`gamma`** (recovery rate) — implicitly sets the infectious period.
- **`N_eff`** — the effective susceptible fraction of the population.

### 3.1 Basic reproduction number (`R0`)

```{r sir-r0, fig.cap="Effect of the R0 prior. A tight prior near R0=2 confines the epidemic growth rate; a loose prior allows both sub-critical and super-critical dynamics."}
plot_prior_effect("R0", title = "SIR: basic reproduction number prior (R0)")
```

**What to watch for:** If the tight prior is mis-specified (e.g. you assume
$R_0 \approx 2$ but the true $R_0$ is 4), the model under-predicts at growth
phases.  Use a loose prior unless you have strong epidemiological knowledge of
the pathogen.

---

## 4. Delay distribution priors

The reporting-delay prior controls how the model thinks delays are distributed
*before* seeing any data.  `diseasenowcasting` supports four delay families.

### 4.1 Log-Normal delay (`lognormal_delay`)

The most widely used family.  Priors on the log-mean (`delay_mu`) and the
natural-scale SD (`delay_sigma`):

```{r delay-lognormal, fig.cap="Effect of the log-normal delay prior. A tight mean prior pins the delay to a specific value; a loose prior lets the data determine it freely."}
plot_prior_effect("delay_mu/sigma (LogNormal)", title = "Log-Normal delay prior")
```

**What to watch for:** The delay prior primarily affects the *level* of the
nowcast at the most recent event times (the rightmost points).  A tight prior
that mismatches the true delay leads to systematic bias: if the prior delay is
too short, the model under-inflates the most recent counts; if too long, it
over-inflates them.

### 4.2 Gamma delay (`gamma_delay`)

```{r delay-gamma, fig.cap="Effect of the Gamma delay prior. Similar structure to log-normal but with lighter tails."}
plot_prior_effect("delay_mu/sigma (Gamma)", title = "Gamma delay prior")
```

### 4.3 Generalised Gamma shape (`generalized_gamma_delay`)

The shape parameter `Q` interpolates between a Weibull ($Q = 1$) and a
log-normal ($Q \to 0$).  The prior on `Q` determines whether the model
prefers heavier or lighter tails:

```{r delay-gengamma, fig.cap="Effect of the GenGamma Q prior. Tight prior near Q=1 (Weibull-like) vs. loose prior that lets Q vary freely."}
plot_prior_effect("delay_Q", title = "GenGamma shape prior (delay_Q)")
```

### 4.4 Dirichlet concentration (`dirichlet_delay`)

The non-parametric Dirichlet delay uses a concentration parameter $\alpha$ that
controls how spread the prior mass is across delay bins.  A high $\alpha$ means
the prior expects roughly equal probability for every delay; a low $\alpha$
(< 1) allows the mass to concentrate on a few bins (sparse delay):

```{r delay-dirichlet, fig.cap="Effect of the Dirichlet concentration prior. High concentration = flat delay distribution; low concentration = concentrated on a few delays."}
plot_prior_effect("dirichlet_alpha", title = "Dirichlet delay concentration prior")
```

---

## 5. Likelihood prior: NB overdispersion (`phi`)

The negative-binomial dispersion parameter $\phi$ controls how much variability
in case counts is *not* explained by the epidemic trend.  Smaller $\phi$ means
more overdispersion (wider count intervals for the same mean); larger $\phi$
means less overdispersion (closer to Poisson).  The default prior is:

```r
phi_nb ~ LogNormal(log(20), 0.5)   # mode ~15, 90% CI roughly [6, 55]
```

```{r phi-nb, fig.cap="Effect of the NB phi prior. A tight prior near phi=50 (low overdispersion, almost Poisson) produces narrow intervals; a loose prior near phi=5 (high overdispersion) produces very wide intervals."}
plot_prior_effect("phi_nb", title = "NB overdispersion prior (phi)")
```

**What to watch for:** This is the single most impactful prior for the *width*
of the prediction intervals.  If the 90% intervals consistently miss the truth
(intervals too narrow), try a prior with a lower mode — e.g.
`lognormal_prior(log(5), 0.5)` — to allow more overdispersion.

---

## 6. Recommendations

The table below summarises what to change first depending on the symptom you
observe in your nowcast.

```{r recommendations, echo=FALSE}
recs <- data.frame(
  Symptom = c(
    "Credible intervals too wide",
    "Credible intervals too narrow (too many truths outside)",
    "Nowcast systematically too high at the newest event",
    "Nowcast systematically too low at the newest event",
    "Epidemic trend too wiggly / noisy",
    "Epidemic trend too smooth / misses peaks",
    "Rapid epidemic growth is under-predicted",
    "Model converges slowly or not at all",
    "Two-stage pipeline produces different results each run"
  ),
  `Start here` = c(
    "Tighten `phi_nb` prior (increase mode, e.g. `lognormal_prior(log(50), 0.3)`)",
    "Loosen `phi_nb` prior (decrease mode, e.g. `lognormal_prior(log(5), 0.5)`)",
    "Delay prior `delay_mu` is too large — move it toward the empirical mean delay",
    "Delay prior `delay_mu` is too small — or `delay_sigma` is too tight, capping the tail",
    "Tighten `gp_alpha` or increase `gp_ell` (longer length-scale → smoother)",
    "Loosen `gp_alpha` or decrease `gp_ell` (shorter length-scale → more flexible)",
    "For SIR: loosen `R0` prior (allow higher values); for HSGP: loosen `gp_alpha`",
    "Tighten all priors slightly; try a different `delay_mu` starting point",
    "Increase `K` (delay imputations) or set `floor_mu` / `floor_sig_frac` higher"
  ),
  `Also consider` = c(
    "Check `gp_alpha` prior — very loose amplitude can inflate intervals independently of phi",
    "Check whether the true epidemic is overdispersed (e.g. clustered outbreaks need low phi)",
    "Verify delay family: LogNormal is most robust; try Gamma or GenGamma if LN over-inflates",
    "Use GenGamma or Dirichlet to capture heavy-tailed delays",
    "Switch from HSGP to AR1 (AR1 can be smoother for short series)",
    "Increase `num_basis` in `hsgp_epidemic()` for long daily series",
    "Consider SIR model — mechanistic growth is built in",
    "Run `backtest()` across multiple dates to identify date-specific failures",
    "`np_spread` controls Dirichlet imputation width; increase for more variability"
  ),
  check.names = FALSE
)
knitr::kable(recs, caption = "Prior tuning recommendations by symptom.")
```

### Quick-reference prior changes

```r
# Wider intervals (more overdispersion) -- set phi on the likelihood:
mdl <- model(nb_likelihood(phi = lognormal_prior(log(5), 0.5)),
             hsgp_epidemic(), lognormal_delay())
nc  <- nowcast(tn, mdl, ...)

# Narrower intervals (less overdispersion):
mdl <- model(nb_likelihood(phi = lognormal_prior(log(50), 0.3)),
             hsgp_epidemic(), lognormal_delay())
nc  <- nowcast(tn, mdl, ...)

# Override any prior via default_priors() + fix_param()
priors <- default_priors(mdl, engine)
priors <- fix_param(priors, "delay_mu",    log(7))   # fix delay mean to log(7)
priors <- fix_param(priors, "delay_sigma", 5)         # fix delay SD to 5
fit_result <- fit(mdl, engine, priors = priors)

# Widen the GP amplitude prior
priors$gp_alpha <- list(dist = lognormal_prior(0,1)@num_id,
                        params = c(log(1), 1.5, 0),
                        is_constant = 0L, fixed = numeric(0))
```

### The most important priors to get right

1. **`phi_nb`** — the single biggest driver of interval width.  Start here.
2. **`delay_mu`** — determines how much the model inflates the most recent
   counts.  If you know the typical reporting delay from domain knowledge,
   anchor the prior around that value.
3. **`gp_ell`** (HSGP) / **`ar_sigma`** (AR1) — control how flexible the
   epidemic trend is.  Tune these if the model tracks noise or misses real peaks.
4. **`R0`** (SIR only) — if using the mechanistic model, a plausible prior on
   $R_0$ is essential for stability, especially at epidemic turning points.
